---
title: Synchronous effects produce major cycles in deer populations
  and deer-vehicle collisions
author:
 - Thomas L. Anderson\textsuperscript{1,2}, Lawrence W. Sheppard\textsuperscript{2}, Jonathan A. Walter\textsuperscript{2,3}, Robert E. Rolley\textsuperscript{4},  Daniel C. Reuman\textsuperscript{2,5}  
 - \textsuperscript{1}Department of Biology, Appalachian State University, 572 Rivers St., Boone, NC 28608, USA; \textsuperscript{2}Deparment of Ecology and Evolutionary Biology and Kansas Biological Survey, University of Kansas, 2101 Constant Ave, Lawrence, KS 66049, USA;  \textsuperscript{3}Department of Environmental Sciences, University of Virginia 291 McCormick Rd, Charlottesville, VA 22904, USA; \textsuperscript{4}Wisconsin Department of Natural Resources, 101 S. Webster St., Madison, WI 53707, USA; \textsuperscript{5}Laboratory of Populations, Rockefeller University, 1230 York Ave., New York, NY 10065, USA
date: 'correspondence to: anderstl@gmail.com, reuman@ku.edu'
output:
  pdf_document:
    fig_caption: yes
    keep_tex: yes
    includes:
      in_header: head_maintext.sty
  html_document:
    df_print: paged
  word_document: default
geometry: left=1cm,right=1cm,top=2.5cm,bottom=2.8cm
link-citations: yes
csl: ecologyletters.csl
tables: yes
bibliography: AndersDeerShortRefs.bib
urlcolor: blue
---

<!--Notes from convos with LWS and TLA, i.e., things we need to do:
1) We need to check that the model with MEI and snow depth is statistically better than the model with just snow depth,
and also statistically better than the model with just MEI, in the 3-7yr timescale band. Tom will do.
2) Lawrence then wants us to check that the phases of the coefficients are consistent between the models with one
predictor and the model with two. This will give us evidence related to whether the two-variable model is overfitted.
3) Dan needs to insert Fourier results. 
4) Lawrence is working on analyses which may provide an estimate of how inflated our fraction of synchrony explained
estimate is.
-->

\noindent Short title: Spatial synchrony leads to populations cycles  
Author Contributions: TLA and DCR conceived the project; RER provided assistance with data curation; TLA, DCR and LWS analyzed the data; TLA and DCR wrote manuscript; all authors provided extensive editing and feedback on the analysis and manuscript.  
Type of Article: Letter  
Key Words: population cycle, synchrony, Moran effect, population dynamics, white-tailed deer  
Number of words in abstract: 150  
Number of words in main text: 4999  
Number of figures and tables: 5 figures, 1 table  
Number of references: 58  
<!--***DAN: don't forget to fill this stuff in--> <!--***Tom: will double check for hte final submission-->

\textbf{Data accessibility.} All data will be made available in the Dryad digital repository. 
<!--***DAN: Tom to mention in the cover letter that data will be so deposited when the paper
is accepted, and Tom also to replace XYZ with the repo number when the time comes, please.-->
All data and code producing this paper are also archived at `www.github.com/reumandc/TomAnderson_WIdeer`.
Analyses leading to the paper can be re-run and the paper re-compiled 
via the codes there - see `Readme.md` in the github repository for instructions.

\newpage
<!--Title length limit 96 characters, inc spaces-->

```{r echo=F, results="hide"}
#load in results
options(scipen=6)
Tab1_results<-readRDS(file = "Results/Table1.rds")
Tab1_results<-cbind(Tab1_results[,1:3],apply(Tab1_results[,4:7],2,function(x){round(x,4)}))
Tab1_MT<-Tab1_results[Tab1_results$Timescale=="3-7",]

#make synchrony explained and cross terms NA for non-spatial drivers
for(i in 1:nrow(Tab1_results)){
  if(grepl("Winter",Tab1_results$Predictor[i])||grepl("Summer",Tab1_results$Predictor[i])){
    Tab1_results$SynchronyExplained[i]<-NA
    Tab1_results$CrossTerms[i]<-NA
  }
}
Tab1_MT<-Tab1_results
abunwmeiSyncExp37<-round(Tab1_results$SynchronyExplained[Tab1_results$Predictor=="Winter MEI"&Tab1_results$Timescale=="3-7"],2)
abunwpdoSyncExp37<-Tab1_results$SynchronyExplained[Tab1_results$Predictor=="Winter PDO" &  Tab1_results$Timescale=="3-7"]
abunwmeiSyncExp47<-round(Tab1_results$SynchronyExplained[Tab1_results$Predictor=="Winter MEI"&Tab1_results$Timescale=="4-7"],2)
abunwpdoSyncExp47<-Tab1_results$SynchronyExplained[Tab1_results$Predictor=="Winter PDO" &  Tab1_results$Timescale=="4-7"]
snowabunSyncExp3_4<-Tab1_results$SynchronyExplained[Tab1_results$Predictor=="Snow Depth" & Tab1_results$Timescale=="3-4"]
snowabunSyncExp3_7<-Tab1_results$SynchronyExplained[Tab1_results$Predictor=="Snow Depth" & Tab1_results$Timescale=="3-7"]
abunmodelSyncExp<-readRDS(file="Results/abunmodelSyncExp.rds")

#load Table S1 results
TabS1_results<-readRDS(file="Results/TabS1_results.rds")
wmei_wpdo_pval<-round(TabS1_results$Pvalue[TabS1_results$Predictor=="Winter MEI" & TabS1_results$Response=="Winter PDO"],4)
wmei_wpdo_phase<-round(TabS1_results$MeanPhase[TabS1_results$Predictor=="Winter MEI" & TabS1_results$Response=="Winter PDO"],4)
abunwmeiSyncExp47<-Tab1_results$SynchronyExplained[Tab1_results$Predictor=="Winter MEI" &  Tab1_results$Timescale=="4-7"]

#load Table S2 results
TabS2_results<-readRDS(file = "Results/TableS2.rds")
usda_dvcabunSyncExp<-TabS2_results$SynchronyExplained[TabS2_results$Response=="DVCs" &
                                                        TabS2_results$Predictor=="Abundance"]
usda_adjdvcabunSyncExp<-TabS2_results$SynchronyExplained[TabS2_results$Response=="Traffic-adj. DVCs" &
                                                        TabS2_results$Predictor=="Abundance"]
usda.snowabunSyncExp3_4<-TabS2_results$SynchronyExplained[TabS2_results$Response=="Abundance" &  TabS2_results$Predictor=="Snow Depth" & TabS2_results$Timescale=="3-4"]
avgdeerfluc<-readRDS(file="Results/totaldeerfluctuations.rds")
avgdvcfluc<-readRDS(file="Results/totaldvcfluctuations.rds")
avgdvccostsLow<-readRDS(file="Results/totaldvccostsLow.rds")
avgdvccostsHigh<-readRDS(file="Results/totaldvccostsHigh.rds")
```

#Abstract
Population cycles are fundamentally linked with spatial synchrony, the
prevailing paradigm being that populations with intrinsically cyclic dynamics are easy to synchronize. 
That is, population cycles help give rise to spatial synchrony. Here we demonstrate this process can work in reverse, 
with synchrony causing population cycles. We show that timescale-specific environmental effects, by synchronizing local 
population dynamics on certain timescales only, cause major population cycles over large areas in white-tailed deer. An 
important aspect of the new mechanism is specificity of synchronizing effects to certain timescales, which causes local 
dynamics to sum across space to a substantial cycle on those timescales. We also demonstrate, to our knowledge for the 
first time, that synchrony can be transmitted from environmental drivers to populations (deer) and from there to human 
systems (deer-vehicle collisions). Because synchrony of drivers may be altered by climate change, 
changes to population cycles may arise via the new mechanism.

<!--Authors must submit a 120-word-maximum statement about the significance of their research paper written at a level understandable to an undergraduate-educated scientist outside their field of specialty. The primary goal of the Significance Statement is to explain the relevance of the work in broad context to a broad readership. Significance statements are not required for Brief Reports. TOM: My first attempt below is currently 140 words-->
<!-- **Significance Statement:** Understanding the mechanisms behind cycles in population abundances has long been a goal in ecology. Here we demonstrate, in white-tailed deer, a new mechanism by which population cycles can arise, and the consequences of cycles for humans. Fluctuations in deer abundances in different counties of Wisconsin, USA, were highly correlated due to correlated fluctuations in winter climate. Importantly, these synchronized dynamics occurred only on 3-7-year timescales. Deer synchrony in turn produced synchronized oscillations in deer-vehicle collisions across Wisconsin, again with 3-7-year periodicity. When aggregated spatially, fluctuations for individual counties reinforced each other on 3-7-year timescales, producing state-wide cycles. In addition to revealing new mechanisms, our results are important for hunting, insurance industries, and ecosystems influenced by deer.    -->

#Introduction

The causes of population cycles have mystified ecologists for over 100 years 
[@Myers2018_cycles]. An early hypothesis for the cause of cycles, for the iconic Canadian 
lynx-snowshoe hare system and 
for other systems, posited that cyclic environmental phenomena such as sun spots drive 
population cycles [@Sinclair1992_sunspots;@Selas2004_sunspots]. However, that hypothesis lacked 
empirical support [@Nilssen2007_nosunspots;@Myers2013_nosunspots], and for decades 
research has focused instead on explanations involving nonlinear or delayed density dependence, 
often via predator-prey interactions, parasitism, and disease 
[@Krebs2018_harereview;@Krebs2013_book;@Myers2013_nosunspots;@MartinezPadilla2014_grouse]. 

Population cycles are fundamentally linked with spatial synchrony, the 
tendency for geographically distinct populations to 
exhibit correlated temporal fluctuations. 
Theory predicts that cyclic populations can easily be brought into synchrony via weak 
coupling from dispersal or via shared exogenous forcing of weather (Moran effects), in a 
process known as nonlinear phase locking [@Bjornstad2000]. Empirical evidence from laboratory 
experiments [@VassuerFox_2009;@HopsonFox_2019] and observational data 
[@Haynes_etal2019] support these predictions. Indeed, the 
lynx-hare system features predator-prey cycles thought to be synchronized regionally 
by phase locking through Moran effects [@stenseth1999;@stenseth2004]. 
As a consequence, the prevailing paradigm is that cycles and synchrony are linked because 
populations whose dynamics are intrinsically cyclic are easy to synchronize; that is, population 
cycles help give rise to spatial synchrony.

We here propose and provide evidence for a mechanism by which the direction of causality can be 
reversed: instead, spatial synchrony can give rise to population cycles. The
mechanism comprises the following two steps. First, an exogenous driver synchronizes non-cyclic local
population fluctuations *on a particular timescale*. Many large climate phenomena
fluctuate, either episodically or consistently, with emphasis on particular timescales. 
For example, from ca.
1965-1995, fluctuations in the North Atlantic Oscillation (NAO) contained a substantial 
10-year oscillatory component 
[@Sheppard2016_aphids], and the El Ni$\~{n}$o Southern Oscillation (ENSO) also has 
periodic content [@tudhope2001]. Thus local weather variables influenced by 
climatic oscillations can contain periodic components which are synchronous across
large areas. These components may be weak and therefore hardly noticable, 
in comparison with asynchronous,
local fluctuations with which they are combined.
Synchronous periodic components in local weather can nevertheless propogate, via Moran
effects, to produce weak, synchronous periodic components 
in local population dynamics. Evidence
indicates that timescale-specific spatial synchrony of populations is probably common 
[@Defriez2016_plankton;@Sheppard2016_aphids;@Defriez2017_marine;@Defriez2017_terrestrial;@Sheppard2017_surrogates;@Walter2017_geogsync;@Anderson2019_LO;@Sheppard_wavmodel;@Haynes_etal2019]. 
Second, the weak periodic components
of local population dynamics, being synchronous across space, reinforce each other
in the total population computed across the region, whereas stronger, asynchronous
components on other timescales cancel in the spatial total. Thus periodicities
in dynamics which may be unnoticable at the local spatial scale become
dominant at the regional scale. 
The key here is that synchrony occurs *on a particular timescale*.
Dynamics of the total, regional population are then dominated by a cycle on that timescale. See
Fig. \ref{Fig1_example} for a simulated example. A theoretical approach to the mechanism is summarized below and elaborated in
Supporting Information (SI) section \ref{SM-SI:theory}. The bulk of the paper is 
spent examining empirical evidence for the mechanism.

One major applied concern is that if
fluctuations in local time series do become amplified in a
timescale-specific manner as outlined above, outbreaks of pests or disease
could follow [@Rohani_etal1999]. For example, timescale-specific
synchrony of local populations of pests may produce major cycles in the
regional total pest population and therefore in the damage they cause. 
Timescale-specific
synchrony could also be important for industries, such as hunting,
that exploit multiple populations across space if it causes cycling in
the regional total population. 
For instance, if the time series of Fig. \ref{Fig1_example}D-F were deer populations 
in counties within a state so that 
Fig. \ref{Fig1_example}J-L are state-total populations, differences between the 
three scenarios of Fig. \ref{Fig1_example} could
impact the hunting industry.
Such cycles may then also cause cascading effects 
on human economic and social systems. Cascading effects of synchrony
on human concerns, though potentially important, are insufficiently 
studied. 

The ecological system for which we explore the importance of these ideas 
centers on white-tailed deer (*Odocoileus virginianus*) populations across
the state of Wisconsin (WI), USA, and on rates of deer-vehicle collisions (DVCs).
White-tailed deer is a species of considerable interest due to its
economic and cultural benefits and ecosystem impacts [@Cote2004_deerimpacts;@Conover2011_bookchp]. Across 
the USA in 2011, US\$$34$Bn were spent on hunting big game, including deer [@usfws2011_survey]. 
In WI, hunting-related economic activity is ~$1\%$ of state GDP [@usfws2011_survey]. DVCs cause about 
50 human fatalities and US\$$1.5$Bn in vehicular damage annually across the USA [@Conover2011_bookchp]. Thus it 
is an important applied question what drives the year-to-year variability in statewide deer abundance and DVC 
totals. We consider potential drivers of deer populations including both environmental factors and the direct 
effects of hunting. We use data from 1981-2016 (Fig.  \ref{Fig2_Timeseries}) aggregated at three 
spatial resolutions: counties ($n=71$), United States Department of Agricultural (USDA) districts 
($n=9$), and statewide (*Methods*). 

Here we demonstrate, using deer data, how timescale-specific
spatial synchrony can produce regional population cycles. We also show that
timescale-specific deer population synchrony begets timescale-specific synchrony
in DVCs across the state, producing cycles in 
state-total DVCs as well. Fluctuations in deer population abundances in
the 71 counties of Wisconsin were synchronous due to
fluctuations in winter climate, and synchrony 
occurred mainly on 3-7 year timescales. Deer synchrony in turn
produced synchrony in DVCs on 3-7 year timescales. 
Timescale-specific synchrony caused large fluctuations on 3-7-year
timescales in state-total deer populations and DVCs. 
Our results provide a theoretical description and empirical support for 
the importance of a new mechanism of population cycles, and for its
potential importance for human concerns. In contrast to the well-understood
mechanism by which cycles can beget synchrony, our mechanism shows 
that synchrony, when specific to certain timescales, can also produce
cycles.

# Summary of theory

The statement of our main theoretical result requires a conceptual understanding 
of the *power spectrum* (or, simply, *spectrum*) of a population or other 
time series, and of the *cospectrum* of two time series. 
The spectrum is useful for studying population cycling because peaks in the 
spectrum of a population indicate cyclic dynamics. Given a time series
or stochastic process, $w_i(t)$, measured in location $i$, the spectrum $S_{w_iw_i}(f)$ 
is a function of frequency, $f$. The frequency of a periodic oscillation is one over
its timescale, i.e., one over its period. The value of the spectrum $S_{w_iw_i}(f)$ is larger for 
frequencies on which $w_i(t)$ oscillates with greater strength. So, 
for instance, a population that exhibits strong (respectively, weak) 10-year-timescale 
oscillations will have a spectrum with a large (respectively, small) peak at 
the frequency $1/10 \text{yr}^{-1}$. The spectrum decomposes the variance, $\var(w_i(t))$, according 
to the timescales on which it occurs, in the sense that an appropriate integral of 
$S_{w_iw_i}(f)$ across all frequencies equals $\var(w_i(t))$. The cospectrum
of two time series, $w_i(t)$ and $w_j(t)$, decomposes their covariance, $\cov(w_i(t),w_j(t))$,
in a similar way: the cospectrum, $C_{w_iw_j}(f)$, is a function of frequency
with peaks occurring at frequencies for which oscillations in $w_i(t)$ and
$w_j(t)$ are strong and coincident in their phases, i.e., synchronous; an 
appropriate integral of $C_{w_i w_j}(f)$ equals $\cov(w_i(t),w_j(t))$.

Our main theoretical result expresses the spectrum, $S_{\text{tot}}$, of a total
population across a region composed of $i=1,\ldots,N$ local areas in the form
\begin{equation}
S_{\text{\emph{tot}}}(f) = F_{\text{popdyn}}(f) \left[ \sum_{i=1}^N S_{\epsilon_{i}\epsilon_{i}}(f) +
\sum_{\substack{i,j=1 \\ i \neq j}}^N C_{\epsilon_{i}\epsilon_{j}}(f)\right],\label{mainresmt}
\end{equation}
\noindent where $F_{\text{popdyn}}(f)$ reflects 
the nature of population dynamics within the local areas, and the $S_{\epsilon_i\epsilon_i}$
are spectra and the $C_{\epsilon_i\epsilon_j}$ are cospectra of environmental noise time 
series in the local areas. The derivation of this result is performed for a model exhibiting 
autoregressive moving average (ARMA) dynamics 
within a habitat patch, and $F_{\text{popdyn}}(f)$
is a function that depends only on the ARMA coefficients (SI section \ref{SM-SI:theory}). 
If dynamics are not intrinsically cycling, 
then $F_{\text{popdyn}}(f)$ has no peaks and does not contribute to whether $S_{\text{tot}}$
is peaked. The term $\sum_{i=1}^N S_{\epsilon_{i}\epsilon_{i}}(f)$ in (\ref{mainresmt}) 
encapsulates the effects of 
periodicities in local environmental noise time series on whether the total 
population exhibits cycling. The term 
$\sum_{i \neq j} C_{\epsilon_{i}\epsilon_{j}}(f)$ in (\ref{mainresmt}) encapsulates
the effects of timescale-specific synchrony on whether the total population
exhibits cycling. The key observation here is that, when $N$ is large so that the
region being considered is a combination of a large number of local areas, the first term
is likely to make a much smaller contribution to $S_{\text{tot}}$ than is the second
term; the first term is a sum of only $N$ quantities and the second is a sum of
$N^2-N$ quantities, and $N^2-N$ is much greater than $N$, for large $N$. Thus theory predicts
that timescale-specific patterns of synchrony are very important for whether
spatial-total populations exhibit cycling. Of course, which of the two terms
in brackets in (\ref{mainresmt}) is larger depends not only on the number of summands
in each term, but also on the values of the summands. We explore 
reasonable scenarios in the SI
and find that the initial intuition is correct, for a general model setup, 
that timescale-specific synchrony is a very important determinant of
regional population cycles. See SI sections \ref{SM-SI:theory}-\ref{SM-sect:mainproof}
for the the general theory and 
SI section \ref{SM-SI:theory} and Figs \ref{SM-TheoryExample1}, \ref{SM-TheoryExample2},
\ref{SM-TheoryExample3} for examples.

# Methods

\textbf{Empirical Data.} Using data provided by the Wisconsin Department of Natural Resources 
(WIDNR), we calculated annual deer abundance estimates for the `r dim(cty.list$Abun)[2]`-yr 
period `r minyear`-`r maxyear` for each of `r length(unique(dat$COUNTY_NAM))` counties in WI
(Fig. \ref{Fig2_Timeseries}). We also calculated estimated hunter abundance during the first 
two days of the annual 9-day gun hunting season [@widnr2001], again using WIDNR data, for each 
county for 1992-`r maxyear` (Fig. \ref{Fig2_Timeseries}). These data were an index of 
hunting pressure. We obtained estimates of DVCs for each county, for 1987-`r maxyear`, 
from the WI Department of Transportation (WIDOT; Fig. \ref{Fig2_Timeseries}). DVC data reflect 
the numbers of police-investigated DVCs. They are an 
index of DVCs, as opposed to an absolute number, 
since not all DVCs are reported [@Marcoux2010_dvc]. Finally, traffic volume estimates 
(millions of miles 
driven), for each county and for 1988-`r maxyear` were obtained from the WIDOT, to control for traffic in our 
analysis of DVCs.

We obtained daily weather data (snow depth, 
precipitation amount, and minimum and maximum air temperature - here called Tmin and Tmax), from all available 
weather stations in each county, from the 
National Climatic Data Center ([`https://www.ncdc.noaa.gov/`](https://www.ncdc.noaa.gov/)). We 
averaged daily values across the winter ($\text{Dec}_{t-1}$ 
to $\text{Mar}_t$) and across stations within a county to produce 
annual time series of winter weather by county (e.g., Fig. \ref{Fig2_Timeseries}). 
We also calculated a Winter Severity Index (WSI), the total number of days for 
which snow depth was greater than 18 inches or Tmin was less than $0^\circ$F, for each county 
and year. WSI is known to relate to winter impacts on deer populations [@widnr2001].

Monthly values of indices of three large-scale climate phenomena were obtained: 
the North Atlantic Oscillation (NAO), the multivariate 
El Ni$\~{n}$o Southern Oscillation Index (MEI), and the Pacific 
Decadal Oscillation (PDO). All of these have been shown to 
influence large ungulate population dynamics in the north central 
USA [@Post1998_popgrowth;@Post2004_greenland]. 
For each index, we averaged monthly values across the 
approximate growing season (Apr-Nov), and, separately, across the winter 
($\text{Dec}_{t-1}$ to $\text{Mar}_t$; Fig. \ref{Fig2_Timeseries}). High 
values of these indices generally correspond to mild winter weather over 
WI [@Rodionov2003_elnino]. A total of 11 weather and climate variables were thus 
used: winter Tmin, Tmax, snow depth, and precipitation; winter NAO, MEI, and PDO; 
growing-season NAO, MEI and PDO; and the WSI. 

\textbf{Data pre-processing.} We performed all analyses described below 
at the three spatial scales mentioned in the *Introduction*: counties (N = 44-`r dim(cty.list$Abun)[1]` depending on 
the variables involved, see SI section \ref{SM-SI:methods}; mean county 
area $\pm$ SD = `r round(mean(gps$Area_km2),0)` $\pm$ `r round(sd(gps$Area_km2),0)` $\text{km}^2$), 
USDA districts ($N = 9$; mean 
district area $\pm$ SD = `r round(mean(usda.centroid$Area_km2),0)` $\pm$ `r round(sd(usda.centroid$Area_km2),0)` $\text{km}^2$), 
and statewide (SI Fig. \ref{SM-FigS1_map}). Analyses at three spatial scales 
were performed because of concerns about the accuracy, on small spatial scales, of deer abundance estimation methods
used by the WIDNR and underlying our data; they used a standard "Sex-Age-Kill" (SAK) approach.
Although this is a standard approach used by management agencies for abundance assessments,
its accuracy at the smallest spatial scales was once questioned
(@Millspaugh2009; see SI section \ref{SM-SI:methods} for further discussion).
USDA district data were 
the sum or average of county-level variables, as districts were always conglomerates of 
counties (SI Fig. \ref{SM-FigS1_map}); likewise for statewide data. 
Results remained unchanged in their main substance using larger 
spatial scales (see *Results*),
indicating results for counties were probably not meaningfully 
influenced, for our purposes, by possible shortcomings of the SAK procedure at the county level.

Each time series was Box-Cox transformed to help ensure Gaussian marginal 
distributions [@Sheppard2016_aphids], which are necessary for
the Fourier-surrogate-based spatial wavelet coherence and wavelet linear
regression testing procedures described below. This process also included linear detrending
(and subtracting the mean), to remove trends that might otherwise
obscure patterns of synchrony [@Buonaccorsi2001], as well as variance standardization 
(SI section \ref{SM-SI:methods}).

<!--***DAN: You probably want to add some material in the below para or nearby explaining 
why it makes sense to use Fourier for the theory and wavelets for the empirical analysis. 
Maybe add a few basic statements in the main text around here, a few more basic statements 
at the end of the "Analysis" subsection, maybe one more summary statement somewhere in the 
Results, and then a sup mat section elaborating. I have done some of this but not all.-->
\textbf{Analysis.} We use wavelet transforms in most analyses described below,
complemented by a few classic Fourier approaches.
Wavelets are ideal for detecting periodic components in time series data
[@addison2002]. Our methods we use are extensions of methods that are
well established, standard tools in many biological fields. 
Therefore, we provide in the main text an operational description, only,
of each method, i.e., what it takes as input, what it provides as output, and how to
interpret the output. Further descriptions and technical details are in SI 
section \ref{SM-SI:methods}, and complete details of all methods are published elsewhere
[@Sheppard2013;@Sheppard2016_aphids]. It is explained in SI section \ref{SM-SI:methods}
why it was an appropriate choice to use Fourier methods for the theoretical
development (above and in SI sections \ref{SM-SI:theory}-\ref{SM-sect:mainproof}), and 
to use a combination of wavelet and Fourier methods
for the empirical analyses (below). 
<!--***DAN: Make sure to do what it says I will do in the above sentence-->
All wavelet analyses were done via software in the open-source `wsyn`
package for the R programming language [@wsyn], which is itself well documented.
<!--***DAN: The perspective we used to use above was that we explain methods operationally and
the reader has to look elsewhere. But the referee does not like that, so I changed the wording 
above, and then you have to go and change substance, i.e., add quite a bit more description of 
the techniques in the sup mat, at least. -->

We characterized the spatial synchrony of spatiotemporal variables (e.g., deer
abundance) using wavelet mean field magnitudes and wavelet
phasor mean field magnitudes (SI Fig.
\ref{SM-PedagogFigWMF}; @Sheppard2013, @Sheppard2016_aphids). Both methods take as input
standardized (see above) timeseries data from multiple locations and provide as
output plots which display the degree of synchrony among locations as a function
of time and timescale (Figs \ref{Fig1_example}G-I, SI 
Fig. \ref{SM-PedagogFigWMF}). Thus the plots reveal at what times, during the 
duration of data collection, and on what timescales, synchrony was most prevalent. 
The wavelet phasor mean field magnitude plot describes the degree 
of phase synchrony among time series, whereas the wavelet mean field magnitude plot describes
the degree of synchrony in both phases and amplitudes of oscillation.
The statistical significance of a wavelet phasor mean field magnitude plot, at any given
time and timescale, can be assessed though 
comparison to a null hypothesis of random phases (SI section \ref{SM-SI:methods}; @Sheppard_wavmodel). 
We used a significance threshold  of $P < 0.001$. 

For both county- and district-level data, we assessed pairwise relationships
between spatiotemporal variables using the spatial wavelet coherence technique
(SI Fig. \ref{SM-PedagogFigSpatCoh}; @Sheppard2013, @Sheppard2016_aphids, @Sheppard2017_surrogates). 
Spatial wavelet coherence tests for timescale-specific relationships between variables. 
Specifically, the technique tests whether two
spatiotemporal variables exhibit consistent phase differences and correlated
magnitudes of oscillation through time and across space, as a function of
timescale. Significance is determined by comparing spatial wavelet coherence values for
real data with spatial wavelet coherences made from appropriately randomized,
Fourier surrogate data sets ($N=$ `r nsurrogs`) that retain the same temporal and
spatial autocorrelation properties as the original time series but are otherwise
unrelated to each other [@Theiler1992;@Schreiber2000;@Sheppard2013;@Sheppard2016_aphids]. 
We calculated how often observed spatial wavelet coherence between pairs of variables was 
ranked higher than coherences of surrogate data sets, aggregating across the timescale 
band of interest using an
average-rank-based method described in detail elsewhere [@Sheppard2016_aphids]. A p-value 
of less than $0.05$ means that the average empirical spatial wavelet coherence rank was higher than 
the same quantity calculated for at least 95% of the surrogate data sets
which represent the null hypothesis of no relationship. Ranks of
actual spatial wavelet coherences in surrogate spatial wavelet coherences were
also plotted against timescale. 
Finally, when two variables had significant spatial wavelet coherence over a timescale
band, we calculated the mean phase difference between the variables over the band,
$\bar{\theta}$, in units of $\pi$ radians, to determine if oscillations had
approximately in-phase ($|\bar{\theta}| = 0$ to $0.25$), quarter-cycle
($|\bar{\theta}| =  0.25$ to $0.75$), or anti-phase ($|\bar{\theta}| =  0.75$ to $1$)
relationships. We tested spatial wavelet coherence between deer abundance and each of 
five weather variables (average winter
Tmin, Tmax, snow depth, precipitation, and WSI), six climate variables (winter and
growing season PDO, MEI and NAO) and hunter abundance, 
over the $3-7$-year timescale band, where the
majority of deer synchrony was present (see *Results*).
We also used spatial wavelet
coherence to test for relationships between deer abundance and DVCs, and between deer
abundance and DVCs adjusted for traffic volume.

<!--***DAN: I rewrote the following para so that it:
1) only states that we can calculate fraction of synchrony explained and predicted synchrony plots
for one-variable models with a spatiotemporal predictor. (OK, Plos Comp Biol used multiple predictors,
but we had no cause to that here, because we never had two spatiotemporal variables which were both significant,
and we wanted to keep it simpler - make this last point later, though, not here - Discussion?).
2) mentions cross terms in a similar way
3) mentions that use of non-spatial variables is complex, and involves inflation. We don't need it for the
main research goals, but see sup mat-->
We calculated percentages of deer synchrony explained by the synchrony of individual weather 
variables, as well the percentage of DVC synchrony explained by deer
synchrony, following the wavelet Moran theorem of @Sheppard2016_aphids.
<!--***DAN: Commented this follwoing text when we decided to not use multi-predictor models:
We also examined the percentage of deer synchrony jointly explained by snow depth, winter MEI, 
and winter PDO (the three primary drivers of deer synchrony, see *Results*) using
wavelet linear regression [@Sheppard_wavmodel]. This method is analogous to
multiple linear regression, with a model describing the relationship between a
response variable and multiple predictor variables [@Sheppard_wavmodel]. However,
all predictor and response variables are wavelet transforms of spatiotemporal
variables. -->
This approach also produces a time- and timescale-specific plot of
what the pattern of synchrony in the response variable (deer abundance, or DVCs) would look
like if the only driver of synchrony were the explanatory variable considered. The predicted synchrony
plot can then be visually compared to the actual wavelet mean field magnitude
plot (of deer or DVCs), giving a visual indication of how well synchrony can be accounted for
by the predictor considered [@Sheppard_wavmodel]. As part of these procedures, we evaluated 
model "cross terms", which are the component of  synchrony that is due to variability 
explained by the drivers at a given location being synchronous with model residuals at other 
locations [@Sheppard2016_aphids]. Cross terms must be small ($<10\%$) 
in order for inferences about synchrony to be valid (SI section \ref{SM-SI:methods}).  
The approach described here is most straightforwardly applied to predictors which 
are spatially as well as temporally
resolved, e.g., local weather variables as opposed to climate indices such as 
MEI. Attempts to use these techniques to quantify the fraction of deer synchrony explained by
MEI were more complex, produced estimates which were probably inflated, and were not needed 
for our main research goals, so were relegated to the SI (sections \ref{SM-SI:methods} for
methods and \ref{SM-SI:results} for results).
<!--***DAN: don't forget to add this material-->

To quantify the degree to which synchronizing effects are responsible for
fluctuations (cycling) in statewide total deer and DVC time series, we adapted the standard
Fourier surrogate procedure of @Theiler1992. We constructed artificial
time series data, for both variables for each location, that had the same spectral
characteristics as the original data, except that they were not spatially synchronized over
the $3-7$-year timescale band of interest. We accomplished this by randomizing
phases of the Fourier transforms of the county-level time series 
(SI section \ref{SM-SI:methods}). We summed each set of
surrogate time series spatially to produce surrogate statewide total time series for both
deer and DVCs, representing the hypothetical case of no synchrony over $3-7$-year
timescales. We then compared these surrogate time series to the 
actual state-total deer abundance and DVC time series to determine to what extent 
additional variance and cycling intensity in the
empirical state-total time series were attributable to synchrony. Further details of all datasets,
data pre-processing, and analytic methods are in SI section \ref{SM-SI:methods}.

<!--***DAN: Added a section here on the fact that we also carried out Fourier analyses.
Need to correspondingly mention the Fourier results at the end of the Results section-->
Although wavelet tools were the most appropriate appropriate 
overall choice for our research goals, some of the main components of our theoretical 
mechanism can be detected in our data using Fourier analyses which may be more familiar to
some readers than wavelets. Supplementary Fourier approaches we used are described in 
SI section \ref{SM-SI:methods}. These approaches
are used, for instance, to examine whether: 1) the spectral power in the state-total deer time series
in the $3-7$-year timescale band is greater than it would without synchrony in that band; and
2) the same spectral peaks in the $3-7$-year band do not occur in county-level deer time series.

\textbf{Data availability.} All data will be made available in the Dryad digital repository. 
<!--***DAN: Tom to mention in the cover letter that data will be so deposited when the paper
is accepted, and Tom also to replace XYZ with the repo number when the time comes, please.-->
All data and code producing this paper are also archived at `www.github.com/reumandc/TomAnderson_WIdeer`.
Analyses leading to the paper can be re-run and the paper re-compiled 
via the codes there - see `Readme.md` in the github repository for instructions.

# Results

The wavelet mean field of county deer abundance time series revealed that the most prominent 
synchrony occurred at $3-7$-year timescales (Fig. \ref{Fig3_DeerSync}A). The wavelet 
phasor mean field showed the strong statistical significance of this phenomenon (Fig.
\ref{Fig3_DeerSync}B). We therefore focused subsequent analysis on the $3-7$-year timescale band.
Wavelet mean field and wavelet phasor mean field magnitude plots for all
winter weather variables are in Fig. \ref{SM-FigS8_WMFsAndWPMFsAncillary}.

Spatial wavelet coherence demonstrated that 
deer abundance synchrony on $3-7$-year timescales was partly caused by climatic Moran effects. 
Synchrony in deer abundance was related to winter snow depth 
(spatial wavelet coherence, $P =$ `r round(Tab1_results$Pvalue[Tab1_results$Predictor=="Snow Depth" &Tab1_results$Timescale=="3-7"],2)`, 
Table \ref{Tab1_CountyCoherenceSummary}), in a near anti-phase relationship (mean phase 
difference between abundance and snow depth across $3-7$-year timescales, in units of ${\pi}$ 
radians, was $\bar{\theta} =$ `r round(Tab1_results$MeanPhase[Tab1_results$Predictor=="Snow Depth"&Tab1_results$Timescale=="3-7"],2)`; SI Fig. \ref{SM-FigS4_PhaseDiffs}D). Thus deer 
abundances tended to peak in years of below-average snow. No other local weather variables were statistically significantly related with deer abundance (Table \ref{Tab1_CountyCoherenceSummary}). 
Deer abundance was also related, with
approximately quarter-cycle phase shifts, 
to the multivariate El Ni$\~{n}$o Southern Oscillation index
($P =$ `r round(Tab1_results$Pvalue[Tab1_results$Predictor=="Winter MEI"&Tab1_results$Timescale=="3-7"],2)`, $\bar{\theta}  =$ `r round(Tab1_results$MeanPhase[Tab1_results$Predictor=="Winter MEI"&Tab1_results$Timescale=="3-7"],2)`) and to the Pacific Decadal Oscillation 
($P =$ `r round(Tab1_results$Pvalue[Tab1_results$Predictor=="Winter PDO"&Tab1_results$Timescale=="3-7"],2)`, $\bar{\theta}  =$ `r round(Tab1_results$MeanPhase[Tab1_results$Predictor=="Winter PDO"&Tab1_results$Timescale=="3-7"],2)`, 
Table \ref{Tab1_CountyCoherenceSummary}, SI Fig. \ref{SM-FigS4_PhaseDiffs}A,C), both averaged over the winter months
($\text{Dec}_{t-1}$ to $\text{Mar}_t$, abbreviations $\text{MEI}_{\text{w}}$ and
$\text{PDO}_{\text{w}}$). The two indices were strongly coherent and in-phase with
each other on $3-7$-year timescales over the period 1981-2016 (wavelet coherence, $P =$ `r round(TabS1_results$Pvalue[TabS1_results$Predictor=="Winter MEI" & TabS1_results$Response=="Winter PDO" ],4)`, $\bar{\theta}  =$ `r round(TabS1_results$MeanPhase[TabS1_results$Predictor=="Winter MEI" & TabS1_results$Response=="Winter PDO" ],2)`, SI Table \ref{SM-TabS1_PairwiseClimVarSpatCoh}). 
At $5$-yr periodicity, the quarter-cycle phase lags between deer abundance and $\text{MEI}_{\text{w}}$/$\text{PDO}_{\text{w}}$
correspond to deer populations
peaking about `r round(-5*Tab1_results$MeanPhase[Tab1_results$Predictor=="Winter PDO"&Tab1_results$Timescale=="3-7"]/2,1)` or 
`r round(-5*Tab1_results$MeanPhase[Tab1_results$Predictor=="Winter MEI"&Tab1_results$Timescale=="3-7"]/2,1)` years after mild winters. 
Significant spatial wavelet coherence with an environmental variable which is itself spatially synchronous 
(e.g., snow depth or some combination of weather variables related to MEI and PDO) is evidence of transmission 
of synchrony from the environment to deer (SI section \ref{SM-SI:methods}; @Sheppard2016_aphids, @Sheppard_wavmodel). 
NAO was not related to deer abundance synchrony (Table \ref{Tab1_CountyCoherenceSummary}).
<!--***DAN: you need to add some text to SI methods arguing the above statement, you can rephrase from the aphid paper-->

Snow depth and climate index results seemed to 
indicate distinct or partly distinct phenomena 
synchronizing deer on different timescales within the $3-7$-year range.
Deer dynamics at $3-4$-year timescales were not significantly spatially coherent with either 
$\text{MEI}_{\text{w}}$ or $\text{PDO}_{\text{w}}$, which instead were strongly coherent with
deer on $4-7$-year timescales (Table \ref{Tab1_CountyCoherenceSummary}, see also 
Fig. \ref{SM-FigS5_RanksInSurrogs}A and Table \ref{SM-TabS2_SummaryDistrict}).
In contrast, spatial wavelet coherence between snow depth and deer was 
not significant on $4-7$-year timescales, but was significant on $3-4$-year timescales
(Table \ref{Tab1_CountyCoherenceSummary}). 
Applying the wavelet Moran theorem, snow depth explained $`r round(snowabunSyncExp3_7,0)`\%$ of 
synchrony in deer abundance at $3-7$-year timescales, but explained $`r round(snowabunSyncExp3_4,0)`\%$ of synchrony in deer
<!--***DAN: Tom, please make the numbers in the above line be autolinks--><!--TOM: DONE-->
abundance at $3-4$-year timescales (Table \ref{Tab1_CountyCoherenceSummary}; see also
USDA district level results for the $3-4$-year band, Table \ref{SM-TabS2_SummaryDistrict},
which were similar but had smaller cross terms). The timescale
bands $3-4$ and $4-7$ were chosen by examination of Fig. \ref{SM-FigS5_RanksInSurrogs}.
Estimates of percentages of synchrony in deer explained by $\text{MEI}_{\text{w}}$
were complicated, and also were artefactually inflated, by the fact that $\text{MEI}_{\text{w}}$ is 
not a spatially resolved variable (*Methods*). Furthermore, such results are not strictly necessary
to substantiate our argument that deer are environmentally synchronized in the $3-7$-year
band, since we already presented results for snow depth, and results showing signficant 
wavelet coherence between deer and MEI. Thus such estimates are relegated to the SI (SI section
\ref{SM-SI:results}), where attempts to quantify the degree of inflation are also described.
<!--***DAN: Don't forget to add the material described above-->
<!--***DAN: Tom, can you please make the following changes to Table 1:-->
<!--1) add back in the column for timescale band--><!--TOM: Done-->
<!--2) put all the results for timescale bands other than 3-7 back into the table - these are in table S1 now--><!--TOM: All results are back in if P < 0.05 for 3-7ts-->
<!--3) remove all the results from the "synchrony explained" and "average cross terms" columns for cases where
the predictor is not a spatially resolved variable. So those columns will be mostly empty, only having 
values for the deer v snow depth results, some of the deer v hunters results, and the DVC results--><!--TOM: made sync explained and xterms both blanks-->
  <!--So this will make Table 1 look a lot like Table S1. We may well no longer need Table S1 (I think we won't,
but you should check by searching for citations to S1 in both the main text and sup mat, verifying that these
could all be replaced by citations to the newly revised Table 1, and then also replacing all those
citations, please), in which case get rid of it.--><!--TOM:-->
  <!--Tom, can you also remove panel C from Fig. 3, please. Keep the code, though, in case we need it.--><!--TOM: Done-->
 <!-- Finally, Tom, you will also have to remove the numbers from Table S3 in the "synchrony explained" and 
"average cross terms" columns that correspond to cases where
the predictor is not a spatially resolved variable.--><!--TOM: Done-->
  <!--Finally-finally, you will have to remove Fig. S11E, though you can keep S11F. You'll have to relabel 
the panels. The caption will have to be changed, but you can leave that to me if you want.--><!--TOM: Removed the figure, need to update caption yet-->
-->
<!--***DAN: Dan, you probably want to add something to the discussion that coherence with MEI/PDO (together
with our results for snow depth) is enough to support our argument that deer are environmentally 
synchroinzed in the 3-7-year band, but we still do not know what specific weather variables (other
than snow depth) are acting here, and for which MEI serves as a surrogate. This could be investigated in
future work-->

Deer abundance synchrony on $3-7$-year timescales was not caused by hunting. County-level hunter and deer abundances were not significantly spatially coherent 
($P =$ `r round(Tab1_results$Pvalue[Tab1_results$Predictor=="Hunters" & Tab1_results$Timescale=="3-7"],2)`) over $3-7$-year timescales 
(Table \ref{Tab1_CountyCoherenceSummary}, SI Figs \ref{SM-FigS5_RanksInSurrogs}B, \ref{SM-FigS6_WMFHunter}). Hunter abundance was significantly related to deer abundance on shorter timescales, around 
$2-2.5$-years ($P =$ `r round(Tab1_results$Pvalue[Tab1_results$Predictor=="Hunters" & Tab1_results$Timescale=="2-2.5"],3)`), and was in approximate anti-phase with deer abundance 
on those timescales ($\bar{\theta}  =$ `r round(Tab1_results$MeanPhase[Tab1_results$Predictor=="Hunters" & Tab1_results$Timescale=="2-2.5"],2)`; 
Table \ref{Tab1_CountyCoherenceSummary}, SI Figs \ref{SM-FigS4_PhaseDiffs}G,
\ref{SM-FigS5_RanksInSurrogs}B). 
<!--***DAN: Tom, I changed what used to be a citation to Table S1, above, into a citation to Table 1, and I am here asking you to move the
2-2.5-year hunting coherence results into Table 1-->
But these short-timescale results, while interesting (*Discussion*), are not particularly
relevant for synchrony because deer showed limited synchrony on 
those timescales (Fig. \ref{Fig3_DeerSync}).

The wavelet mean field and wavelet phasor mean field magnitude plots for DVCs revealed time-and timescale-specific synchrony, and visual patterns were remarkably similar to those shown by deer (compare Figs \ref{Fig3_DeerSync}A,B and \ref{Fig4_DvcSync}A,B). Spatial wavelet coherence showed that deer abundance and DVCs were significantly related 
($P =$ `r round(Tab1_results$Pvalue[Tab1_results$Response=="DVCs" & Tab1_results$Predictor=="Abundance"],3)`, 
Table \ref{Tab1_CountyCoherenceSummary}) on $3-7$-year timescales, and had a nearly in-phase 
relationship ($\bar{\theta} =$ `r round(Tab1_results$MeanPhase[Tab1_results$Response=="DVCs" & Tab1_results$Predictor=="Abundance"],2)`, SI Fig. \ref{SM-FigS4_PhaseDiffs}E). 
Deer abundance explained $`r round(usda_dvcabunSyncExp,0)`\%$ of the synchrony in 
DVCs when data were aggregated to USDA districts (SI section \ref{SM-SI:results}, 
Fig. \ref{SM-FigS7_WMFAndPredictionsDistricts} and Table \ref{SM-TabS2_SummaryDistrict}). 
As a visual manifestation of this statistical result, patterns of synchrony in DVCs predicted by a
wavelet regression model with sole predictor deer abundance match the patterns actually present in DVC
synchrony (Fig. \ref{Fig4_DvcSync}C). 
Statistical significance, phase relationships, and the amount of DVC synchrony explained by deer
abundance remained similar after adjusting DVCs to account for
fluctuations in traffic volume (Table \ref{Tab1_CountyCoherenceSummary}, SI section \ref{SM-SI:results} and Table \ref{SM-TabS2_SummaryDistrict}, Fig. \ref{SM-FigS4_PhaseDiffs}F).

Statewide oscillations due to synchrony correspond to a typical difference of $~ `r round(avgdeerfluc,-4)`$ deer and $~ `r round(avgdvcfluc,-2)`$ 
DVCs between peak and trough years 
(Fig. \ref{Fig5_StateWideFluctuations}). Importantly, these high amplitude 
statewide oscillations in deer and DVCs disappear (Fig. 
\ref{Fig5_StateWideFluctuations}) when synchrony between counties on $3-7$-year 
timescales is removed via a randomization procedure (*Methods*). Therefore, it is spatial synchrony 
that causes local dynamics to sum to a substantial 
semi-periodic oscillatory phenomenon on statewide spatial scales. Depending on 
assumptions made about costs and reporting rates of DVCs (SI section \ref{SM-SI:methods}; @Huijser2008_dvc;@Marcoux2010_dvc;@Conover2011_bookchp), 
statewide DVC damages 
vary by US\$$`r floor(round(2*avgdvccostsLow,-6))/1000000`-`r floor(round(2*avgdvccostsHigh,-6))/1000000`$Mn 
between peak and trough years. 

Supplementary results using Fourier approaches (*Methods*, SI section \ref{SM-SI:methods}) which may be more familiar to
some readers than our wavelet tools, demonstrated clearly that: 1) there was a meaningful spectral peak in the
$3-7$-year timescale band for the state-total deer time series for WI, and it was substantially caused by 
synchrony because when $3-7$-year synchrony was removed from county-level time series the peak was removed;
2) the same spectral peaks did not occur in county-level deer time series (SI section \ref{SM-SI:results}).
Similar results also held for DVCs (SI section \ref{SM-SI:results}). 

# Discussion

<!--summary-->
Overall, we demonstrated that timescale-specific Moran effects, by 
synchronizing local population dynamics, can cause major cycles in the 
regional total population. This is a new mechanism by which the environment can 
produce population cycles. This new mechanism differs fundamentally from prior,
now refuted hypotheses that population cycles result from highly regular
environmental fluctuations [e.g., sun spots;
@Sinclair1992_sunspots;@Selas2004_sunspots] because the new mechanism does not
require a highly regular environmental 
oscillator. Rather, it is sufficient simply for irregular environmental drivers 
such as winter climatic factors to be coherent with populations over a particular timescale band, thereby inducing 
synchrony in that band and leading to total-population cycles on the same timescales.
We have also presented what is, to our knowledge, the first demonstration that synchrony 
can be transmitted in a 
Moran-like effect from ecological to human systems (deer to DVCs). Prior work demonstrated the
transmission of synchrony through trophic interactions within ecosystems [e.g., @Haynes2009],
but our results show similar mechanisms can also cause synchrony to pass from ecological
to human systems. Thus if climate change is altering synchrony, as a growing number
of papers indicate is likely occurring 
[@Post2004_greenland;@Allstad2015_gypsysync;@Defriez2016_plankton;
@Sheppard2016_aphids;@Koenig2016_birdsync;@Shestakova2016_forestsync;@Black2018], 
<!--***DAN: add the citation to the new Grotan review here-->
our results show that large-scale cycles
in both populations and human systems may change as a consequence.

<!--Variants of the mechanism. -->
There are two alternative elaborations of the mechanism we present by which
synchronous effects produce population cycles.
First, synchronous fluctuations in environmental drivers may have produced
small oscillations on $3-7$-year timescales in county-level deer population time series, and 
also brought these oscillations into spatial synchrony, thereby causing them to sum across
counties to produce noticable statewide fluctuations. Second, the local population oscillations 
may have been due partly to the intrinsic dynamics of deer, resulting 
from density-dependent effects, maternal effects, or 
other mechanisms of nonlinear population dynamics. Synchronous environmental drivers 
would then have brought the intrinsically oscillating local dynamics into synchrony. 
<!--This second hypothesis is the pattern typically described from theoretical and experimental studies, where exogenous drivers or dispersal couple intrinsically cyclic populations via nonlinear phase-locking [@Bjornstad2000;@VassuerFox_2009;@HopsonFox_2019]. Recent evidence from observational studies on gypsy moths also support the hypothesis of cyclic-dynamics-leading-to-synchrony [@Haynes_etal2019]. -->
The difference between these alternatives is whether local population
oscillations were produced and synchronized, or merely synchronized but 
not produced, by synchronous environmental drivers. But in either case our mechanism is the
same whereby Moran effects synchronize local dynamics in a timescale-specific manner,
and then local dynamics sum to produce population cycles.
The alternatives would be difficult to distinguish with certainty, but the former mechanism
may be more likely because winter climatic factors are known to influence ungulate
dynamics [@Mysterud2008_deerclimate; @Hegel2012_caribou; @Post2004_greenland; @Grotan2005_roedeer],
and we are unaware of any evidence suggesting that deer populations can be intrinsically
cyclic. Eliminating synchrony via randomizations eliminated the state-wide 
cyclic oscillation (Fig. \ref{Fig5_StateWideFluctuations}), even though the
randomizations preserved the structure of the local fluctuations, demonstrating that
it was, indeed, the synchrony that produced the observed large-spatial-scale cycling.

<!--Reasonableness of the drivers we found in relation to other studies and biological
mechanisms-->
Large-scale climatic patterns and local winter weather conditions are known to influence spatial synchrony 
via Moran effects in other ungulates, including reindeer, 
caribou [@post2002;@aanes2003;@Post2004_greenland;@Hegel2012_caribou], roe deer [@Grotan2005_roedeer], 
and musk oxen [@post2002], and thus the cycling mechanism we observed may occur as well 
in other ungulates. Here, winter snow depth,  $\text{MEI}_{\text{w}}$ and $\text{PDO}_{\text{w}}$ 
were the primary determinants of deer synchrony. Deer populations lagged $\text{MEI}_{\text{w}}$ and 
$\text{PDO}_{\text{w}}$ by 1-2 years, a similar delay to other reported lag times of climatic effects on deer 
populations [@Post1998_popgrowth;@forchhammer1998]. Winter weather and climate probably influenced deer
populations because deeper snow limits access to winter food resources in ungulates, which subsequently 
affects body condition, leading to greater susceptibility to starvation or predators and reduced 
parturition rates [@Mysterud2008_deerclimate;@Hegel2012_caribou]. 
The influence of winter weather on deer extends outside as well as inside WI,
so for larger regions including WI, cycling of total regional deer populations and 
DVCs are likely to be even stronger than what we have documented because
a larger number of synchronized local populations accentuates the effect we described.

<!--It aint hunters. -->
Hunters were not the principle driver of deer synchrony. These results are consistent with the
hypothesis that hunters can regulate deer populations over short timescales and small spatial 
areas [@fryxell1991;@denicola2008] but have more limited effects over longer timescales and at 
large spatial extents [@brown2000;@simard2013]. The lack of longer-timescale ($3-7$-year) spatial wavelet 
coherence between hunters and deer abundance is likely also related to the facts that hunters 
typically prefer to shoot bucks, the removal of
which has limited population effects; and the state regulates antlerless harvest in an attempt to
maintain populations at a desirable size.

<!--applied reasons to care, link to climate change-->
Our results provide an important but previously unrecognized applied motivation for studies 
of synchrony in populations.
Fluctuations of a quarter million deer, statewide, which we showed are caused by synchrony, 
likely will have considerable consequences for agricultural crop damage, forest regeneration 
[@Cote2004_deerimpacts], and hunting success, all of which can have socioeconomic impacts. 
We also showed that Moran effects subsequently pass from deer to DVCs in a cascading manner. 
Deer synchrony almost certainly caused DVC synchrony rather than the opposite causation,
because DVCs were too infrequent to
meaningfully drive deer abundance fluctuations (Fig. \ref{Fig2_Timeseries}), and if they did an anti-phase
relationship would have been found (as opposed to the in-phase relationship we observed).
<!--***DAN: Tom, I guess you made the change you did because you did not like the phrase "rather
than DVCs driving deer fluctuations" because of course DVCs do drive deer fluctuations
because most DVCs directly result in the death of a deer? I think it is important, for
clarity, to identify that we are contrasting with the opposite causal pathway, so I tried to
reword to bring that contrast back while avoiding this phrases which I guessed s what you
didn't like. If you want to reword again, please just keep in the specific identification 
of the alternative causal pathway that we are arguing did not happen.-->
Thus Moran effects are ultimately responsible for synchrony of DVCs, and synchrony 
causes fluctuations of more than 2500 annual documented DVCs through time.
<!--***DAN: In response to one of the referee comments, you may want to caveat the following 
sentence with something like "...though we have not examined whether a fluctuation of 2500
DVCs is sufficient, relative to other fluctuations that occur in accident rates, to be of concern..."-->
Such fluctuations may influence insurance and accident recovery industries.
<!--***Tom: It doesn't feel like this sentence is needed, and also was an abrupt transition that didn't flow well to me to I commented it out The influence of climate change on synchrony is probably important
[@Post2004_greenland;@Allstad2015_gypsysync;@Defriez2016_plankton;
@Sheppard2016_aphids;@Koenig2016_birdsync;@Shestakova2016_forestsync;@Black2018], but is still, 
in our opinion, greatly understudied.--> Our results provide another reason that
additional studies of climate-change influences on synchrony should be undertaken: 
because such influences may alter the cycling of populations, which can in turn
influence human activities. 

<!--specificity to timescales, and the strengths of our methods-->
An important aspect in all of the phenomena we explore is their specificity to 
certain timescales. Deer population cycles would not have manifested had deer synchrony 
not been specific to certain timescales ($3-7$-years).
Timescale specificity in synchrony is now know to be common 
[@Sheppard2016_aphids;@Sheppard2017_surrogates;@Walter2017_geogsync;@Sheppard_wavmodel;@Haynes_etal2019;@Anderson2019_LO].
Many past analyses of synchrony have 
used correlation methods [@Leibhold2004_sync] that conflate timescales, but this can hinder 
detection of synchronous phenomena and inferences of drivers of synchrony 
by obscuring synchrony on some timescales with 
asynchronous or noisy dynamics on other timescales [@Sheppard2016_aphids].
The analytic methods used here [@Sheppard2016_aphids;@Sheppard2017_surrogates;@Walter2017_geogsync;@Sheppard_wavmodel]
were able to isolate the timescales on which synchrony occurred ($3-7$ years),
facilitating more precise determination of drivers of synchrony.
Historically, determination of the causes of synchrony was identified as a
difficult problem [@Leibhold2004_sync;@Kendall2000_moran;@Abbott2007_moran], but in our previous
work [@Sheppard2016_aphids;@Sheppard2017_surrogates;@Walter2017_geogsync;@Sheppard_wavmodel],
and again here, we showed that using timescale-specific methods can help 
ameliorate this difficulty and can reveal new phenomena and previously unrecognized mechanisms.

# Acknowledgements
D. Lyden, J. Steinglein and the WIDNR for data; D. Storm, T. Van Deelen, J. Millspaugh, J. Fox, K. Haynes, and A. Liebhold for consultation; and annonymous referees for very useful feedback. Funding was provided by the James S McDonnell Foundation, United States National Science Foundation grants 1442595 and 1714195, United States Department of Agriculture award 2016-67012-24694, and The Nature
Conservancy. 

<!--Load sup mat workspace for the methods, which used to be in the sup mat-->
```{r echo=F, results="hide"}
load(file="WholeWorkspace.RData")
```

# References {-}

<div id="refs"></div>

\newpage


```{r Table1,echo=F,results='markup'}
options(knitr.kable.NA = '') #makes it put blanks instead of "NA" for NAs
rownames(Tab1_MT)<-NULL
knitr::kable(Tab1_MT[,c(1:7)],
             format="latex", align="l",escape=F,
             digits=4,
             col.names=c("Predictor","Response", "Timescale","P-value","Mean phase ($\\bar{\\theta}$)","Synchrony explained","Average cross terms"),
             caption = "Summary of spatial wavelet coherence tests and related results using county-level data. $P$-values and mean phases are from tests of spatial wavelet coherence (\\textit{Methods}) of the variables in columns 1 and 2 over the $3-7$-year timescale band. Mean phases over the timescale band (units of $\\pi$ radians) are provided only when meaningful, i.e., when spatial wavelet coherence was significant, $P < 0.05$, or marginally significant. A positive mean phase indicates the variable in the column \\textit{Response} is leading, and a negative mean phase indicates the variable in the column \\textit{Predictor} is leading. The \\textit{Synchrony explained} and \\textit{Average cross terms}  columns, which give percentages, are based on wavelet linear regressions (\\textit{Methods}) of the \\textit{Response} transforms against the \\textit{Predictor} transforms. When cross terms are low, synchrony explained is interpretable as the approximate percentage of synchrony in the response explainable as due to the synchrony in the predictor. Cross terms were low enough except for tests involving DVCs, for which USDA district-level results were more reliable (SI Table \\ref{SM-TabS2_SummaryDistrict}). \\textit{Synchrony Explained} \\textit{Average cross terms} are not reported for climate indices because of potential bias in that metric due to them not being spatially-explicit. All values were rounded to display four digits. Abundance = deer abundance. \\label{Tab1_CountyCoherenceSummary}",
             booktabs=T)
```
<!--TOM: Took this out of the caption above as it wasn't needed anymore: Cross-terms are zero for tests involving climate indices because for those tests the same values were used for all locations (\\textit{Methods})-->.              
\newpage

\begin{figure}[!h]  
\includegraphics[width=.5\textwidth]{Results/Fig1.png}
\caption{Simulations demonstrating our mechanism whereby
synchrony can cause population cycling. In scenario 1, 
environmental noise in location $i$ ($i=1,\dots,71$; first $15$ locations
shown in A) was modeled as 
$e^{(1)}_i(t)=\sin(2 \pi t/w^{(1)}+2 \pi u^{(1)}_i)+3\epsilon^{(1)}_i(t)$, where $w^{(1)}=3$, 
the $u^{(1)}_i$ were independent
draws from the uniform distribution on $(0,1)$, and the local noises $\epsilon^{(1)}_i(t)$ were
standard normal and independent across locations and times. In scenario 2,
noise ($15$ locations shown in B) 
was $e^{(2)}_i(t)=\sin(2 \pi t/w^{(2)}+2 \pi u^{(2)})+3\epsilon^{(2)}_i(t)$
where $w^{(2)}=3$, $u^{(2)}$ was one single standard-normally distributed random number used for all
locations, and the $\epsilon^{(2)}_i(t)$ were again
standard normal and independent.
Noise for scenario 3 (C) was like scenario 2 but used $w^{(3)}=10$. Thus the noises in 
scenarios 2 and 3 had weak and not visually obvious timescale-specific synchrony and those in 1 had no synchrony.
Populations were modelled as $p^{(s)}_{i}(t+1)=\rho p^{(s)}_i (t)+e^{(s)}_i(t)$ using $\rho=0.4$
for scenarios $s=1,2,3$, and are displayed (D-F) for locations 1-15.
Although synchrony is again weak and 
cannot readily be detected visually, wavelet mean field magnitide plots (\emph{Methods})
revealed that synchrony was present, in scenarios 2 and 3, on 3- and 10-year
timescales (G-I); on such plots, higher values mean greater synchrony at the given 
time on the given timescale. Total populations (J-L) show clear cycling in scenarios 2 and 3,
of periods $3$ and $10$, respectively, consistent with the proposed mechanism 
(\emph{Introduction}): synchrony on a given timescale 
produces cycling on that timescale in the total population.
We used $71$ locations because the empirical analyses of this study focus on white-tailed
deer populations in Wisconsin, and Wisconsin has 71 counties for which deer 
data were available. Time series on A-F were
shifted vertically for readability.
\label{Fig1_example}}
\end{figure}

\newpage

\begin{figure}[h!]
\includegraphics[width=\textwidth]{Results/Fig2.png}
\caption{Raw time series of (A) deer abundance, (B) deer-vehicle collisions (DVCs), 
(C) average winter ($\text{Dec}_{t-1}$ to $\text{Mar}_t$) snow depth, and 
(D) winter climate indices (Multivariate El Ni$\~{n}$o Southern Oscillation Index, 
MEI; Pacific Decadal Oscillation, PDO), and hunter abundances (E). 
Each line in A-C, E represents a county. Other variables were 
used in analyses (see text) but are not shown here due to lack of significant 
assocations with the deer and DVC variables that were the focus of the 
study.\label{Fig2_Timeseries}}
\end{figure}

\newpage

\begin{figure}[!h]
\includegraphics[height=.7\textheight]{Results/Fig3.png}
\caption{Wavelet mean field (A) and wavelet phasor mean field (B) magnitudes<!--, and predicted synchrony (C)--> 
of deer abundance time series (\textit{Methods}). Higher values in A, B indicate greater synchrony (A) or phase synchrony (B) 
at the indicated times and timescales. Contours on (B) indicate statistical significance of phase synchrony at 
$P =$ 0.001. Horizontal lines in (A) and (B) indicate the timescales that were the focus of subsequent statistical 
analysis, $3-7$-years. Colors on (C) show synchrony in deer abundance predicted from a wavelet linear model (\textit{Methods}) using 
snow depth, $\text{MEI}_{\text{w}}$, and $\text{PDO}_{\text{w}}$ as predictors, while contours in (C) show the 
actual wavelet mean field magnitude of deer abundance from (A), for comparison. Agreement is good, giving a 
visual indication of the statistical result (see \emph{Results}) that climatic Moran effects explain deer synchrony.
\label{Fig3_DeerSync}}
\end{figure}

\newpage

\begin{figure}[!h]
\includegraphics[height=.7\textheight]{Results/Fig4.png}
\caption{Wavelet mean field (A) and wavelet phasor mean field (B) magnitudes, and predicted synchrony (C) 
of DVCs. Colors, contours, and horizontal lines in A and B indicate the same as in 
Fig. \ref{Fig3_DeerSync}A,B, but for DVC data. Colors in (C) 
show synchrony in DVCs predicted by a wavelet model with sole predictor deer abundance, while contours in (C) 
show the actual wavelet mean field magnitude of DVCs from (A), for comparison. Agreement is good, giving a 
visual indication of the statistical result that deer synchrony explains DVC synchrony.
\label{Fig4_DvcSync}}
\end{figure}

\newpage

\begin{figure}[!h]
\includegraphics[width=.6\textwidth]{Results/Fig5.png}
\caption{Magnitude of the effects of synchrony on state-total deer abundance
and DVC fluctuations. Small panels show state-total deer abundance (A, in millions) and DVC (C, in thousands) time 
series (black lines) and surrogate totals obtained via a randomization procedure 
(\textit{Methods}) that eliminates synchrony among counties at $3-7$-year timescales 
(grey lines, 1000 surrogates plotted). Increased variation and semi-periodic 
oscillations due to synchrony are particularly notable when trends and long-timescale 
(longer than 7 years) variability are filtered out by subtracting the mean of the 
grey lines (B for deer, in thousands; D for DVCs). $\Delta$ represents the difference in values from 
the surrogate mean. Red arrows highlight examples of the strong cycling that is 
attributable to synchrony. For DVCs, the typical fluctuation in numbers of DVCs due 
to synchrony amounts to an estimated difference between peak and trough years of 
US\$$`r floor(round(2*avgdvccostsLow,-6))/1000000`-`r floor(round(2*avgdvccostsHigh,-6))/1000000`$Mn in damages 
statewide (\textit{Methods}). 
\label{Fig5_StateWideFluctuations}}
\end{figure}


