---
title: Synchronous effects produce major cycles in deer populations
  and deer-vehicle collisions
author:
 - Thomas L. Anderson\textsuperscript{1,2}, Lawrence W. Sheppard\textsuperscript{2}, Jonathan A. Walter\textsuperscript{2,3}, Robert E. Rolley\textsuperscript{4},  Daniel C. Reuman\textsuperscript{2,5}  
 - \textsuperscript{1}Department of Biology, Appalachian State University, 572 Rivers St., Boone, NC 28608, USA; \textsuperscript{2}Deparment of Ecology and Evolutionary Biology and Kansas Biological Survey, University of Kansas, 2101 Constant Ave, Lawrence, KS 66049, USA;  \textsuperscript{3}Department of Environmental Sciences, University of Virginia 291 McCormick Rd, Charlottesville, VA 22904, USA; \textsuperscript{4}Wisconsin Department of Natural Resources, 101 S. Webster St., Madison, WI 53707, USA; \textsuperscript{5}Laboratory of Populations, Rockefeller University, 1230 York Ave., New York, NY 10065, USA
date: 'correspondence to: anderstl@gmail.com, reuman@ku.edu'
output:
  pdf_document:
    fig_caption: yes
    keep_tex: yes
    includes:
      in_header: head_maintext.sty
  html_document:
    df_print: paged
  word_document: default
geometry: left=1cm,right=1cm,top=2.5cm,bottom=2.8cm
link-citations: yes
csl: pnas.csl
tables: yes
bibliography: AndersDeerShortRefs.bib
urlcolor: blue

---
\newpage
<!--Title length limit 96 characters, inc spaces-->

```{r echo=F, results="hide"}
#load in results
options(scipen=6)
TabS1_results<-readRDS(file = "Results/TableS1.rds")
TabS1_results<-cbind(TabS1_results[,1:3],apply(TabS1_results[,4:7],2,function(x){round(x,4)}))
abunwmeiSyncExp37<-round(TabS1_results$SynchronyExplained[TabS1_results$Predictor=="Winter MEI"&TabS1_results$Timescale=="3-7"],2)
abunwpdoSyncExp37<-TabS1_results$SynchronyExplained[TabS1_results$Predictor=="Winter PDO" &  TabS1_results$Timescale=="3-7"]
abunwmeiSyncExp47<-round(TabS1_results$SynchronyExplained[TabS1_results$Predictor=="Winter MEI"&TabS1_results$Timescale=="4-7"],2)
abunwpdoSyncExp47<-TabS1_results$SynchronyExplained[TabS1_results$Predictor=="Winter PDO" &  TabS1_results$Timescale=="4-7"]
snowabunSyncExp3_4<-TabS1_results$SynchronyExplained[TabS1_results$Predictor=="Snow Depth" & TabS1_results$Timescale=="3-4"]
abunmodelSyncExp<-readRDS(file="Results/abunmodelSyncExp.rds")

#load Table S2 results
TabS2_results<-readRDS(file="Results/TabS2_results.rds")
wmei_wpdo_pval<-round(TabS2_results$Pvalue[TabS2_results$Predictor=="Winter MEI" & TabS2_results$Response=="Winter PDO"],4)
wmei_wpdo_phase<-round(TabS2_results$MeanPhase[TabS2_results$Predictor=="Winter MEI" & TabS2_results$Response=="Winter PDO"],4)
abunwmeiSyncExp47<-TabS1_results$SynchronyExplained[TabS1_results$Predictor=="Winter MEI" &  TabS1_results$Timescale=="4-7"]

#load Table S4 results
TabS4_results<-readRDS(file = "Results/TableS4.rds")
usda_dvcabunSyncExp<-TabS4_results$SynchronyExplained[TabS4_results$Response=="DVCs" &
                                                        TabS4_results$Predictor=="Abundance"]
usda_adjdvcabunSyncExp<-TabS4_results$SynchronyExplained[TabS4_results$Response=="Traffic-adj. DVCs" &
                                                        TabS4_results$Predictor=="Abundance"]
usda.snowabunSyncExp3_4<-TabS4_results$SynchronyExplained[TabS4_results$Response=="Abundance" &  TabS4_results$Predictor=="Snow Depth" & TabS4_results$Timescale=="3-4"]
avgdeerfluc<-readRDS(file="Results/totaldeerfluctuations.rds")
avgdvcfluc<-readRDS(file="Results/totaldvcfluctuations.rds")
avgdvccostsLow<-readRDS(file="Results/totaldvccostsLow.rds")
avgdvccostsHigh<-readRDS(file="Results/totaldvccostsHigh.rds")
```

**Abstract:** Population cycles are a profound enigma of the natural world, the causes of which have been investigated for over 100 years. An early hypothesized explanation of cycles, for the iconic lynx-hare and for other systems, posited that cyclic environmental phenomena such as sun spots drive population cycles. But that hypothesis was refuted, and for several decades research focused instead on explanations involving nonlinear or delayed density dependence, often via species interactions. Here we show that timescale-specific environmental effects, by synchronizing local population dynamics on certain timescales only, cause major population cycles over large areas in white-tailed deer and deer-vehicle collisions (DVCs). We thereby demonstrate a new mechanism of cycles that centers once again on environmental influences. The mechanism differs fundamentally from earlier hypotheses that relied on a strongly cyclic driver such as sun spots. An important aspect of the new mechanism is specificity of synchronizing effects to certain timescales, which causes local dynamics to sum across space to a substantial cycle on those timescales. We also demonstrate for the first time that synchrony can be transmitted, in a cascading effect, from environmental drivers, to populations (deer), to human systems (DVCs). The cycles we explain can have substantial effects on agriculture, hunting, and accident recovery industries. Research shows that synchrony of some environmental drivers may be changing, possibly as part of climate change. Such changes may alter population cycles that arise via the mechanism described here.

<!--***DAN: Old first para
**Correlated fluctuations through time in abundances of distant populations, called spatial synchrony, is 
a widespread and important phenomenon in population ecology [@Leibhold2004_sync]. Synchrony of exploited 
or economically important taxa should, theoretically, have substantial impacts for humans, because
synchrony can produce high-amplitude oscillations through time in total populations across large 
regions. However, synchrony has seldom been investigated empirically for exploited species, 
and its anthropogenic importance is not often considered. Additionally, it has historically been 
difficult to distinguish drivers of synchrony in natural populations of any 
taxa[@Kendall2000_moran;@Abbott2007_moran], though improved analytical methods that decompose 
synchrony by timescale have furthered these 
efforts[@Allstad2015_gypsysync;@Sheppard2016_aphids;@Sheppard2017_surrogates;@Walter2017_geogsync]. 
Here we identify the causes of spatial synchrony in the dynamics of populations of white-tailed deer, 
a widely exploited species, and its consequences on deer-vehicle collisions (DVCs) across Wisconsin, 
USA. Deer synchrony occurred principally at $3-7$-year timescales of fluctuation, and synchronous 
winter environmental variation explained almost 90% of deer synchrony through a process called the 
Moran effect[@Moran1953;@Ranta1997_moran]. DVCs were in turn synchronized by deer populations. 
Timescale-specific synchrony caused large semi-periodic fluctuations in statewide total deer 
populations and DVCs that would not have occurred without synchrony. Thus synchrony caused local 
dynamics to sum to a major cyclic phenomenon on statewide spatial scales. Our results are 
principally important as a demonstration of a new and potentially widespread mechanism for 
population cycles on large spatial scales, a phenomenon of long-standing interest.**
-->

<!--Authors must submit a 120-word-maximum statement about the significance of their research paper written at a level understandable to an undergraduate-educated scientist outside their field of specialty. The primary goal of the Significance Statement is to explain the relevance of the work in broad context to a broad readership. Significance statements are not required for Brief Reports. TOM: My first attempt below is currently 140 words-->
**Significance Statement:** Understanding the mechanisms behind cycles in population abundances has long been a goal in ecology. Here we demonstrate, in white-tailed deer, a new mechanism by which population cycles can arise, and the consequences of cycles for humans. Fluctuations in deer abundances in different counties of Wisconsin, USA, were highly correlated due to correlated fluctuations in winter climate. Importantly, these synchronized dynamics occurred only on 3-7-year timescales. Deer synchrony in turn produced synchronized oscillations in deer-vehicle collisions across Wisconsin, again with 3-7-year periodicity. When aggregated spatially, fluctuations for individual counties reinforced each other on 3-7-year timescales, producing state-wide cycles. In addition to revealing new mechanisms, our results are important for hunting, insurance industries, and ecosystems influenced by deer.   

#Introductory text

The causes of population cycles have interested and mystified ecologists for over 100 
years [@Myers2018_cycles]. Although early hypotheses, for the 
iconic lynx-hare as well as for other 
systems, proposed that cyclic environmental phenomena such as sun spots drive population 
cycles [@Sinclair1992_sunspots;@Selas2004_sunspots], these hypotheses were robustly
refuted [@Nilssen2007_nosunspots;@Myers2013_nosunspots], and research has since shifted to 
explaining cycles via non-environmental mechanisms based on nonlinear or delayed density 
dependence, often operating through species 
interactions [@Krebs2018_harereview;@Krebs2013_book;@Myers2013_nosunspots;@MartinezPadilla2014_grouse]. 
However, spatial population synchrony, i.e., the tendency for populations of a species in 
different locations to exhibit correlated fluctuations through time, is a 
ubiquitous phenomenon [@Leibhold2004_sync]; and, if synchrony of local populations
occurs predominantly on specific timescales, then, in theory, it can produce cycling 
of total populations across large spatial scales (Fig. \ref{Fig1_example}A-F).
Essentially, the synchronous frequency 
components of local time series reinforce each 
other in the total time series, whereas asynchronous components on other 
timescales cancel out when summed. Recent work shows that timescale structure 
in synchrony is common - synchrony can occur on some timescales of population
dynamics, but be absent on others [@Sheppard2016_aphids;@Sheppard2017_surrogates;@Walter2017_geogsync;@Sheppard_wavmodel;@Anderson2019_LO]. Synchrony is commonly caused by synchronized environmental 
variation, a process called the Moran effect [@Moran1953]. 
Thus timescale-specific synchrony 
is an unexplored but potentially important mechanism of population cycles
that may help renew interest in environmental contributors to population cycles.
Timescale-specific synchrony could 
also be important for industries such as 
hunting and agriculture that exploit multiple populations across space if it 
causes cycling in the sum of the exploited populations, and synchrony 
of pests may likewise produce cycles in regional total damage 
caused by the pests [@Allstad2015_gypsysync]. 

We investigate timescale-specific synchrony and related cycling of white-tailed deer (*Odocoileus virginianus*) and DVCs in Wisconsin (WI), USA. White-tailed deer is a species of considerable interest due to its major economic and cultural benefits and impacts on ecosystems [@Cote2004_deerimpacts;@Conover2011_bookchp]. Across the USA in 2011, US\$$34$Bn were spent on hunting big game, including deer [@usfws2011_survey]. 
In WI, hunting-related economic activity accounts for ~$1\%$ of state GDP [@usfws2011_survey]. DVCs cause about 50 human fatalities and US\$$1.5$Bn in vehicular damage annually across the USA [@Conover2011_bookchp]. Thus it is an important applied question what drives the year-to-year variability in statewide deer abundance and DVC totals. We consider potential drivers of deer populations including both environmental factors and the direct effects of hunting. We use data from 1981-2016 (SI Appendix Fig.  \ref{SM-FigS1_Timeseries}) aggregated at three spatial resolutions (*Methods*): counties ($n=71$), United States Department of Agricultural (USDA) districts ($n=9$), and statewide (SI Appendix Fig. \ref{SM-FigS2_map}). 

#Results and Discussion

The wavelet mean field [@Sheppard2016_aphids], a technique for visualizing the strength of synchrony as a
function of time and timescale (*Methods*, Fig \ref{Fig1_example}G-I and SI Appendix Fig. \ref{SM-PedagogFigWMF} provide examples), revealed that 
the most prominent synchrony in county-level deer abundances occurred at $3-7$-year timescales 
(Fig. \ref{Fig2_DeerSync}A). A related technique, the wavelet phasor mean field (*Methods*, SI Appendix Fig. \ref{SM-PedagogFigWMF}), 
demonstrated the strong statistical significance of this phenomenon (Fig. \ref{Fig2_DeerSync}B). We
therefore focused our analysis on the $3-7$-year timescale band. Most past analyses of synchrony have
used correlation methods that conflate timescales, but recent results show this can hinder detection of
synchronous phenomena by obscuring important synchrony on some timescales with asynchronous or noisy dynamics on other timescales [@Sheppard2016_aphids]. Moreover, synchrony must be specific to certain timescales to produce cyclic dynamics of
total populations on those timescales (Fig. \ref{Fig1_example}).  

Spatial wavelet coherence ([@Sheppard2016_aphids;@Sheppard2017_surrogates], SI Appendix Fig. \ref{SM-PedagogFigSpatCoh} provides an example), wavelet regression techniques [@Sheppard_wavmodel], and a wavelet Moran theorem ([@Sheppard2016_aphids;@Sheppard_wavmodel, *Methods*) demonstrated that <!--***DAN: Is there a way to get the parens to not appear around the above citations? No big deal if not, the typesetters can take care of it, but it might create a better impression in the reviewers if we fix it now--><!--***TOM: I've tried using both StackOverflow and Twitter to find an answer and no response so we may just have to go with it. Given we might have to do some post hoc editing to get it in Science format anyways, we could just manually change it.--><!--***DAN: OK, that's fine, leaving these comments here to remind us the issue still exists, though we may never do anything about it.-->
deer abundance synchrony on $3-7$-year timescales was principally caused by climatic Moran effects. 
Although it is known that a spatially synchronous environmental driver can induce 
population synchrony via Moran effects [@Leibhold2004_sync; @Moran1953; 
@Ranta1997_moran], historically it has been 
difficult to attribute synchrony 
in natural populations to specific environmental drivers versus other potential 
causes of synchrony such as dispersal [@Kendall2000_moran;@Abbott2007_moran]. 
The analytical advances listed above can facilitate more precise 
identification of Moran drivers 
[@Sheppard2016_aphids;@Sheppard2017_surrogates;@Walter2017_geogsync;@Sheppard_wavmodel]. 
Their effectiveness is due in part to their capacity to decompose 
synchrony according to the distinct timescales on which it can occur (Fig. 
\ref{Fig1_example}G-I).
Synchrony in deer abundance was related to winter snow depth 
(spatial wavelet coherence, $P =$ `r round(TabS1_results$Pvalue[TabS1_results$Predictor=="Snow Depth" &TabS1_results$Timescale=="3-7"],2)`, SI Appendix
Table \ref{SM-TabS1_CountyCoherenceSummary}), in a near anti-phase relationship (mean phase 
difference between abundance and snow depth across $3-7$-year timescales, in units of ${\pi}$ 
radians, was $\bar{\theta} =$ `r round(TabS1_results$MeanPhase[TabS1_results$Predictor=="Snow Depth"&TabS1_results$Timescale=="3-7"],2)`; SI Appendix Fig. \ref{SM-FigS4_PhaseDiffs}D): deer 
abundances tended to peak in years of below-average snow. This effect probably occurred because 
deeper snow limits access to winter food resources in ungulates, which subsequently affects body 
condition, leading to greater susceptibility to starvation or predators and reduced parturition 
rates [@Mysterud2008_deerclimate;@Hegel2012_caribou]. Deer abundance was also related, with
approximately quarter-cycle phase shifts, 
to the multivariate El Ni$\~{n}$o Southern Oscillation index
($P =$ `r round(TabS1_results$Pvalue[TabS1_results$Predictor=="Winter MEI"&TabS1_results$Timescale=="3-7"],2)`, $\bar{\theta}  =$ `r round(TabS1_results$MeanPhase[TabS1_results$Predictor=="Winter MEI"&TabS1_results$Timescale=="3-7"],2)`) and to the Pacific Decadal Oscillation 
($P =$ `r round(TabS1_results$Pvalue[TabS1_results$Predictor=="Winter PDO"&TabS1_results$Timescale=="3-7"],2)`, $\bar{\theta}  =$ `r round(TabS1_results$MeanPhase[TabS1_results$Predictor=="Winter PDO"&TabS1_results$Timescale=="3-7"],2)`, SI Appendix 
Table \ref{SM-TabS1_CountyCoherenceSummary}, Fig. \ref{SM-FigS4_PhaseDiffs}A,C), both averaged over the winter months
($\text{Dec}_{t-1}$ to $\text{Mar}_t$, abbreviations $\text{MEI}_{\text{w}}$ and
$\text{PDO}_{\text{w}}$). High values of these indices generally correspond to mild 
winter weather over
WI [@Rodionov2003_elnino]. The two indices were strongly coherent and in-phase with
each other on $3-7$-year timescales over the period 1981-2016 (wavelet coherence, $P =$ `r round(TabS2_results$Pvalue[TabS2_results$Predictor=="Winter MEI" & TabS2_results$Response=="Winter PDO" ],4)`, $\bar{\theta}  =$ `r round(TabS2_results$MeanPhase[TabS2_results$Predictor=="Winter MEI" & TabS2_results$Response=="Winter PDO" ],2)`, SI Appendix Table \ref{SM-TabS2_PairwiseClimVarSpatCoh}). 
At $5$-yr periodicity, the quarter-cycle phase lags correspond to deer populations
peaking about `r round(-5*TabS1_results$MeanPhase[TabS1_results$Predictor=="Winter PDO"&TabS1_results$Timescale=="3-7"]/2,1)` or 
`r round(-5*TabS1_results$MeanPhase[TabS1_results$Predictor=="Winter MEI"&TabS1_results$Timescale=="3-7"]/2,1)` years after mild winters, similar to other reported lag times of climatic effects on deer 
populations [@Post1998_popgrowth]. Winter weather and climatic oscillations have been identified previously as mechanisms influencing synchrony in other ungulates across multiple 
regions [@Post2004_greenland;@Grotan2005_roedeer;@Hegel2012_caribou]. Applying the wavelet Moran theorem ([@Sheppard2016_aphids;@Sheppard_wavmodel], *Methods*), 
<!--***DAN: I would also like the parens in the citation above to not appear here-->
snow depth, 
$\text{MEI}_{\text{w}}$, and $\text{PDO}_{\text{w}}$ together explained 
$`r round(abunmodelSyncExp,0)`\%$ of deer synchrony on $3-7$-year timescales. 
See Fig. \ref{Fig2_DeerSync}C for a visual manifestation of this statistical result. 

Snow depth and climate index results seemed to indicate distinct or partly distinct phenomena 
synchronizing deer on different timescales within the $3-7$-year range. Again applying the wavelet 
Moran theorem, 
snow depth explained $`r round(usda.snowabunSyncExp3_4,0)`\%$ 
of synchrony in deer abundance
at $3-4$-year timescales when data were aggregated to USDA districts (*Methods*, SI Appendix text and Table \ref{SM-TabS4_SummaryDistrict}). Dynamics at $3-4$-year timescales were not well explained by either $\text{MEI}_{\text{w}}$ or $\text{PDO}_{\text{w}}$, which instead contributed more
to deer synchrony over $4-7$-year timescales (SI Appendix Tables \ref{SM-TabS1_CountyCoherenceSummary} and \ref{SM-TabS4_SummaryDistrict}, Fig. \ref{SM-FigS6_RanksInSurrogs}A). 
$\text{MEI}_{\text{w}}$ and $\text{PDO}_{\text{w}}$ explained $`r round(abunwmeiSyncExp47,)`\%$ and $`r round(abunwpdoSyncExp47,0)`\%$ of deer synchrony on $4-7$-year timescales, respectively (SI Appendix Table \ref{SM-TabS1_CountyCoherenceSummary}). Snow depth did not explain $4-7$-year synchrony: spatial wavelet coherence between snow depth and deer abundance on those timescales was non-significant (SI Appendix Table \ref{SM-TabS1_CountyCoherenceSummary}, \ref{SM-TabS4_SummaryDistrict}).

Deer abundance synchrony was not caused by hunting on $3-7$-year timescales. County-level hunter and deer
abundances were not significantly spatially coherent 
($P =$ `r round(TabS1_results$Pvalue[TabS1_results$Predictor=="Hunters" & TabS1_results$Timescale=="3-7"],2)`) over $3-7$-year timescales 
(SI Appendix Table \ref{SM-TabS1_CountyCoherenceSummary}, Figs \ref{SM-FigS6_RanksInSurrogs}B, \ref{SM-FigS7_WMFHunter}). 
However, hunter abundance was significantly related to deer abundance on shorter timescales, around 
$2-2.5$-years ($P =$ `r round(TabS1_results$Pvalue[TabS1_results$Predictor=="Hunters" & TabS1_results$Timescale=="2-2.5"],3)`), and were in approximate anti-phase with deer abundance 
on those timescales ($\bar{\theta}  =$ `r round(TabS1_results$MeanPhase[TabS1_results$Predictor=="Hunters" & TabS1_results$Timescale=="2-2.5"],2)`; 
SI Appendix Table \ref{SM-TabS1_CountyCoherenceSummary}, Figs \ref{SM-FigS4_PhaseDiffs}G,
\ref{SM-FigS6_RanksInSurrogs}B). These patterns are consistent with the hypothesis that hunters can
regulate deer populations over short timescales but do not do so on longer timescales. The lack of
longer-timescale spatial wavelet coherence between hunters and deer abundance is likely also related to the facts that
hunters prefer to shoot bucks, the removal of which has limited population effects; and the state
regulates antlerless harvest in an attempt to maintain populations at a desirable size.  

Having attributed the bulk of the synchrony in deer abundance at $3-7$-year timescales to climatic Moran 
effects, we also found strong coherence between deer abundance and DVCs on $3-7$-year timescales: Moran 
effects pass from climate to deer to DVCs in a cascading manner, and in this way are ultimately
responsible for synchrony of DVCs. The wavelet mean field and phasor mean field of DVCs revealed time-
and timescale-specific synchrony, and visual patterns were remarkably similar to those shown by deer 
(compare Figs \ref{Fig2_DeerSync}A,B and \ref{Fig3_DvcSync}A,B). Spatial wavelet 
coherence ([@Sheppard2016_aphids;@Sheppard2017_surrogates, *Methods*) 
<!--***DAN: Would be nice if we could fix the parens in the above citations-->
showed that deer abundance and DVCs were significantly related 
($P =$ `r round(TabS1_results$Pvalue[TabS1_results$Response=="DVCs" & TabS1_results$Predictor=="Abundance"],3)`, 
SI Appendix Table \ref{SM-TabS1_CountyCoherenceSummary}) on $3-7$-year timescales, and had a nearly in-phase 
relationship ($\bar{\theta} =$ `r round(TabS1_results$MeanPhase[TabS1_results$Response=="DVCs" & TabS1_results$Predictor=="Abundance"],2)`, SI Appendix Fig. \ref{SM-FigS4_PhaseDiffs}E). 
Greater deer abundance almost certainly caused greater DVCs rather 
than DVCs driving deer fluctuations because DVCs were too infrequent to meaningfully impact deer 
abundance (SI Appendix Fig. \ref{SM-FigS1_Timeseries}), and if they did an anti-phase relationship would have 
been expected. Deer abundance explained $`r round(usda_dvcabunSyncExp,0)`\%$ of the synchrony in 
DVCs when data were aggregated to USDA districts (*Methods*, SI Appendix text and Table \ref{SM-TabS4_SummaryDistrict}). 
As a visual manifestation of this statistical result, patterns of synchrony in DVCs predicted by a
wavelet regression model with sole predictor deer abundance match the patterns actually present in DVC
synchrony (Fig. \ref{Fig3_DvcSync}C). 
Statistical significance, phase relationships, and the amount of DVC synchrony explained by deer
abundance remained similar after adjusting DVCs to account for
fluctuations in traffic volume (SI Appendix text and Tables \ref{SM-TabS1_CountyCoherenceSummary}, \ref{SM-TabS4_SummaryDistrict}, Fig. \ref{SM-FigS4_PhaseDiffs}F).

As mentioned above, timescale-specific synchrony may lead, theoretically, to cyclic 
dynamics in spatial-total or spatial-average time series, which can mean 
alternating relative abundance and scarcity of important species over large areas 
[@Allstad2015_gypsysync]. Our data show the empirical reality of this effect and its 
strength for deer and DVCs. Statewide oscillations due to synchrony correspond to a typical 
difference of $~ `r round(avgdeerfluc,-4)`$ deer and $~ `r round(avgdvcfluc,-2)`$ 
DVCs between peak and trough years 
(Fig. \ref{Fig4_StateWideFluctuations}). Importantly, these high amplitude 
statewide oscillations in deer and DVCs disappear (Fig. 
\ref{Fig4_StateWideFluctuations}) when synchrony between counties on $3-7$-year 
timescales is removed via a randomization procedure 
(*Methods*); spatial synchrony causes local dynamics to sum to a substantial 
semi-periodic oscillatory phenomenon on statewide spatial scales. Depending on 
assumptions made about costs and reporting rates of DVCs 
[@Huijser2008_dvc;@Marcoux2010_dvc;@Conover2011_bookchp], statewide DVC damages 
vary by US\$$`r floor(round(2*avgdvccostsLow,-6))/1000000`-`r floor(round(2*avgdvccostsHigh,-6))/1000000`$Mn 
between peak and trough years (*Methods*). Similarly, fluctuations of a quarter 
million deer likely have considerable consequences for agricultural crop damage 
and hunting success, both of which can have socioeconomic impacts. See 
SI Appendix for further details on the cycling mechanism.

Overall, we have demonstrated that timescale-specific Moran effects, by 
synchronizing local dynamics, can cause major cycles in the 
regional total population. This new mechanism
differs fundamentally from prior, now refuted hypotheses that population cycles in 
snowshoe hare [@Sinclair1992_sunspots] and moth pests [@Selas2004_sunspots] are due 
to highly regular environmental fluctuations such as 
sun spots because the new mechanism does not require a highly regular environmental
oscillator. Rather it is sufficiently simply for irregular environmental drivers 
such as winter climatic factors to be coherent with populations in a timescale 
band, thereby inducing synchrony
in that band and leading to total-population cycles on the same timescales. 
Timescale specificity in synchrony is 
common [@Sheppard2016_aphids;@Sheppard2017_surrogates;@Walter2017_geogsync;@Sheppard_wavmodel]. 
Resulting cycles affect the exploiting 
industry (hunting) and related industries 
(emergency medicine and accident recovery). We have also 
presented the first demonstration that synchrony can be transmitted in a 
Moran-like effect from ecological to human systems (deer to DVCs). An 
important aspect of all of the phenomena we described is their specificity 
to certain timescales. Recent work on several taxa 
[@Allstad2015_gypsysync;@Sheppard2016_aphids;@Koenig2016_birdsync;@Shestakova2016_forestsync] shows 
that synchrony of some environmental drivers may be changing, 
possibly due to climate change. These changes may alter population synchrony, 
and therefore also population cycles that arise by our new mechanism. 
In particular, the predicted increase 
in the frequency of strong El Ni$\~{n}$o events 
[@Cai2014_elnino] may lead to altered cycling period in deer 
and DVCs, though our mechanisms are general so we would not 
expect effects to be limited to the system we have studied.

<!--Load sup mat workspace for the methods, which used to be in the sup mat-->
```{r echo=F, results="hide"}
load(file="WholeWorkspace.RData")
```

# METHODS

\textbf{Deer and hunter abundance data.} We obtained deer abundance estimates from the Wisconsin Department of 
Natural Resources (WIDNR) spanning the `r dim(cty.list$Abun)[2]`-yr period `r minyear`-`r maxyear` (Fig. \ref{SM-FigS1_Timeseries}). 
From 1981-2013, deer abundance was estimated for deer management units 
(DMUs, $N=$ 101-139 depending on the year). For most years and DMUs, 
abundance was estimated using a sex-age-kill (SAK) model, with aerial surveys or 
accounting models used for a smaller number of years and DMUs [@widnr2001]. 
In $2014$, DMU boundaries were changed to match counties ($N=72$). Because of this, and because other 
spatial variables (e.g., DVCs, climate) were gathered at the county level, 
we transformed the 1981-2013 abundance estimates for DMUs 
into county-level estimates. To do so, we first calculated the areas 
of intersection between DMU and county boundaries for each year (as 
DMU boundaries changed frequently), using the 'intersect' function in 
Arc GIS 10.3 (ESRI, Redlands, CA). We then calculated deer abundance 
for each intersection area by multiplying the amount of deer habitat in 
each intersection area by the WIDNR density estimate of deer for the DMU 
that the intersection fell within. Deer habitat was considered to comprise 
areas of permanent cover (e.g., forests) that were $>4$ ha in size, plus agriculture and grass fields within 100 m of permanent cover. DMU deer 
densities were calculated by dividing the DMU abundances originally provided to us
by the amount of deer habitat within the DMU [@widnr2001]. This procedure 
assumes approximately spatially uniform densities of deer within deer 
habitat across each DMU. The abundance for a county was the sum of 
the estimated abundances of the county's intersections with DMUs. 
We dropped one county (Menominee) due to incomplete data for most 
variables, leaving `r length(unique(dat$COUNTY_NAM))` counties for statistical analysis.

We obtained from the WIDNR estimated hunter abundance data for each DMU for the years 1992-`r maxyear`
(Fig. \ref{SM-FigS1_Timeseries}); these data constituted an index of the estimated 
number of hunters during the first two days of the 9-day gun 
hunting season [@widnr2001], assuming this was a good proxy 
for hunter numbers across the season. Gun hunters on average represent $`r gunhunter_percent`\%$ of the total hunters in WI each year over the years of this study. We estimated hunter abundances for counties using the same method as for deer abundance. We 
dropped 18 counties due to incomplete survey 
efforts of hunters or missing data, leaving 53
counties for analysis that used the hunter density data. 

\textbf{DVC data.} We obtained estimates of DVCs for each county, spanning the 
years 1987-`r maxyear`, from the WI Department 
of Transportation (WIDOT; Fig. \ref{SM-FigS1_Timeseries}). These data reflect 
the number of police-investigated DVCs; they are an 
index of DVCs, as opposed to an absolute number of incidents, 
since not all DVCs are reported. In Michigan, nearly half of 
DVCs went unreported to the police or insurance companies 
over a five-year period [@Marcoux2010_dvc]. Additionally, 
DVC information was collected with reporting effort that 
may have varied for different counties 
(Daniel Storm, WIDNR, Personal Communication), and the 
threshold of damage for the police to respond to and report on 
an incident increased statewide in 1996 from US\$500 to US\$1000. 
However, features of the data such as these could only have obscured 
and could not have artefactually generated the strongly 
significant spatial wavelet coherence relationships we report between 
DVCs and deer abundance. Shortcomings of the DVC data added 
noise to the DVC data and likely obscured the relationship 
with deer abundance, so that if better DVC data were available 
this relationship may have been even stronger than reported. 
We also obtained traffic volume estimates (millions of miles 
driven) for each county spanning the years 1988-`r maxyear` from the 
WIDOT, to control for traffic in our analysis of DVCs.

\textbf{Weather and climate data.} We obtained daily weather data (snow depth, 
precipitation, and minimum and maximum air temperature - here called Tmin and Tmax) from 
all available weather stations in each county from the 
National Climatic Data Center ([`https://www.ncdc.noaa.gov/`](https://www.ncdc.noaa.gov/)). We 
averaged daily values across the winter ($\text{Dec}_{t-1}$ 
to $\text{Mar}_t$) and across stations within a county to produce 
annual time series of winter weather by county which were used in 
all subsequent analyses (Fig. \ref{SM-FigS1_Timeseries}). The number of 
weather stations per county and data collected at each station 
varied, resulting in different spatial extents for each variable 
in order to retain the full temporal extent: Tmin, $N=$ `r dim(winter.clim$Tmin[!is.na(rowMeans(winter.clim$Tmin)),])[1]`
counties; Tmax, $N=$ `r dim(winter.clim$Tmax[!is.na(rowMeans(winter.clim$Tmax)),])[1]`; precipitation, $N=$ `r dim(winter.clim$Prcp[!is.na(rowMeans(winter.clim$Prcp)),])[1]`; snow depth, $N=$ `r dim(winter.clim$Snwd[!is.na(rowMeans(winter.clim$Snwd)),])[1]`. 
We also calculated a Winter Severity Index (WSI, $N=$ `r dim(winter.clim$WSI[!is.na(rowMeans(winter.clim$WSI)),])[1]`), 
a metric known to impact deer populations [@widnr2001]. WSI is the sum of the total number of days for which snow depth was greater 
than 18 inches or Tmin was less than $0^\circ$F. 

We obtained monthly estimates of three global climate indices: 
the North Atlantic Oscillation (NAO), the multivariate 
El Ni$\~{n}$o Southern Oscillation Index (MEI), and the Pacific 
Decadal Oscillation (PDO). All of these have been shown to 
influence large ungulate population dynamics in the north central 
USA [@Post1998_popgrowth;@Post2004_greenland]. 
We averaged monthly values across the 
approximate growing season (Apr-Nov), and the winter 
($\text{Dec}_{t-1}$ to $\text{Mar}_t$) for each index 
(Fig. \ref{SM-FigS1_Timeseries}). We used a total of 11 weather and 
climate variables in subsequent analyses: winter Tmin, Tmax, 
snow depth, and precipitation; winter NAO, MEI, and PDO; 
growing-season NAO, MEI and PDO; and the WSI. 

\textbf{Data pre-processing.} We performed all analyses described below and in the main text 
at three spatial scales: counties (N = 44-`r dim(cty.list$Abun)[1]` depending on 
the variables involved; mean county 
area $\pm$ SD = `r round(mean(gps$Area_km2),0)` $\pm$ `r round(sd(gps$Area_km2),0)` $\text{km}^2$), 
United States Department of Agriculture (USDA) districts ($N = 9$; mean 
district area $\pm$ SD = `r round(mean(usda.centroid$Area_km2),0)` $\pm$ `r round(sd(usda.centroid$Area_km2),0)` $\text{km}^2$), 
and statewide (Fig. \ref{SM-FigS2_map}). Analyses at three spatial scales 
were performed because the SAK model used by the WIDNR to estimate deer 
abundance for each management unit has been criticized as less accurate 
at smaller spatial scales [@Millspaugh2009]. The bias in abundance 
estimates previously reported [@Millspaugh2009] may have been overestimated, as more recent analyses 
suggest the uncertainty in SAK estimates is substantially less 
($\pm 20\%$ error rates). 
In addition, the criticism of the SAK [@Millspaugh2009] was based 
on a model that did not match perfectly with practices of the WIDNR. 
Nevertheless, 
to ensure our results were not biased, we analyzed 
synchrony on multiple spatial scales including two larger scales 
(districts and statewide) for which SAK data have not been criticized. 
USDA district data were 
produced as the sum or 
average of county-level variables, as districts were always conglomerates 
of counties (Fig. \ref{SM-FigS2_map}). 
Statewide analysis used state total or average time series for each 
variable. In general, results remained unchanged in their main substance using data 
aggregated to larger spatial scales (Table \ref{SM-TabS4_SummaryDistrict}, 
Fig. \ref{SM-FigS8_WMFAndPredictionsDistricts}), suggesting results for counties were not 
meaningfully influenced by possible inaccuracies in SAK abundance estimation 
procedures at the county level.

We performed an optimal Box-Cox transformation on each time series to help ensure Gaussian marginal distributions [@Sheppard2016_aphids], which are necessary for the Fourier-surrogate-based spatial wavelet coherence and wavelet linear regression procedure described below. This process also included linear detrending (and subtracting the mean) to remove any long-term trends that might otherwise obscure patterns of synchrony [@Buonaccorsi2001]. Time series
variances were also standardized to 1. These steps were done via the `cleandat` function in the `wsyn` package for the R programming language, available at `www.github.com/reumandc/wsyn/CRANv1p0p0` and on the Comprehensive R Archive Network (CRAN).

\textbf{The wavelet mean field and wavelet phasor mean field.} We assessed the strength of synchrony 
and its dependence on time and timescale for 
all spatiotemporal variables using two 
statistical tools: the wavelet mean field and 
the wavelet phasor mean field [@Sheppard2013;@Sheppard2016_aphids]. 
Both of these are standard tools in 
some biological fields and well established; they are described in 
mathematical detail elsewhere [@Sheppard2013;@Sheppard2016_aphids] and 
operationally here. Both techniques take as 
input time series measured at the same times 
in several locations, and return as output 
complex-number-valued functions of time 
and timescale. When the magnitudes of these complex functions are plotted against time 
and timescale, the plots characterize the degree 
of synchrony among the input time series and how 
it may have changed over time or differed by 
timescale (Figs \ref{Fig1_example}, \ref{SM-PedagogFigWMF}). 
The wavelet phasor mean field magnitude 
quantifies the degree to which phases of 
oscillation in the input time series were the 
same across locations, as a function of time 
and timescale. The wavelet mean field 
magnitude incorporates information not only 
on phase synchronization but also on the 
tendency of oscillatory components on any 
given timescale to have magnitudes which are 
correlated through time across the sampling 
locations. Both quantities are based on complex Morlet 
wavelet transforms of each input time series. 
Statistical significance of the wavelet phasor mean 
field, compared to a null hypothesis of no phase 
synchrony, can be assessed for each combination of 
time and timescale. For any given time and 
timescale, significance testing compares the degree 
of phase consistency between sampling locations for 
real time series to a null hypothesis of random 
phases. If real phase consistency was greater than 
that of $99.9\%$ of sets of random phases 
($P < 0.001$, one-tailed test, see contours on, e.g., Figs \ref{Fig2_DeerSync}B and \ref{Fig3_DvcSync}B), synchrony was 
deemed significant and important 
[@Sheppard_wavmodel]. Wavelet mean field and 
wavelet phasor mean field magnitude plots for 
winter weather variables are in 
Fig. \ref{SM-FigS9_WMFsAndWPMFsAncillary}. Wavelet mean fields were computed using the
function `wmf` and wavelet phasor mean fields were computed using the function `wpmf` in the `wsyn` package for the R programming language.

All plots based on wavelet transforms are "scalloped", i.e., values corresponding to longer timescales and times closer to the start or end of available data are not plotted. This is because the wavelet transform is based on convolutions of wavelet functions with the timeseries being transformed. Convolutions for which a substantial portion of the wavelet function overhangs the edge of the data are unreliable. "Scalloping" wavelet plots is standard practice.

\textbf{Spatial wavelet coherence and wavelet coherence.} For both county- and district-level data, we 
assessed pairwise relationships between 
spatiotemporal variables using the spatial wavelet coherence 
technique. This technique is also a well-established and standard tool in some biological fields; it is described in detail 
elsewhere [@Sheppard2013;@Sheppard2016_aphids;@Sheppard2017_surrogates] and  operationally
here. Spatial wavelet coherence tests whether two 
spatiotemporal variables exhibit consistent phase 
differences and correlated magnitudes of 
oscillation through time and across space, as a 
function of timescale. Significance is determined 
by comparing spatial wavelet coherence of data with 
spatial wavelet coherences of appropriately randomized, 
surrogate data sets ($N=$ `r nsurrogs` surrogates 
were used for all tests) 
that retain the same 
temporal and spatial autocorrelation properties as 
the original time series but are otherwise 
unrelated to each other [@Theiler1992;@Schreiber2000]. Standard Fourier 
surrogates were used [@Schreiber2000;@Sheppard2013;@Sheppard2016_aphids]. 
We calculated how often observed spatial wavelet coherence 
between pairs of variables was ranked higher than 
coherences of surrogate data sets, aggregating 
across the timescale band of interest using an 
average-rank-based method described in detail 
elsewhere [@Sheppard2016_aphids]. A $p$-value of $0.05$ 
means that the empirical spatial wavelet coherence was higher, in an average 
sense across the timescale band of interest, than the same 
quantity calculated for `r 0.95*nsurrogs` of `r nsurrogs` 
surrogate data sets representing the null hypothesis of no relationship between variables. Ranks of actual spatial wavelet
coherences in surrogate spatial wavelet coherences were 
also plotted as a function of timescale 
(Fig. \ref{SM-FigS4_PhaseDiffs}). Spatial wavelet coherences were computed and results were processed using the `coh`, `bandtest`, `plotrank`, `plotphase` and related functions in the `wsyn` package for the R programming language. The fast algorithm of Sheppard *et al.* [@Sheppard2017_surrogates] for assessing the statistical significance of spatial wavelet coherence was used.

Correlation-based approaches have been standard practice in synchrony studies [@Buonaccorsi2001], but these methods assume in-phase relationships. In contrast, spatial wavelet coherence 
allows for any kind of phase relationship and 
provides as part of its output the average phase that occurred between two variables. 
Therefore, when two variables had significant 
spatial wavelet coherence over a timescale band, we 
calculated the mean phase difference between the 
variables over the band, $\bar{\theta}$, in units of $\pi$ 
radians, to determine if oscillations had 
approximately in-phase ($|\bar{\theta}|=0$ to 
$0.25$), quarter-cycle ($|\bar{\theta}|=0.25$ to 
$0.75$), or anti-phase ($|\bar{\theta}|=0.75$ to 
$1$) relationships. We used "scalloped" 
wavelet transforms for 
calculating phase differences. Phase difference was not meaningful when spatial
wavelet coherence was not significant and so was not reported.

As with correlations, causal relationships between variables are not directly 
revealed by the spatial wavelet coherence. But biological arguments, combined 
with significant coherence and phase information, 
can strongly support one causal hypothesis over 
others. Climate and weather variables could not reasonably be influenced 
by deer abundance. Therefore, a highly significant 
spatial wavelet coherence between a weather variable and 
deer abundance on a timescale band of interest 
provides strong evidence that either the weather 
variable helped cause deer fluctuations and 
synchrony for those timescales, possibly 
indirectly, or some other weather variable, 
highly spatially 
coherent with the first variable, caused the 
fluctuations and synchrony. The highly significant 
coherence between deer abundances and DVCs we 
present in the main text is likewise probably 
indicative of greater deer abundances resulting in 
more DVCs. DVCs are too uncommon to meaningfully 
influence deer abundance (Fig. \ref{SM-FigS1_Timeseries}), 
and the 
in-phase relationship our analyses revealed  between 
deer abundances and DVCs on $3-7$-year timescales 
is more consistent with deer 
dynamics driving DVC dynamics than the reverse. 

We used spatial wavelet coherence to test 
five weather variables (average winter Tmin, 
Tmax, snow depth, precipitation, and WSI), six 
climate variables (winter and growing season PDO, 
MEI and NAO) and hunter abundance as potential 
drivers of synchrony in deer abundance over the 
$3-7$-year timescale band. We also used spatial wavelet
coherence to test for a relationship of deer 
abundance with DVCs, and deer abundance with DVCs 
adjusted for traffic volume. When the temporal and 
spatial extents of available data for two spatiotemporal variables 
were not identical, we retained only the 
locations and time range over which data were 
available for both variables when testing for 
spatial wavelet coherence. When testing for spatial 
wavelet coherence between a spatiotemporal variable and a 
climate index (PDO, MEI or NAO, each of which 
comprised only a single time series), we considered 
that the same index time series pertained to all 
locations for which the spatiotemporal variable was 
measured. 

For statewide analyses (SI Appendix), for which all variables 
were represented only by single, state-average or 
total time series rather than spatiotemporal data, timescale-specific relationships 
between variables were tested using ordinary 
wavelet coherence. Wavelet coherence is a 
non-spatial version of the spatial wavelet coherence, and 
tests whether two time series exhibit consistent 
phase differences and correlated magnitudes of 
oscillation through time, as a function of 
timescale. We tested for relationships between 
statewide deer abundance time series and climate 
indices, statewide averages for winter weather, 
statewide hunter abundance time series, and 
statewide total DVCs. The same functions in the `wsyn` package listed above for spatial wavelet coherence were used, as they can compute both ordinary wavelet coherence and spatial wavelet coherence.

\textbf{Wavelet Moran theorem, fractions of synchrony explained, and predicted synchrony.} We calculated the amount of deer abundance 
synchrony explained by individual weather and 
climate variables, as well as the amount of DVC 
synchrony explained by deer abundance, following 
the wavelet Moran theorem of Sheppard *et al.* [@Sheppard2016_aphids]. 
As part of this procedure, we also calculated the 
average "cross terms" of Sheppard *et al.* [@Sheppard2016_aphids] over 
the timescale band of interest (see supplementary 
files of ref @Sheppard2016_aphids, equation 22), again following the 
procedures described in that reference. The precise 
mathematical meaning of cross terms is explained by Sheppard *et al.* 
[@Sheppard2016_aphids]; for operational purposes, they 
provide a test of whether modelling assumptions of Sheppard *et al.*
[@Sheppard2016_aphids] are sufficiently met to use the 
wavelet Moran theorem to estimate fractions of 
synchrony explained. Cross terms less than $10\%$ 
of estimated fractions of synchrony explained were 
considered necessary to make statements about fractions of synchrony explained. Cross terms associated with a spatial wavelet coherence between a spatiotemporal variable and a non-spatial variable (such as one of the climate indices) were always $0$ by mathematical necessity [@Sheppard2016_aphids].

We calculated the amount of deer abundance 
synchrony jointly explained by snow depth, 
$\text{MEI}_{\text{w}}$, and $\text{PDO}_{\text{w}}$ using a wavelet multiple regression 
approach [@Sheppard_wavmodel]. This method, the detailed mathematics of which are described by Sheppard *et al.* [@Sheppard_wavmodel] along with an application, 
is analogous to multiple linear regression, with a 
model describing the relationship between a 
response variable and multiple predictor variables; 
but in contrast to standard regression, response 
and predictor variables are 
wavelet transforms of spatiotemporal variables. The approach gives several outputs which we use. First, the approach gives the fraction of synchrony (in our case, $`r round(abunmodelSyncExp,0)`\%$ for $3-7$-year timescales) in the response variable (county-level deer abundance) that can be explained by the synchrony of the predictors (snow depth, $\text{MEI}_{\text{w}}$, and $\text{PDO}_{\text{w}}$). The validity of this result is again subject to whether cross terms were small (less than $10\%$ of the fraction of synchrony explained). They were small, $`r round(abunmodelXterms,0)`\%$, for our model. The approach also gives a "predicted synchrony" plot (colors in Fig. \ref{Fig2_DeerSync}C), which is a function of time and timescale. The plot shows what the pattern of synchrony in the response variable (deer abundance) would look like if the only drivers of synchrony were the predictor variables. Predicted synchrony plots can be compared directly to the wavelet mean field magnitude plot of the response variable, similarity of the two plots giving a visual indication of whether the predictors in the model adequately explain synchrony in the response variable. The function `wlm` and related functions from `wsyn` implement these tools.

Proper partitioning of spatial synchrony, into a component explained by the synchronous drivers in a wavelet linear model and a component corresponding to factors independent of those drivers (residuals in the model), is only possible if cross terms, which also appear in the model, are small compared to the other components [@Sheppard2016_aphids; @Sheppard_wavmodel].
Large cross terms correspond to a general association between the model (drivers, or predictor variables) at one location and the residuals of the model at other locations [@Sheppard2016_aphids; @Sheppard_wavmodel].  Such associations may occur as a result of the population at the first location dispersing to the second location and being included in the count there. Considering a coarser spatial scale limits the potential for dispersal to have such an influence on our statistics, so one expects that if cross terms are large at the county spatial scale, the problem may be mitigated at the district scale, as happened in some
of our analyses (see section "USDA district results" in SI Appendix). 

\textbf{Randomization procedure for removing synchrony at $3-7$-year timescales.} In order to quantify the degree to which 
synchronizing effects are responsible for 
fluctuations in statewide total deer and DVC time 
series (Fig. \ref{Fig4_StateWideFluctuations} and associated text), 
we adapted the standard Fourier surrogate procedure 
of Theiler [@Theiler1992]. We constructed artificial 
deer abundance and DVC time series for each 
location, having the same variance and spectral 
characteristics as the original data but no spatial 
synchrony in the $3-7$-year timescale band of 
interest. Components of variability and synchrony 
outside this band (such as longer-term trends or 
fluctuations, and higher-frequency components 
associated with 'spikiness' in the data) were left 
unchanged. To produce surrogate data for a given 
location, we took the Fourier transform of the time 
series and randomized the phases of the Fourier 
components inside the $3-7$-year timescale band 
before inverse transforming to produce the 
surrogate time series. Because only phases of the 
Fourier transform were randomized, power spectra and 
variances of time series are unchanged. 
Additionally, because phases in the $3-7$-year 
timescale band are randomized independently, 
surrogate time series so produced no longer exhibit 
a tendency to synchrony in the $3-7$-year timescale 
range [@Schreiber2000]. Phase randomization was 
accomplished by adding a random and uniformly distributed phase increment 
between $0$ and $2\pi$ to the phase of each Fourier 
component with a frequency greater than or equal to 
$1/7$, but less than $1/3$ cycles per year. 
Additionally, because the phase values of the 
Fourier components of a real signal are 
anti-symmetric about the Nyquist frequency ($1/2$ 
cycles/year), Fourier components with frequencies 
less than or equal to $6/7$ and more than $2/3$ 
cycles per year had the same set of phase 
increments subtracted to preserve the anti-symmetry 
property of Fourier transforms. We summed each set 
of surrogate time series spatially to produce a 
surrogate state-total time series representative of 
the hypothetical case in which $3-7$-year timescale 
synchrony was absent. Such surrogate state-total 
time series were compared to the actual time series 
(deer abundance or DVCs) to show the nature of the 
additional variance in the state-total time series 
that was attributable to synchrony (Fig. \ref{Fig4_StateWideFluctuations}).

\textbf{Approximate monetary impact of DVCs.} We estimated the approximate monetary impact of 
synchrony in DVCs based on several reported 
assumptions about DVCs. Using an average difference in DVCs
between peak and trough years 
(Fig. \ref{Fig4_StateWideFluctuations}) of `r avgdvcfluc`, and if we assume 
that each reported DVC results in an average of 
US\$2024 of damage [@Conover2011_bookchp], the 
difference in damages statewide between peak and 
trough years is more than US\$5M. Given that DVCs 
may be only reported about 50% of the time 
[@Marcoux2010_dvc], this number is actually likely 
greater than US\$10M. Using different assumptions 
of the cost of DVCs [@Huijser2008_dvc], the 
difference between peak and trough years in DVC 
costs may exceed US\$22M.    

\textbf{Data Availability.} We have adopted an open and reproducible science approach throughout the project. As indicated in the sections above on individual methods, all computations were done using functions from the `wsyn` package for the R programming language, the source code of which is available from `www.github.com/reumandc/wsyn/CRANv1p0p0`. The package can be installed into R from CRAN or via `devtools::install_github(repo="reumandc/wsyn/CRANv1p0p0")`.
The `wsyn` package includes a vignette documenting how to use the functions implemented there, and giving examples; the vignette provides additional description of the methods of this paper. All data and code producing this paper are archived at `www.github.com/reumandc/TomAnderson_WIdeer`, including the complete revision history of the process that led to the presented analyses.
The suite of analyses leading to this paper can be re-run and the paper re-compiled at the touch of a button via the codes in that repository - see the `Readme.md` file in the repository for instructions.

**Acknowledgements:** D. Lyden, J. Steinglein and the WIDNR for data; D. Storm, T. Van Deelen, J. Millspaugh, and J. Fox for consultation. Funding was provided by the James S McDonnell Foundation, United States National Science Foundation grants 1442595 and 1714195, United States Department of Agriculture award 2016-67012-24694, and The Nature
Conservancy. 

**Author Contributions:** TLA and DCR conceived the project; RER provided assistance with data curation; TLA, DCR and LWS analyzed the data; TLA and DCR wrote manuscript; LWS, RER and JAW provided extensive editing and feedback on the analysis and manuscript. 

# References {-}

<div id="refs"></div>

\newpage

\begin{figure}[!h]  
\includegraphics[width=.6\textwidth]{Results/Fig1.png}
\caption{Theoretical simulation showing that timescale-specific synchrony between populations may cause cycling in the total population. Panels A-C show
three alternative scenarios for time series measured at 15 locations across a region, and differ in that
populations are unsynchronized (A), or are partially synchronized on short (B, 3-year period) or long (C, 10-year
period) timescales, though synchrony cannot readily be detected visually. Differences in synchrony produce
differences in variability and periodicity of the corresponding spatially summed population time series (D-F
correspond to A-C, respectively). Synchrony on a given timescale produces cycling on that timescale in the
total. If time series in A-C were deer populations in 15 locales so that D-F were the total population across the
region, differences among D-F could impact the hunting industry even though local differences across scenarios are
not so noticeable. Differences between D and E are due solely to differences in synchrony. The population at
location $i$ ($i=1,\ldots,15$) and time $t$ was  
$p_i(t) = \sin \left( \frac{2\pi t}{w}+2 \pi u_i \right) + \epsilon_i(t)$, where $w=3$ (A,B) or $10$ (C), 
$u_i = 0$ for all $i$ (B,C) or the $u_i$ are independent draws from a uniform distribution on $(0,1)$ (A), and the $\epsilon_i(t)$ are independent standard-normally distributed. 
Wavelet mean field magnitude plots (\textit{Methods}) computed from A-C (G-I, respectively) 
sensitively reveal the timescale 
structure of synchrony. Higher values mean greater synchrony at the given time on the given timescale. 
\label{Fig1_example}}
\end{figure}

\newpage

\begin{figure}[!h]
\includegraphics[height=.7\textheight]{Results/Fig2.png}
\caption{Wavelet mean field (A) and wavelet phasor mean field (B) magnitudes, and predicted synchrony (C) 
of deer abundance (\textit{Methods}). Higher values in A, B indicate greater synchrony (A) or phase synchrony (B) 
at the indicated times and timescales. Contours on (B) indicate statistical significance of phase synchrony at 
$P =$ 0.001. Horizontal lines in (A) and (B) indicate the timescales that were the focus of statistical 
analysis, $3-7$-years. Colors on (C) show synchrony in deer abundance predicted from a wavelet model using 
snow depth, $\text{MEI}_{\text{w}}$, and $\text{PDO}_{\text{w}}$ as predictors (\textit{Methods}), while contours in (C) show the 
actual wavelet mean field magnitude of deer abundance from (A), for comparison. Agreement is good, giving a 
visual indication of the statistical result that climatic Moran effects explain deer synchrony.
\label{Fig2_DeerSync}}
\end{figure}

\newpage

\begin{figure}[!h]
\includegraphics[height=.7\textheight]{Results/Fig3.png}
\caption{Wavelet mean field (A) and wavelet phasor mean field (B) magnitudes, and predicted synchrony (C) 
of DVCs. Colors, contours, and horizontal lines in A and B indicate the same as in 
Fig. \ref{Fig2_DeerSync}A,B, but for DVC data. Colors in (C) 
show synchrony in DVCs predicted by a wavelet model with sole predictor deer abundance, while contours in (C) 
show the actual wavelet mean field magnitude of DVCs from (A), for comparison. Agreement is good, giving a 
visual indication of the statistical result that deer synchrony explains DVC synchrony.
\label{Fig3_DvcSync}}
\end{figure}

\newpage

\begin{figure}[!h]
\includegraphics[width=.6\textwidth]{Results/Fig4.png}
\caption{Magnitude of the effects of synchrony on state-total deer abundance
and DVC fluctuations. Small panels show state-total deer abundance (A, in millions) and DVC (C, in thousands) time 
series (black lines) and surrogate totals obtained via a randomization procedure 
(\textit{Methods}) that eliminates synchrony among counties at $3-7$-year timescales 
(grey lines, 1000 surrogates plotted). Increased variation and semi-periodic 
oscillations due to synchrony are particularly notable when trends and long-timescale 
(longer than 7 years) variability are filtered out by subtracting the mean of the 
grey lines (B = deer, in thousands; D = DVCs). $\Delta$ represents the difference in values from 
the surrogate mean. Red arrows highlight examples of the strong cycling that is 
attributable to synchrony. For DVCs, the typical fluctuation in numbers of DVCs due 
to synchrony amounts to an estimated difference between peak and trough years of 
US\$$`r floor(round(2*avgdvccostsLow,-6))/1000000`-`r floor(round(2*avgdvccostsHigh,-6))/1000000`$Mn in damages 
statewide (\textit{Methods}). 
\label{Fig4_StateWideFluctuations}}
\end{figure}


