---
title: Synchronous effects produce major cycles in deer populations
  and deer-vehicle collisions
author:
 - Thomas L. Anderson\textsuperscript{1,2}, Lawrence W. Sheppard\textsuperscript{2}, Jonathan A. Walter\textsuperscript{2,3}, Robert E. Rolley\textsuperscript{4},  Daniel C. Reuman\textsuperscript{2,5}  
 - \textsuperscript{1}Department of Biology, Appalachian State University, 572 Rivers St., Boone, NC 28608, USA; \textsuperscript{2}Deparment of Ecology and Evolutionary Biology and Kansas Biological Survey, University of Kansas, 2101 Constant Ave, Lawrence, KS 66049, USA;  \textsuperscript{3}Department of Environmental Sciences, University of Virginia 291 McCormick Rd, Charlottesville, VA 22904, USA; \textsuperscript{4}Wisconsin Department of Natural Resources, 101 S. Webster St., Madison, WI 53707, USA; \textsuperscript{5}Laboratory of Populations, Rockefeller University, 1230 York Ave., New York, NY 10065, USA
date: 'correspondence to: anderstl@gmail.com, reuman@ku.edu'
output:
  pdf_document:
    fig_caption: yes
    keep_tex: yes
    includes:
      in_header: head_maintext.sty
  html_document:
    df_print: paged
  word_document: default
geometry: left=1cm,right=1cm,top=2.5cm,bottom=2.8cm
link-citations: yes
csl: ecologyletters.csl
tables: yes
bibliography: AndersDeerShortRefs.bib
urlcolor: blue

---
**Short title:** Spatial synchrony leads to populations cycles  
**Author Contributions:** TLA and DCR conceived the project; RER provided assistance with data curation; TLA, DCR and LWS analyzed the data; TLA and DCR wrote manuscript; LWS, RER and JAW provided extensive editing and feedback on the analysis and manuscript.  
**Type of Article:** Letter 
**Key Words:** population cycle, synchrony, Moran effect, population dynamics,  
**Number of words in abstract:**  
**Number of words in main text:**  
**Number of figures and tables:** 5 figures, 1 table  
**Number of references:**  

\newpage
<!--Title length limit 96 characters, inc spaces-->

```{r echo=F, results="hide"}
#load in results
options(scipen=6)
Tab1_results<-readRDS(file = "Results/Table1.rds")
Tab1_results<-cbind(Tab1_results[,1:3],apply(Tab1_results[,4:7],2,function(x){round(x,4)}))
abunwmeiSyncExp37<-round(Tab1_results$SynchronyExplained[Tab1_results$Predictor=="Winter MEI"&Tab1_results$Timescale=="3-7"],2)
abunwpdoSyncExp37<-Tab1_results$SynchronyExplained[Tab1_results$Predictor=="Winter PDO" &  Tab1_results$Timescale=="3-7"]
abunwmeiSyncExp47<-round(Tab1_results$SynchronyExplained[Tab1_results$Predictor=="Winter MEI"&Tab1_results$Timescale=="4-7"],2)
abunwpdoSyncExp47<-Tab1_results$SynchronyExplained[Tab1_results$Predictor=="Winter PDO" &  Tab1_results$Timescale=="4-7"]
snowabunSyncExp3_4<-Tab1_results$SynchronyExplained[Tab1_results$Predictor=="Snow Depth" & Tab1_results$Timescale=="3-4"]
abunmodelSyncExp<-readRDS(file="Results/abunmodelSyncExp.rds")

#load Table S1 results
TabS1_results<-readRDS(file="Results/TabS1_results.rds")
wmei_wpdo_pval<-round(TabS1_results$Pvalue[TabS1_results$Predictor=="Winter MEI" & TabS1_results$Response=="Winter PDO"],4)
wmei_wpdo_phase<-round(TabS1_results$MeanPhase[TabS1_results$Predictor=="Winter MEI" & TabS1_results$Response=="Winter PDO"],4)
abunwmeiSyncExp47<-Tab1_results$SynchronyExplained[Tab1_results$Predictor=="Winter MEI" &  Tab1_results$Timescale=="4-7"]

#load Table S2 results
TabS2_results<-readRDS(file = "Results/TableS2.rds")
usda_dvcabunSyncExp<-TabS2_results$SynchronyExplained[TabS2_results$Response=="DVCs" &
                                                        TabS2_results$Predictor=="Abundance"]
usda_adjdvcabunSyncExp<-TabS2_results$SynchronyExplained[TabS2_results$Response=="Traffic-adj. DVCs" &
                                                        TabS2_results$Predictor=="Abundance"]
usda.snowabunSyncExp3_4<-TabS2_results$SynchronyExplained[TabS2_results$Response=="Abundance" &  TabS2_results$Predictor=="Snow Depth" & TabS2_results$Timescale=="3-4"]
avgdeerfluc<-readRDS(file="Results/totaldeerfluctuations.rds")
avgdvcfluc<-readRDS(file="Results/totaldvcfluctuations.rds")
avgdvccostsLow<-readRDS(file="Results/totaldvccostsLow.rds")
avgdvccostsHigh<-readRDS(file="Results/totaldvccostsHigh.rds")
```

<!--***DAN: Dan has not edited this yet for Ecol Lett; TOM: I started editing- there is a 150 word limit-->
#Abstract
Population cycles are fundamentally linked with spatial synchrony, with the
prevailing paradigm being that populations whose dynamics are intrinsically cyclic are easy to synchronize. That is, population cycles help give rise to spatial synchrony. Here we demonstrate this process can work in reverse, with synchrony causing population cycles. We show that timescale-specific environmental effects, by synchronizing local population dynamics on certain timescales only, cause major population cycles over large areas in white-tailed deer. An important aspect of the new mechanism is specificity of synchronizing effects to certain timescales, which causes local dynamics to sum across space to a substantial cycle on those timescales. We also demonstrate for the first time that synchrony can be transmitted from environmental drivers to populations (deer) and human systems (DVCs). Because synchrony of some environmental drivers may be changing, possibly as part of climate change, subsequent changes to population cycles may arise via the mechanism described here.
<!--OLD ABSTRACT: Population cycles are a profound enigma of the natural world, the causes of which have been investigated for over 100 years. An early hypothesized explanation of cycles, for the iconic lynx-hare and for other systems, posited that cyclic environmental phenomena such as sun spots drive population cycles. But that hypothesis was refuted, and for several decades research focused instead on explanations involving nonlinear or delayed density dependence, often via species interactions. Here we show that timescale-specific environmental effects, by synchronizing local population dynamics on certain timescales only, cause major population cycles over large areas in white-tailed deer and deer-vehicle collisions (DVCs). We thereby demonstrate a new mechanism of cycles that centers once again on environmental influences. The mechanism differs fundamentally from earlier hypotheses that relied on a strongly cyclic driver such as sun spots. An important aspect of the new mechanism is specificity of synchronizing effects to certain timescales, which causes local dynamics to sum across space to a substantial cycle on those timescales. We also demonstrate for the first time that synchrony can be transmitted, in a cascading effect, from environmental drivers, to populations (deer), to human systems (DVCs). The cycles we explain can have substantial effects on agriculture, hunting, and accident recovery industries. Research shows that synchrony of some environmental drivers may be changing, possibly as part of climate change. Such changes may alter population cycles that arise via the mechanism described here.-->

<!--***DAN: Old first para
**Correlated fluctuations through time in abundances of distant populations, called spatial synchrony, is 
a widespread and important phenomenon in population ecology [@Leibhold2004_sync]. Synchrony of exploited 
or economically important taxa should, theoretically, have substantial impacts for humans, because
synchrony can produce high-amplitude oscillations through time in total populations across large 
regions. However, synchrony has seldom been investigated empirically for exploited species, 
and its anthropogenic importance is not often considered. Additionally, it has historically been 
difficult to distinguish drivers of synchrony in natural populations of any 
taxa[@Kendall2000_moran;@Abbott2007_moran], though improved analytical methods that decompose 
synchrony by timescale have furthered these 
efforts[@Allstad2015_gypsysync;@Sheppard2016_aphids;@Sheppard2017_surrogates;@Walter2017_geogsync]. 
Here we identify the causes of spatial synchrony in the dynamics of populations of white-tailed deer, 
a widely exploited species, and its consequences on deer-vehicle collisions (DVCs) across Wisconsin, 
USA. Deer synchrony occurred principally at $3-7$-year timescales of fluctuation, and synchronous 
winter environmental variation explained almost 90% of deer synchrony through a process called the 
Moran effect[@Moran1953;@Ranta1997_moran]. DVCs were in turn synchronized by deer populations. 
Timescale-specific synchrony caused large semi-periodic fluctuations in statewide total deer 
populations and DVCs that would not have occurred without synchrony. Thus synchrony caused local 
dynamics to sum to a major cyclic phenomenon on statewide spatial scales. Our results are 
principally important as a demonstration of a new and potentially widespread mechanism for 
population cycles on large spatial scales, a phenomenon of long-standing interest.**
-->

<!--Authors must submit a 120-word-maximum statement about the significance of their research paper written at a level understandable to an undergraduate-educated scientist outside their field of specialty. The primary goal of the Significance Statement is to explain the relevance of the work in broad context to a broad readership. Significance statements are not required for Brief Reports. TOM: My first attempt below is currently 140 words-->
<!-- **Significance Statement:** Understanding the mechanisms behind cycles in population abundances has long been a goal in ecology. Here we demonstrate, in white-tailed deer, a new mechanism by which population cycles can arise, and the consequences of cycles for humans. Fluctuations in deer abundances in different counties of Wisconsin, USA, were highly correlated due to correlated fluctuations in winter climate. Importantly, these synchronized dynamics occurred only on 3-7-year timescales. Deer synchrony in turn produced synchronized oscillations in deer-vehicle collisions across Wisconsin, again with 3-7-year periodicity. When aggregated spatially, fluctuations for individual counties reinforced each other on 3-7-year timescales, producing state-wide cycles. In addition to revealing new mechanisms, our results are important for hunting, insurance industries, and ecosystems influenced by deer.    -->

#Introduction

The causes of population cycles have interested and mystified ecologists for over 100 years 
[@Myers2018_cycles]. An early hypothesis for the cause of cycles, for the iconic Canadian 
lynx-snowshoe hare system as 
well as for other systems, posited that cyclic environmental phenomena such as sun spots drive 
population cycles [@Sinclair1992_sunspots;@Selas2004_sunspots]. However, that hypothesis lacked 
empirical support [@Nilssen2007_nosunspots;@Myers2013_nosunspots], and for several decades 
research has focused instead on explanations involving nonlinear or delayed density dependence, 
often via predator-prey interactions, parasitism, and disease 
[@Krebs2018_harereview;@Krebs2013_book;@Myers2013_nosunspots;@MartinezPadilla2014_grouse]. 

Population cycles are fundamentally linked with spatial synchrony, the 
tendency for geographically distinct populations to 
exhibit correlated temporal fluctuations. 
Theory predicts that individual cyclic populations can easily be brought into synchrony via weak 
coupling from dispersal or via shared exogenous forcing of weather (Moran effects), in a 
process known as nonlinear phase locking [@Bjornstad2000]. Empirical evidence from laboratory 
experiments [@VassuerFox_2009;@HopsonFox_2019] and analysis of observational data 
[@Haynes_etal2019] support these predictions. Indeed, the 
lynx-hare system features predator-prey cycles thought to be synchronized regionally 
by phase locking through Moran effects [@stenseth1999;@stenseth2004]. 
As a consequence, the prevailing paradigm is that cycles and synchrony are linked because 
populations whose dynamics are intrinsically cyclic are easy to synchronize; that is, population 
cycles help give rise to spatial synchrony.

We here propose and provide evidence for a mechanism by which the direction of causality can be 
reversed: instead, spatial synchrony can give rise to population cycles. The
mechanism comprises the following two conceptual steps. First, an exogenous driver synchronizes non-cyclic local
population fluctuations *on a particular timescale*. Many large climate phenomena
fluctuate, either episodically or consistently, with emphasis on particular timescales. 
For example, from ca.
1965-1995, fluctuations in the North Atlantic Oscillation (NAO) contained a substantial 
10-year oscillatory component 
[@Sheppard2016_aphids], and the El Ni$\~{n}$o Southern Oscillation (ENSO) also has 
periodic content. Thus local weather variables influenced by 
climatic oscillations can contain periodic components which are synchronous across
large areas. These components may be weak and therefore hardly noticable, 
in comparison with asynchronous,
local fluctuations with which they are combined.
Synchronous periodic components in local weather can nevertheless propogate, via Moran
effects, to produce weak but synchronous periodic components 
in local population dynamics. Growing evidence
indicates that timescale-specific spatial synchrony of populations is probably common 
[@Defriez2016_plankton;@Sheppard2016_aphids;@Defriez2017_marine;@Defriez2017_terrestrial;@Sheppard2017_surrogates;@Walter2017_geogsync;@Anderson2019_LO;@Sheppard_wavmodel;@Haynes_etal2019]. 
Second, the weak periodic components
of local population dynamics, being synchronous across space, reinforce each other
in the total population computed across the region, whereas stronger, asynchronous
components on other timescales cancel in the spatial total. Thus periodicities
in dynamics which may be unnoticable at the local spatial scale become
dominant at the regional scale. This effect can be strong, as we will see.
The key here is that synchrony occurs *on a particular timescale*.
Dynamics of the total, regional population are then dominated by a cycle on that timescale. 
Fig. \ref{Fig1_example} provides an initial example of this phenomenon using a
simple model. A theoretical approach to the mechanism is summarized below and elaborated in
SI section \ref{SM-SI:theory}. The bulk of the paper is spent examining empirical evidence for the mechanism.
<!--***DAN: Tom, all references to the sup mat need to be made consistent with this: see SI section X,
or see SI Fig. X, or see SI table X, where the X is an automated link. I have been doing this when I see 
these things, but some searches need to be done to make it comprehensive.-->

One major applied concern is that if
fluctuations in local time series do become amplified in a
timescale-specific manner as outlined above, outbreaks of pests or disease
could follow [@Rohani_etal1999]. For example, timescale-specific
synchrony of local populations of pests may produce major cycles in the
regional total pest population and therefore in the damage caused by the pests. 
Timescale-specific
synchrony could also be important for industries, such as hunting,
that exploit multiple populations across space if it causes cycling in
the regional total population. 
For instance, if time series of Fig. \ref{Fig1_example}D-F were deer populations 
in counties within a state so that 
Fig. \ref{Fig1_example}J-L are state-total populations, differences between the 
three scenarios of Fig. \ref{Fig1_example} could substantially
impact the hunting industry.
Such cycles may then also cause cascading effects 
on human economic and social systems. Cascading effects of synchrony
on human concerns, though potentially important, are seldom 
discussed in the literature. 

The ecological system for which we explore the importance of these ideas 
centers on white-tailed deer (*Odocoileus virginianus*) populations across
the state of Wisconsin (WI), USA, and on rates of deer-vehicle collisions (DVCs).
White-tailed deer is a species of considerable interest due to its major 
economic and cultural benefits and impacts on ecosystems [@Cote2004_deerimpacts;@Conover2011_bookchp]. Across 
the USA in 2011, US\$$34$Bn were spent on hunting big game, including deer [@usfws2011_survey]. 
In WI, hunting-related economic activity accounts for ~$1\%$ of state GDP [@usfws2011_survey]. DVCs cause about 
50 human fatalities and US\$$1.5$Bn in vehicular damage annually across the USA [@Conover2011_bookchp]. Thus it 
is an important applied question what drives the year-to-year variability in statewide deer abundance and DVC 
totals. We consider potential drivers of deer populations including both environmental factors and the direct 
effects of hunting. We use data from 1981-2016 (Fig.  \ref{Fig2_Timeseries}) aggregated at three 
spatial resolutions (*Methods*): counties ($n=71$), United States Department of Agricultural (USDA) districts 
($n=9$), and statewide (SI Fig. \ref{SM-FigS1_map}). 

Here we demonstrate, using deer data, how timescale-specific
spatial synchrony can produce regional population cycles. We also show that
timescale-specific deer population synchrony begets timescale-specific synchrony
in DVCs across the state, producing cycles in 
state-total DVC time series as well. These cycles have economic and public-health
consequences. Fluctuations in deer population abundances in
the 71 counties of Wisconsin were synchronous due to
fluctuations in winter climate, and synchrony 
occurred only on 3-7 year timescales. Deer synchrony in turn
produced synchrony in DVCs on 3-7 year timescales. 
Timescale-specific synchrony caused large fluctuations on 3-7-year
timescales in state-total time series for both deer and DVCs. 
Our results provide a theoretical description and strong empirical support for 
the importance of a new mechanism of population cycles, and for its
potential importance for human concerns. In contrast to the well-understood
mechanism by which cycles can beget synchrony, our new mechanism shows 
that synchrony, when it is specific to certain timescales, can also produce
cycles.

# Methods

\textbf{Summary of theory.} The statement of our main theoretical result requires a conceptual understanding 
of the *power spectrum* (or, simply, *spectrum*) of a population or other 
time series, and of the *cospectrum* of two time series. The spectrum $S_{w_iw_i}(f)$ 
of a time series or stochastic process measured in location $i$, $w_i(t)$,
is a function of frequency, $f$. The frequency of a periodic oscillation is one over
its timescale, i.e., one over its period. The value of the spectrum $S_{w_iw_i}(f)$ is larger for 
frequencies (or timescales) on which $w_i(t)$ oscillates with greater strength. So, 
for instance, a population that exhibits a strong (rspectively, weak) 10-year-timescale 
oscillations will have a spectrum with a large (respectively, small) peak at 
the frequency $1/10$. The spectrum decomposes the variance, $\var(w_i(t))$, according 
to the timescales on which it occurs, in the sense that an integral of 
$S_{w_iw_i}(f)$ across all frequencies equals $\var(w_i(t))$. The cospectrum
of two time series, $w_i(t)$ and $w_j(t)$, decomposes their covariance, $\cov(w_i(t),w_j(t))$,
in a similar way: the cospectrum, $C_{w_iw_j}(f)$, is a function of frequency
with peaks occurring at frequencies for which oscillations in $w_i(t)$ and
$w_j(t)$ are strong and coincident in their phases, i.e., synchronous.

The spectrum is useful for studying population cycling because peaks in the 
spectrum of a population indicate cyclic behavior. 
Our main theoretical result expresses the the spectrum, $S_{\text{tot}}$ of a total
population across a region composed of $i=1,\ldots,N$ local areas in the form
$$S_{\text{\emph{tot}}}(f) = F_{\text{popdyn}}(f) \left[ \sum_{i=1}^N S_{\epsilon_{i}\epsilon_{i}}(f) +
\sum_{\substack{i,j=1 \\ i \neq j}}^N C_{\epsilon_{i}\epsilon_{j}}(f)\right],\label{mainresmt}$$ 
\noindent where $F_{\text{popdyn}}(f)$ reflects 
the nature of population dynamics within the local areas, and the $S_{\epsilon_i\epsilon_i}$
are spectra and the $C_{\epsilon_i\epsilon_j}$ are cospectra of environmental noise time 
series in the local areas. If dynamics are not intrinsically cycling, 
then $F_{\text{popdyn}}(f)$ has no peaks and does not contribute to whether $S_{\text{tot}}$
is peaked. The first term in brackets in \ref{mainresmt},
$\sum_{i=1}^N S_{\epsilon_{i}\epsilon_{i}}(f)$, encapsulates the effects of 
periodicities in local environmental noise time series on whether the total 
population exhibits cycling. The second term in the brackets,
$\sum_{i \neq j} C_{\epsilon_{i}\epsilon_{j}}(f)$, encapsulates
the effects of timescale-specific synchrony on whether the total population
exhibits cycling. The key observation here is that, when $N$ is large so that the
region being considered is combination of a large number of local areas, the first term
is likely to make a much smaller contribution to $S_{\text{tot}}$ than is the second
term; the first term is a sum of only $N$ quantities and the second is a sum over
$N^2-N$, which is much greater than $N$ for large enough $N$. Thus theory predicts
that timescale-specific patterns of synchrony are very important for whether
spatial-total populations exhibit synchrony. Of course which of the two terms
in brackets in \ref{mainresmt} is larger depends on the specific values of the summands
contributing to each term as well as on the number of summands. But we explore 
reasonable scenarios in the Supporting Information
and find that the initial intuition is correct, for the quite-general model setup that we
study, that typically timescale-specific synchrony is a very important determinant of
whether a regional population cycles. See SI section \ref{SM-SI:theory}
for the the general theoretical development and 
SI Figs \ref{SM-TheoryExample1}, \ref{SM-TheoryExample2},
\ref{SM-TheoryExample3} for specific examples.

\textbf{Empirical Data.} We calculated deer abundance estimates based on data from deer management units provided by the Wisconsin Department of Natural Resources (WIDNR) spanning the `r dim(cty.list$Abun)[2]`-yr period `r minyear`-`r maxyear` (Fig. \ref{Fig2_Timeseries}) for `r length(unique(dat$COUNTY_NAM))` counties. We also calculated estimated hunter abundance also from data from WINDR for each county for the years 1992-`r maxyear`
(Fig. \ref{Fig2_Timeseries}); these data constituted an index of the estimated 
number of hunters during the first two days of the 9-day gun 
hunting season [@widnr2001].

We obtained estimates of DVCs for each county, spanning the 
years 1987-`r maxyear`, from the WI Department 
of Transportation (WIDOT; Fig. \ref{Fig2_Timeseries}). These data reflect 
the number of police-investigated DVCs and are only an 
index of DVCs, as opposed to an absolute number of incidents, 
since not all DVCs are reported [@Marcoux2010_dvc]. We also obtained traffic volume estimates (millions of miles driven) for each county spanning the years 1988-`r maxyear` from the WIDOT, to control for traffic in our analysis of DVCs.

We obtained daily weather data (snow depth, 
precipitation, and minimum and maximum air temperature - here called Tmin and Tmax) from all available weather stations in each county from the 
National Climatic Data Center ([`https://www.ncdc.noaa.gov/`](https://www.ncdc.noaa.gov/)). We 
averaged daily values across the winter ($\text{Dec}_{t-1}$ 
to $\text{Mar}_t$) and across stations within a county to produce 
annual time series of winter weather by county which were used in 
all subsequent analyses (Fig. \ref{Fig2_Timeseries}). 

We obtained monthly estimates of three global climate indices: 
the North Atlantic Oscillation (NAO), the multivariate 
El Ni$\~{n}$o Southern Oscillation Index (MEI), and the Pacific 
Decadal Oscillation (PDO). All of these have been shown to 
influence large ungulate population dynamics in the north central 
USA [@Post1998_popgrowth;@Post2004_greenland]. 
We averaged monthly values across the 
approximate growing season (Apr-Nov), and the winter 
($\text{Dec}_{t-1}$ to $\text{Mar}_t$) for each index 
(Fig. \ref{Fig2_Timeseries}). High values of these indices generally correspond to mild winter weather over WI [@Rodionov2003_elnino]. We used a total of 11 weather and climate variables in subsequent analyses: winter Tmin, Tmax, 
snow depth, and precipitation; winter NAO, MEI, and PDO; 
growing-season NAO, MEI and PDO; and the WSI. 

\textbf{Data pre-processing.} We performed all analyses described below and in the main text 
at three spatial scales: counties (N = 44-`r dim(cty.list$Abun)[1]` depending on 
the variables involved; mean county 
area $\pm$ SD = `r round(mean(gps$Area_km2),0)` $\pm$ `r round(sd(gps$Area_km2),0)` $\text{km}^2$), 
United States Department of Agriculture (USDA) districts ($N = 9$; mean 
district area $\pm$ SD = `r round(mean(usda.centroid$Area_km2),0)` $\pm$ `r round(sd(usda.centroid$Area_km2),0)` $\text{km}^2$), 
and statewide (Fig. \ref{SM-FigS1_map}). Analyses at three spatial scales 
were performed because of concerns on the accuracy of abundance estimation methods at smaller spatial extents [@Millspaugh2009; see SI for further discussion]. USDA district data were produced as the sum or average of county-level variables, as districts were always conglomerates of counties (Fig. \ref{SM-FigS1_map}). 
Statewide analysis used state total or average time series for each 
variable. In general, results remained unchanged in their main substance using data aggregated to larger spatial scales (Table \ref{SM-TabS2_SummaryDistrict}, 
Fig. \ref{SM-FigS7_WMFAndPredictionsDistricts}), suggesting results for counties were not meaningfully influenced by possible inaccuracies in SAK abundance estimation procedures at the county level.

We performed an optimal Box-Cox transformation on each time series to help ensure Gaussian marginal distributions [@Sheppard2016_aphids], which are necessary for
the Fourier-surrogate-based spatial wavelet coherence and wavelet linear
regression procedure described below. This process also included linear detrending
(and subtracting the mean) to remove any long-term trends that might otherwise
obscure patterns of synchrony [@Buonaccorsi2001]. Time series
variances were also standardized to 1. 

\textbf{Analysis.} We use wavelet transforms in most analyses described below,
which are ideal for detecting periodic components in time series data
[@addison2002]. The methods we use are recently developed extensions to analytical
methods that are standard tools in some biological fields and well established.
The mathematical details are described elsewhere
[@Sheppard2013;@Sheppard2016_aphids] and below we provide operational descriptions
of what each method takes as input, what it provides as output, and how to
interpret the information. All analyses were done via functions in the `wsyn`
package for the R programming language [@Rstats2018].

We characterized the spatial synchrony of spatiotemporal variables (deer
abundance, DVCs and weather) using wavelet mean field magnitudes and wavelet
phasor mean field (SI Appendix Fig.
\ref{SM-PedagogFigWMF}) [@Sheppard2013;@Sheppard2016_aphids]. Both methods take as input
standardized (see above) timeseries data from multiple locations and provide as
output complex-numbered value functions of time and timescale. When plotted, these
values characterize the degree of synchrony among locations in phase and amplitude
(mean field magnitude) and phase (phasor mean field) (Figs \ref{Fig1_example}, SI Appendix Fig. \ref{SM-PedagogFigWMF}). They also show where in the time series the synchrony
occurs, and at which timescales synchrony is more prevalent. The statistical
significance of phasor mean fields can be compared to simulated data sets
containing random phase distributions [@Sheppard_wavmodel]. In this study we
compared phasor mean fields to 1000 simulated data sets, using a significance
threshold  of $P < 0.001$ (one-tailed test). 

For both county- and district-level data, we assessed pairwise relationships
between spatiotemporal variables using the spatial wavelet coherence technique
([@Sheppard2013;@Sheppard2016_aphids;@Sheppard2017_surrogates]; SI Appendix Fig.
\ref{SM-PedagogFigSpatCoh}). Spatial wavelet coherence tests whether two
spatiotemporal variables exhibit consistent phase differences and correlated
magnitudes of oscillation through time and across space, as a function of
timescale. Significance is determined by comparing spatial wavelet coherence of
real data with spatial wavelet coherences made from appropriately randomized,
Fourier surrogate data sets ($N=$ `r nsurrogs`) that retain the same temporal and
spatial autocorrelation properties as the original time series but are otherwise
unrelated to each other [@Theiler1992;@Schreiber2000;@Sheppard2013;@Sheppard2016_aphids]. We calculated how often observed spatial wavelet coherence between pairs of variables was ranked higher than coherences of surrogate data sets, aggregating across the timescale band of interest using an
average-rank-based method described in detail elsewhere [@Sheppard2016_aphids]. A
$p$-value of $0.05$ means that the empirical spatial wavelet coherence was higher
than the same quantity calculated for 95% of the surrogate data sets,
representing the null hypothesis of no relationship between variables. Ranks of
actual spatial wavelet coherences in surrogate spatial wavelet coherences were
also plotted as a function of timescale (Fig. \ref{SM-FigS4_PhaseDiffs}). Finally,
when two variables had significant spatial wavelet coherence over a timescale
band, we calculated the mean phase difference between the variables over the band,
$\bar{\theta}$, in units of $\pi$ radians, to determine if oscillations had
approximately in-phase ($\bar{\theta} = 0$ to $0.25$), quarter-cycle
($\bar{\theta} =  0.25$ to $0.75$), or anti-phase ($\bar{\theta} =  0.75$ to $1$)
relationships. 

We used spatial wavelet coherence to test five weather variables (average winter
Tmin, Tmax, snow depth, precipitation, and WSI), six climate variables (winter and
growing season PDO, MEI and NAO) and hunter abundance as potential drivers of
synchrony in deer abundance over the $3-7$-year timescale band, where the
majority of deer synchrony was present (see Results) We also used spatial wavelet
coherence to test for a relationship of deer abundance with DVCs, and deer
abundance with DVCs adjusted for traffic volume.

We calculated the amount of deer synchrony explained by individual weather and
climate variables, as well the amount of DVC synchrony explained by deer
synchrony, following the wavelet Moran theorem of Sheppard *et al.* [@Sheppard2016_aphids].
We also examined the amount of deer synchrony jointly explained by snow depth, winter MEI, and winter PDO (the three primary drivers of deer synchrony; see *Results*) using
wavelet multiple regression [@Sheppard_wavmodel]. This method is analogous to
multiple linear regression, with a model describing the relationship between a
response variable and multiple predictor variables [@Sheppard_wavmodel]. However,
all predictor and response inputs are wavelet transforms of spatiotemporal
variables. This approach also can produce a time- and timescale-specific plot of
what the pattern of synchrony in the response variable (deer abundance) would look
like if the drivers of synchrony were the predictors. This predicted synchrony
plot can then be visually compared to the actual wavelet mean field magnitude
plot, giving a visual indication of the degree of model adequacy. As part of these procedures, we evaluted model "cross terms", which are the component of  synchrony that is due to variability explained by the drivers at a given location being synchronous with residuals at other locations [@Sheppard2016_aphids;@Sheppard_wavmodel]. Cross terms must be small (< 10%) in order to inferences about synchrony to be valid.  

To quantify the degree to which synchronizing effects are responsible for
fluctuations in statewide total deer and DVC time series, we adapted the standard
Fourier surrogate procedure of Theiler [@Theiler1992]. We constructed artificial
time series data for both variables for each location that had the same spectral
characteristics are the original data, except that they were not synchronized over
$3-7$-year timescale band of interest. We accomplished this by randomizing the phases of the Fourier transforms in the surrogate data sets. We summed each set of
surrogate time series spatially to produce surrogate statewide totals for both
deer and DVCs, representing the hypothetical case of no synchrony over $3-7$-year
timescales. We then compared these surrogate time series to their respective
actual time series to show the nature of the additional variance in the
state-total series that was attributable to synchrony. 

\textbf{Data Availability.} All data and code will be deposited into Dryad digital repository. All data and code producing this paper are also archived at `www.github.com/reumandc/TomAnderson_WIdeer`, including the complete revision history of the process that led to the presented analyses.
The suite of analyses leading to this paper can be re-run and the paper re-compiled at the touch of a button via the codes in that repository - see the `Readme.md` file in the repository for instructions.

#Results

The wavelet mean field of county-level deer abundances revealed that the most prominent synchrony feature occurred at $3-7$-year timescales (Fig. \ref{Fig3_DeerSync}A). A
related technique, the wavelet phasor mean field, demonstrated the strong statistical significance of this phenomenon (Fig.
\ref{Fig3_DeerSync}B). We therefore focused our analysis on the $3-7$-year timescale band.
<!-- TOM: Moved this bit to methods. Most past analyses of synchrony have used correlation methods that conflate timescales, but recent results show this can hinder detection of -->
<!-- synchronous phenomena by obscuring important synchrony on some timescales with asynchronous or noisy dynamics on other timescales [@Sheppard2016_aphids]. Moreover, synchrony must be specific to certain timescales to produce cyclic dynamics of -->
<!-- total populations on those timescales (Fig. \ref{Fig1_example}).   -->

Spatial wavelet coherence, wavelet regression techniques, and a wavelet Moran theorem demonstrated that 
deer abundance synchrony on $3-7$-year timescales was principally caused by climatic Moran effects. 
<!-- Tom: moved to Discussion para on timescale specifity. Although it is known that a spatially synchronous environmental driver can induce 
population synchrony via Moran effects [@Leibhold2004_sync; @Moran1953;  
@Ranta1997_moran], historically it has been  
difficult to attribute synchrony  
in natural populations to specific environmental drivers versus other potential  
causes of synchrony such as dispersal [@Kendall2000_moran;@Abbott2007_moran].  
The analytical advances listed above can facilitate more precise  
identification of Moran drivers  
[@Sheppard2016_aphids;@Sheppard2017_surrogates;@Walter2017_geogsync;@Sheppard_wavmodel].  
Their effectiveness is due in part to their capacity to decompose  
synchrony according to the distinct timescales on which it can occur (Fig.  
\ref{Fig1_example}G-I). -->
Synchrony in deer abundance was related to winter snow depth 
(spatial wavelet coherence, $P =$ `r round(Tab1_results$Pvalue[Tab1_results$Predictor=="Snow Depth" &Tab1_results$Timescale=="3-7"],2)`, 
Table \ref{Tab1_CountyCoherenceSummary}), in a near anti-phase relationship (mean phase 
difference between abundance and snow depth across $3-7$-year timescales, in units of ${\pi}$ 
radians, was $\bar{\theta} =$ `r round(Tab1_results$MeanPhase[Tab1_results$Predictor=="Snow Depth"&Tab1_results$Timescale=="3-7"],2)`; SI Appendix Fig. \ref{SM-FigS4_PhaseDiffs}D): deer 
abundances tended to peak in years of below-average snow. No other weather variables were statistically significant with deer abundance (Table \ref{Tab1_CountyCoherenceSummary}). Wavelet mean field and 
wavelet phasor mean field magnitude plots for all
winter weather variables are in 
Fig. \ref{SM-FigS8_WMFsAndWPMFsAncillary}. Deer abundance was also related, with
approximately quarter-cycle phase shifts, 
to the multivariate El Ni$\~{n}$o Southern Oscillation index
($P =$ `r round(Tab1_results$Pvalue[Tab1_results$Predictor=="Winter MEI"&Tab1_results$Timescale=="3-7"],2)`, $\bar{\theta}  =$ `r round(Tab1_results$MeanPhase[Tab1_results$Predictor=="Winter MEI"&Tab1_results$Timescale=="3-7"],2)`) and to the Pacific Decadal Oscillation 
($P =$ `r round(Tab1_results$Pvalue[Tab1_results$Predictor=="Winter PDO"&Tab1_results$Timescale=="3-7"],2)`, $\bar{\theta}  =$ `r round(Tab1_results$MeanPhase[Tab1_results$Predictor=="Winter PDO"&Tab1_results$Timescale=="3-7"],2)`, 
Table \ref{Tab1_CountyCoherenceSummary}, Fig. \ref{SM-FigS4_PhaseDiffs}A,C), both averaged over the winter months
($\text{Dec}_{t-1}$ to $\text{Mar}_t$, abbreviations $\text{MEI}_{\text{w}}$ and
$\text{PDO}_{\text{w}}$). The two indices were strongly coherent and in-phase with
each other on $3-7$-year timescales over the period 1981-2016 (wavelet coherence, $P =$ `r round(TabS1_results$Pvalue[TabS1_results$Predictor=="Winter MEI" & TabS1_results$Response=="Winter PDO" ],4)`, $\bar{\theta}  =$ `r round(TabS1_results$MeanPhase[TabS1_results$Predictor=="Winter MEI" & TabS1_results$Response=="Winter PDO" ],2)`, SI Appendix Table \ref{SM-TabS1_PairwiseClimVarSpatCoh}). 
At $5$-yr periodicity, the quarter-cycle phase lags correspond to deer populations
peaking about `r round(-5*Tab1_results$MeanPhase[Tab1_results$Predictor=="Winter PDO"&Tab1_results$Timescale=="3-7"]/2,1)` or 
`r round(-5*Tab1_results$MeanPhase[Tab1_results$Predictor=="Winter MEI"&Tab1_results$Timescale=="3-7"]/2,1)` years after mild winters. Applying the wavelet Moran theorem, snow depth, 
$\text{MEI}_{\text{w}}$, and $\text{PDO}_{\text{w}}$ together explained 
$`r round(abunmodelSyncExp,0)`\%$ of deer synchrony on $3-7$-year timescales. 
See Fig. \ref{Fig3_DeerSync}C for a visual manifestation of this statistical result. None of the other climate indices were related to deer abundance (Table \ref{Tab1_CountyCoherenceSummary}). 

Snow depth and climate index results seemed to indicate distinct or partly distinct phenomena 
synchronizing deer on different timescales within the $3-7$-year range. Again applying the wavelet
Moran theorem, snow depth explained $`r round(usda.snowabunSyncExp3_4,0)`\%$ of synchrony in deer
abundance at $3-4$-year timescales when data were aggregated to USDA districts (SI Appendix text and Table \ref{SM-TabS2_SummaryDistrict}). Dynamics at $3-4$-year timescales were not well explained by either $\text{MEI}_{\text{w}}$ or $\text{PDO}_{\text{w}}$, which instead contributed more
to deer synchrony over $4-7$-year timescales (Table \ref{Tab1_CountyCoherenceSummary}, SI Appendix Table \ref{SM-TabS2_SummaryDistrict}, Fig. \ref{SM-FigS5_RanksInSurrogs}A). 
$\text{MEI}_{\text{w}}$ and $\text{PDO}_{\text{w}}$ explained $`r round(abunwmeiSyncExp47,)`\%$ and $`r round(abunwpdoSyncExp47,0)`\%$ of deer synchrony on $4-7$-year timescales, respectively (SI Appendix Table \ref{Tab1_CountyCoherenceSummary}). Snow depth did not explain $4-7$-year synchrony: spatial wavelet coherence between snow depth and deer abundance on those timescales was non-significant (Table \ref{Tab1_CountyCoherenceSummary}, \ref{SM-TabS2_SummaryDistrict}).

Deer abundance synchrony was not caused by hunting on $3-7$-year timescales. County-level hunter and deer abundances were not significantly spatially coherent 
($P =$ `r round(Tab1_results$Pvalue[Tab1_results$Predictor=="Hunters" & Tab1_results$Timescale=="3-7"],2)`) over $3-7$-year timescales 
(Table \ref{Tab1_CountyCoherenceSummary}, Figs \ref{SM-FigS5_RanksInSurrogs}B, \ref{SM-FigS6_WMFHunter}). However, hunter abundance was significantly related to deer abundance on shorter timescales, around 
$2-2.5$-years ($P =$ `r round(Tab1_results$Pvalue[Tab1_results$Predictor=="Hunters" & Tab1_results$Timescale=="2-2.5"],3)`), and were in approximate anti-phase with deer abundance 
on those timescales ($\bar{\theta}  =$ `r round(Tab1_results$MeanPhase[Tab1_results$Predictor=="Hunters" & Tab1_results$Timescale=="2-2.5"],2)`; 
Table \ref{Tab1_CountyCoherenceSummary}, Figs \ref{SM-FigS4_PhaseDiffs}G,
\ref{SM-FigS5_RanksInSurrogs}B). 

<!-- Tom: Moved to discussion: Having attributed the bulk of the synchrony in deer abundance at $3-7$-year timescales to climatic Moran effects, we also found strong coherence between deer abundance and DVCs on $3-7$-year timescales: Moran effects pass from climate to deer to DVCs in a cascading manner, and in this way are ultimately responsible for synchrony of DVCs. -->The wavelet mean field and phasor mean field of DVCs revealed time-and timescale-specific synchrony, and visual patterns were remarkably similar to those shown by deer (compare Figs \ref{Fig3_DeerSync}A,B and \ref{Fig4_DvcSync}A,B). Spatial wavelet coherence showed that deer abundance and DVCs were significantly related 
($P =$ `r round(Tab1_results$Pvalue[Tab1_results$Response=="DVCs" & Tab1_results$Predictor=="Abundance"],3)`, 
Table \ref{Tab1_CountyCoherenceSummary}) on $3-7$-year timescales, and had a nearly in-phase 
relationship ($\bar{\theta} =$ `r round(Tab1_results$MeanPhase[Tab1_results$Response=="DVCs" & Tab1_results$Predictor=="Abundance"],2)`, SI Appendix Fig. \ref{SM-FigS4_PhaseDiffs}E). 
<!--Tom: Moved to discussion. Greater deer abundance almost certainly caused greater DVCs rather 
than DVCs driving deer fluctuations because DVCs were too infrequent to meaningfully impact deer 
abundance (Fig. \ref{Fig2_Timeseries}), and if they did an anti-phase relationship would have 
been expected.--> Deer abundance explained $`r round(usda_dvcabunSyncExp,0)`\%$ of the synchrony in 
DVCs when data were aggregated to USDA districts (SI Appendix text and Table \ref{SM-TabS2_SummaryDistrict}). 
As a visual manifestation of this statistical result, patterns of synchrony in DVCs predicted by a
wavelet regression model with sole predictor deer abundance match the patterns actually present in DVC
synchrony (Fig. \ref{Fig4_DvcSync}C). 
Statistical significance, phase relationships, and the amount of DVC synchrony explained by deer
abundance remained similar after adjusting DVCs to account for
fluctuations in traffic volume (Table \ref{Tab1_CountyCoherenceSummary}, SI Appendix text and Table \ref{SM-TabS2_SummaryDistrict}, Fig. \ref{SM-FigS4_PhaseDiffs}F).

<!-- Tom: Moved to discussion. As mentioned above, timescale-specific synchrony may lead, theoretically, to cyclic  
dynamics in spatial-total or spatial-average time series, which can mean  
alternating relative abundance and scarcity of important species over large areas  
[@Allstad2015_gypsysync]. Our data show the empirical reality of this effect and its  
strength for deer and DVCs. -->Statewide oscillations due to synchrony correspond to a typical difference of $~ `r round(avgdeerfluc,-4)`$ deer and $~ `r round(avgdvcfluc,-2)`$ 
DVCs between peak and trough years 
(Fig. \ref{Fig5_StateWideFluctuations}). Importantly, these high amplitude 
statewide oscillations in deer and DVCs disappear (Fig. 
\ref{Fig5_StateWideFluctuations}) when synchrony between counties on $3-7$-year 
timescales is removed via a randomization procedure; spatial synchrony causes local dynamics to sum to a substantial 
semi-periodic oscillatory phenomenon on statewide spatial scales. Depending on 
assumptions made about costs and reporting rates of DVCs 
[@Huijser2008_dvc;@Marcoux2010_dvc;@Conover2011_bookchp], statewide DVC damages 
vary by US\$$`r floor(round(2*avgdvccostsLow,-6))/1000000`-`r floor(round(2*avgdvccostsHigh,-6))/1000000`$Mn 
between peak and trough years. <!--Tom: Moved to discussion. Similarly, fluctuations of a quarter 
million deer likely have considerable consequences for agricultural crop damage 
and hunting success, both of which can have socioeconomic impacts. See 
SI Appendix for further details on the cycling mechanism.-->

#Discussion

Overall, we have demonstrated that timescale-specific Moran effects, by 
synchronizing local dynamics, can cause major cycles in the 
regional total population. This new mechanism differs fundamentally from prior,
now refuted hypotheses that population cycles in are due to highly regular
environmental fluctuations such as sun spots [*e.g.*,
@Sinclair1992_sunspots;@Selas2004_sunspots]because the new mechanism does not
require a highly regular environmental 
oscillator. Rather, it is sufficient simply for irregular environmental drivers 
such as winter climatic factors to be coherent with populations in a timescale 
band, thereby inducing synchrony in that band and leading to total-population cycles on the same timescales.  
We have also presented the first demonstration that synchrony can be transmitted in a 
Moran-like effect from ecological to human systems (deer to DVCs). The applied ramificatons of synchrony are primarily known from theory in the context of metapopulation collapse when all populations are low simultaneously [@Earn2000] or in human disease dyanmics [@Rohani1999], but our demostration provides a quantitive projection of monetary impacts resulting from synchronized animal populations.

There are two alternative elaborations of the mechanisms we present by which
synchronous effects of environmental drivers produce state-wide population cycles in deer.
First, synchronous fluctuations in environmental drivers associated with winter climate may have produced
small oscillations on $3-7$-year timescales in county-level deer population time series, and 
also brought these oscillations into spatial synchrony, thereby causing them to sum, across
counties, to produce statewide fluctuations. Second, the local population oscillations on $3-7$-year timescales may have been due partly to the intrinsic dynamics of deer, resulting from density-dependent effects, maternal effects, or 
other mechanisms of nonlinear populations dynamics. Synchronous environmental drivers 
then brought the intrinsically oscillating local dynamics into synchrony. This second hypothesis is the pattern typically described from theoretical and experimental studies, where exogenous drivers or dispersal couple intrinsically cyclic populations via nonlinear phase-locking [@Bjornstad2000;@VassuerFox_2009;@HopsonFox_2019]. However, recent evidence from observational studies on gypsy moths also support the hypothesis of cyclic-dynamics-leading-to-synchrony [@Haynes_etal2019]. The
difference between these two alternatives is whether local population
oscillations were produced and synchronized, or merely synchronized but 
not produced, by synchronous environmental drivers. These alternative hypothesis would be difficult to 
distinguish, but the available evidence makes the hypothesis of synchrony-leading-to-cycles more likely for our system. Snow depth and large-scale climatic oscillations have previously been demonstrated to induce synchrony in deer and other ungulates [@Mysterud2008_deerclimate; @Hegel2012_caribou; @Post2004_greenland; @Grotan2005_roedeer]. We are also unaware of evidence to indicate that deer populations are intrinsically cyclic. Under either alternative, a Moran effect caused the observed strong, 
state-wide oscillations in the sense that it synchronized weak local oscillations, 
causing them to sum to a major phenomenon. Eliminating synchrony via randomizations also
eliminated the state-wide oscillations (Fig. \ref{Fig5_StateWideFluctuations}), even though the
randomizations preserved the structure of the local fluctuations. 

An important aspect in all of the phenomena we describe is the specificity of fluctuations in both population and environmental data to certain timescales. Population cycles would not have manifested in deer had their synchrony not been specific to certain timescale ($3-7$-years). Timescale specificity in synchrony is common [@Sheppard2016_aphids;@Sheppard2017_surrogates;@Walter2017_geogsync;@Sheppard_wavmodel;@Haynes_etal2019;@Anderson2019_LO], making our mechanism of population cycles likely to be highly generalizable and the effects to not be limited to the system we have studied. 
The analytical methods used here [@Sheppard2016_aphids;@Sheppard2017_surrogates;@Walter2017_geogsync;@Sheppard_wavmodel] facilitated more precise 
identification of synchrony drivers, with their effectiveness occuring in part due to their capacity to decompose synchrony according to the distinct timescales on which it can occur (Fig. \ref{Fig1_example}G-I). Historically it has been difficult to distinguish environmental drivers versus other potential causes of synchrony such as dispersal [@Kendall2000_moran;@Abbott2007_moran], but the timescale-specific methods applied here effectively demonstrate how such causality can be inferred. 

Large-scale climatic patterns and local weather conditions are well known to influence spatial synchrony via Moran effects[@Leibhold2004_sync]. Both processes have specfically been attributed to synchrony of large ungulates across multiple regions, including reindeer, caribou [@post2002;@aanes2003;@Post2004_greenland;@Hegel2012_caribou], roe deer[@Grotan2005_roedeer], and musk oxen [@post2002]. Here, winter snow depth,  $\text{MEI}_{\text{w}}$ and $\text{PDO}_{\text{w}}$ cycles were the primary predictors of deer synchrony. Furthermore, deer synchrony lagged both climate indices at a temporal scale of about two years, which is similar to other reported lag times of climatic effects on deer 
population dynamics [@Post1998_popgrowth;@forchhammer1998]. Winter weather and climate were likely important because deeper snow limits access to winter food resources in ungulates, which subsequently affects body condition, leading to greater susceptibility to starvation or predators and reduced parturition rates [@Mysterud2008_deerclimate;@Hegel2012_caribou]. In WI, this situation likely played out to a greater extent in northern areas, where snow depth is greater and access to high quality food resources (i.e., agricultral crops) is available. Determining whether such landscape or habitat differences across regions may further decompose how climatic Moran efects influence the spatial patterning synchrony in deer [@Walter2017_geogsync]. Recent work on several taxa 
[@Allstad2015_gypsysync;@Sheppard2016_aphids;@Koenig2016_birdsync;@Shestakova2016_forestsync] shows that synchrony of some environmental drivers may be changing, possibly due to climate change. These changes may alter population synchrony, 
and therefore also population cycles that arise by our new mechanism. In our case study, the predicted increase in the frequency of strong El Ni$\~{n}$o events [@Cai2014_elnino] may lead to altered cycling period in deer and DVCs in WI.

Hunters were not the principle driver of deer synchrony. These results are consistent with the
hypothesis that hunters can regulate deer populations over short timescales and small spatial areas [@fryxell1991, @denicola2008] but have more limited effects over longer timescales and at large spatial extents [@brown2000,@simard2013]. The lack of longer-timescale spatial wavelet coherence between hunters and deer
abundance is likely also related to the facts that hunters typically prefer to shoot bucks, the removal of
which has limited population effects; and the state regulates antlerless harvest in an attempt to
maintain populations at a desirable size.

Our study shows a novel way in which synchrony contributes to applied ecology, beyond the previously demonstrated hypotheses of metapopulation collapse [@Earn]. Having attributed the bulk of the synchrony in deer abundance at $3-7$-year timescales to climatic
Moran effects, we also found strong coherence between deer abundance and DVCs on $3-7$-year
timescales: Moran effects pass from climate to deer to DVCs in a cascading manner, and in this way
are ultimately responsible for synchrony of DVCs. Greater deer abundance almost certainly caused
greater DVCs rather than DVCs driving deer fluctuations because DVCs were too infrequent to
meaningfully impact deer abundance (Fig. \ref{Fig2_Timeseries}), and if they did an anti-phase
relationship would have been found, as opposed to the in-phase relationship we observed. While this result is perhaps intuitive, especially as DVCs are sometimes used as a
proxy for deer population size [@Grotan2005_roedeer], it has important consequences for
humans, namely that statewide DVCs occur in a semi-period manner, tracking fluctuations in
total deer populations. Thus, as climate change continues to modify ecosytems, ensuing changes for
human industries that rely on wildlife and plant poulations seem likely. Changes in winter climate
have aleady been implicated as important for forestry practices in WI [@Rittenhouse2015]; other industries that exploit wildlife and plant populations would also be affected, and in some cases, like white-tailed deer, with semi-periodic regularity. Similarly, fluctuations of a quarter 
million deer statewide likely will have considerable consequences for agricultural crop damage or forest regeneration [@Cote2004_deerimpacts]
and hunting success, both of which can have socioeconomic impacts. 

\textbf{Conclusion.} Our study demonstrates a novel mechanism of population cycles through the cumulative dynamics of synchronized local populations of an exploited species, with important consequences for humans. Further investigation of systems and drivers of synchrony that could lead to similar region-wide dynamics will help unravel the generality of this phenomena. Investigating synchrony and regional fluctuations of species that have impacts on humans (e.g., crop pests) would be particularly vaulable, especially to complement work in the traditional paradigm  of cycles leading to synchrony [@Haynes_etal2019]. 

<!--Load sup mat workspace for the methods, which used to be in the sup mat-->
```{r echo=F, results="hide"}
load(file="WholeWorkspace.RData")
```


**Acknowledgements:** D. Lyden, J. Steinglein and the WIDNR for data; D. Storm, T. Van Deelen, J. Millspaugh, and J. Fox for consultation. Funding was provided by the James S McDonnell Foundation, United States National Science Foundation grants 1442595 and 1714195, United States Department of Agriculture award 2016-67012-24694, and The Nature
Conservancy. 


# References {-}

<div id="refs"></div>

\newpage


```{r Table1,echo=F,results='markup'}
options(knitr.kable.NA = '') #makes it put blanks instead of "NA" for NAs
rownames(Table1)<-NULL
knitr::kable(Table1,
             format='pandoc',
             digits=4,
             col.names=c("Predictor","Response","Timescale", "P-value","Mean phase ($\\bar{\\theta}$)","Synchrony explained","Average cross terms"),
             caption = "Summary of spatial wavelet coherence tests and related results using county-level data. $P$-values and mean phases are from tests of spatial wavelet coherence (*Methods*) of the variables in columns 1 and 2 over the indicated timescale band. Mean phases over the timescale band (units of $\\pi$ radians) are provided only when meaningful, i.e., when spatial wavelet coherence was significant, $P < 0.05$, or marginally significant. A positive mean phase indicates the variable in the column *Response* is leading, and a negative mean phase indicates the variable in the column *Predictor* is leading. The *Synchrony explained* and *Average cross terms*  columns, which give percentages, are based on wavelet linear regressions (*Methods*) of the *Response* transforms against the *Predictor* transforms. When cross terms are low, synchrony explained is interpretable as the approximate percentage of synchrony in the response explainable as due to the synchrony in the predictor. Cross terms were low enough except for tests involving DVCs, for which USDA district-level results were more reliable (main text and Table \\ref{TabS2_SummaryDistrict}). Cross-terms are zero for test involving climate indices because for those tests the same values were used for all locations (*Methods*). All values were rounded to display four digits. \\label{Tab1_CountyCoherenceSummary}",
             booktabs=T)
```

\newpage

\begin{figure}[!h]  
\includegraphics[width=.5\textwidth]{Results/Fig1.png}
\caption{Simulations demonstrating our mechanism whereby
synchrony can cause population cycling. In scenario 1, 
environmental noise in location $i$ ($i=1,\dots,71$; first $15$ locations
shown in A) was modeled as 
$e^{(1)}_i(t)=\sin(2 \pi t/w^{(1)}+2 \pi u^{(1)}_i)+3\epsilon^{(1)}_i(t)$, where $w^{(1)}=3$, 
the $u^{(1)}_i$ were independent
draws from the uniform distribution on $(0,1)$, and the local noises $\epsilon^{(1)}_i(t)$ were
standard normal and independent across locations and times. In scenario 2,
noise ($15$ locations shown in B) 
was $e^{(2)}_i(t)=\sin(2 \pi t/w^{(2)}+2 \pi u^{(2)})+3\epsilon^{(2)}_i(t)$
where $w^{(2)}=3$, $u^{(2)}$ was one single standard-normally distributed random number used for all
locations, and the $\epsilon^{(2)}_i(t)$ were again
standard normal and independent.
Noise for scenario 3 (C) was like scenario 2 but used $w^{(3)}=10$. Thus the noises in 
scenarios 2 and 3 had weak and not visually obvious timescale-specific synchrony and those in 1 had no synchrony.
Populations were modelled as $p^{(s)}_{i}(t+1)=\rho p^{(s)}_i (t)+e^{(s)}_i(t)$ using $\rho=0.4$
for scenarios $s=1,2,3$, and are displayed (D-F) for locations 1-15.
Although synchrony is again weak and 
cannot readily be detected visually, wavelet mean field magnitide plots (Methods)
revealed that synchrony was present, in scenarios 2 and 3, on 3- and 10-year
timescales (G-I); on such plots, higher values mean greater synchrony at the given 
time on the given timescale. Total populations (J-L) show clear cycling in scenarios 2 and 3,
of periods $3$ and $10$, respectively, consistent with the proposed mechanism (Introduction): 
synchrony on a given timescale 
produces cycling on that timescale in the total population.
We used $71$ locations the empirical analyses of this study focus on white-tailed
deer populations in Wisconsin, and Wisconsin has 71 counties for which deer 
data were available. Time series on A-F were
shifted vertically for readability.
\label{Fig1_example}}
\end{figure}

\newpage

\begin{figure}[h!]
\includegraphics[width=\textwidth]{Results/Fig2.png}
\caption{Raw time series of (A) deer abundance, (B) deer-vehicle collisions (DVCs), 
(C) average winter ($\text{Dec}_{t-1}$ to $\text{Mar}_t$) snow depth, and 
(D) winter climate indices (Multivariate El Ni$\~{n}$o Southern Oscillation Index, 
MEI; Pacific Decadal Oscillation, PDO), and hunter abundances (E). 
Each line in A-C, E represents a county. Other variables were 
used in analyses (see text) but are not shown here due to lack of significant 
assocations with the deer and DVC variables that were the focus of the 
study.\label{Fig2_Timeseries}}
\end{figure}

\newpage

\begin{figure}[!h]
\includegraphics[height=.7\textheight]{Results/Fig3.png}
\caption{Wavelet mean field (A) and wavelet phasor mean field (B) magnitudes, and predicted synchrony (C) 
of deer abundance (\textit{Methods}). Higher values in A, B indicate greater synchrony (A) or phase synchrony (B) 
at the indicated times and timescales. Contours on (B) indicate statistical significance of phase synchrony at 
$P =$ 0.001. Horizontal lines in (A) and (B) indicate the timescales that were the focus of statistical 
analysis, $3-7$-years. Colors on (C) show synchrony in deer abundance predicted from a wavelet model using 
snow depth, $\text{MEI}_{\text{w}}$, and $\text{PDO}_{\text{w}}$ as predictors (\textit{Methods}), while contours in (C) show the 
actual wavelet mean field magnitude of deer abundance from (A), for comparison. Agreement is good, giving a 
visual indication of the statistical result that climatic Moran effects explain deer synchrony.
\label{Fig3_DeerSync}}
\end{figure}
<!--***DAN: Tom, the label for this used to be Fig2_DeerSync but now it has been changed to
Fig3_DeerSync. However, I found at least one citation to the old label that now does not
work (I fixed it). This is why we do not include numbers in figure labels. It defeats the
purpose of having labels and automatic numbering! Henceforth please use labels without 
numbers that reflect the content of the figure/table/whatever.-->

\newpage

\begin{figure}[!h]
\includegraphics[height=.7\textheight]{Results/Fig4.png}
\caption{Wavelet mean field (A) and wavelet phasor mean field (B) magnitudes, and predicted synchrony (C) 
of DVCs. Colors, contours, and horizontal lines in A and B indicate the same as in 
Fig. \ref{Fig3_DeerSync}A,B, but for DVC data. Colors in (C) 
show synchrony in DVCs predicted by a wavelet model with sole predictor deer abundance, while contours in (C) 
show the actual wavelet mean field magnitude of DVCs from (A), for comparison. Agreement is good, giving a 
visual indication of the statistical result that deer synchrony explains DVC synchrony.
\label{Fig4_DvcSync}}
\end{figure}

\newpage

\begin{figure}[!h]
\includegraphics[width=.6\textwidth]{Results/Fig5.png}
\caption{Magnitude of the effects of synchrony on state-total deer abundance
and DVC fluctuations. Small panels show state-total deer abundance (A, in millions) and DVC (C, in thousands) time 
series (black lines) and surrogate totals obtained via a randomization procedure 
(\textit{Methods}) that eliminates synchrony among counties at $3-7$-year timescales 
(grey lines, 1000 surrogates plotted). Increased variation and semi-periodic 
oscillations due to synchrony are particularly notable when trends and long-timescale 
(longer than 7 years) variability are filtered out by subtracting the mean of the 
grey lines (B = deer, in thousands; D = DVCs). $\Delta$ represents the difference in values from 
the surrogate mean. Red arrows highlight examples of the strong cycling that is 
attributable to synchrony. For DVCs, the typical fluctuation in numbers of DVCs due 
to synchrony amounts to an estimated difference between peak and trough years of 
US\$$`r floor(round(2*avgdvccostsLow,-6))/1000000`-`r floor(round(2*avgdvccostsHigh,-6))/1000000`$Mn in damages 
statewide (\textit{Methods}). 
\label{Fig5_StateWideFluctuations}}
\end{figure}


