---
title: Synchronous effects produce cycles in deer populations
  and deer-vehicle collisions
author:
 - Thomas L. Anderson\textsuperscript{1,2}, Lawrence W. Sheppard\textsuperscript{2}, Jonathan A. Walter\textsuperscript{2,3}, Robert E. Rolley\textsuperscript{4},  Daniel C. Reuman\textsuperscript{2,5}  
 - \textsuperscript{1}Department of Biology, Appalachian State University, 572 Rivers St., Boone, NC 28608, USA; \textsuperscript{2}Deparment of Ecology and Evolutionary Biology and Kansas Biological Survey, University of Kansas, 2101 Constant Ave, Lawrence, KS 66049, USA;  \textsuperscript{3}Department of Environmental Sciences, University of Virginia 291 McCormick Rd, Charlottesville, VA 22904, USA; \textsuperscript{4}Wisconsin Department of Natural Resources, 101 S. Webster St., Madison, WI 53707, USA; \textsuperscript{5}Laboratory of Populations, Rockefeller University, 1230 York Ave., New York, NY 10065, USA
date: 'correspondence to: anderstl@gmail.com, reuman@ku.edu'
output:
  pdf_document:
    fig_caption: yes
    keep_tex: yes
    includes:
      in_header: head_maintext.sty
  html_document:
    df_print: paged
  word_document: default
geometry: left=1cm,right=1cm,top=2.5cm,bottom=2.8cm
link-citations: yes
csl: ecologyletters.csl
tables: yes
bibliography: AndersDeerShortRefs.bib
urlcolor: blue
---

\noindent Short title: Spatial synchrony leads to populations cycles  
Author Contributions: TLA and DCR conceived the project; RER provided assistance with data curation; TLA, DCR and LWS analyzed the data; TLA and DCR wrote manuscript; all authors provided extensive editing and feedback on the analysis and manuscript.  
Type of Article: Letter  
Key Words: population cycle, synchrony, Moran effect, population dynamics, white-tailed deer  
Number of words in abstract: 150  
Number of words in main text: 5000  
Number of figures and tables: 5 figures, 1 table  
Number of references: 58  
<!--***DAN: I updated this on 2020 05 29--> <!--***Tom: will double check for the final submission-->

\textbf{Data accessibility.} All data will be made available in the Dryad digital repository. 
<!--***DAN: Tom to mention in the cover letter that data will be so deposited when the paper
is accepted, and Tom also to replace XYZ with the repo number when the time comes, please.-->
All data and code producing this paper are also archived at `www.github.com/reumandc/TomAnderson_WIdeer`.
Analyses leading to the paper can be re-run and the paper re-compiled 
via the codes there.

\newpage
<!--Title length limit 96 characters, inc spaces-->

```{r echo=F, results="hide"}
#load in results
options(scipen=6)
Tab1_results<-readRDS(file = "Results/Table1.rds")
Tab1_results<-cbind(Tab1_results[,1:3],apply(Tab1_results[,4:7],2,function(x){round(x,4)}))
#Tab1_MT<-Tab1_results[Tab1_results$Timescale=="3-7",]

#make synchrony explained and cross terms NA for non-spatial drivers
for(i in 1:nrow(Tab1_results)){
  if(grepl("Winter",Tab1_results$Predictor[i])||grepl("Summer",Tab1_results$Predictor[i])){
    Tab1_results$SynchronyExplained[i]<-NA
    Tab1_results$CrossTerms[i]<-NA
  }
}
Tab1_MT<-Tab1_results
abunwmeiSyncExp37<-round(Tab1_results$SynchronyExplained[Tab1_results$Predictor=="Winter MEI"&Tab1_results$Timescale=="3-7"],2)
abunwpdoSyncExp37<-Tab1_results$SynchronyExplained[Tab1_results$Predictor=="Winter PDO" &  Tab1_results$Timescale=="3-7"]
abunwmeiSyncExp47<-round(Tab1_results$SynchronyExplained[Tab1_results$Predictor=="Winter MEI"&Tab1_results$Timescale=="4-7"],2)
abunwpdoSyncExp47<-Tab1_results$SynchronyExplained[Tab1_results$Predictor=="Winter PDO" &  Tab1_results$Timescale=="4-7"]
snowabunSyncExp3_4<-Tab1_results$SynchronyExplained[Tab1_results$Predictor=="Snow Depth" & Tab1_results$Timescale=="3-4"]
snowabunSyncExp3_7<-Tab1_results$SynchronyExplained[Tab1_results$Predictor=="Snow Depth" & Tab1_results$Timescale=="3-7"]
abunmodelSyncExp<-readRDS(file="Results/abunmodelSyncExp.rds")

#load Table S1 results
TabS1_results<-readRDS(file="Results/TabS1_results.rds")
wmei_wpdo_pval<-round(TabS1_results$Pvalue[TabS1_results$Predictor=="Winter MEI" & TabS1_results$Response=="Winter PDO"],4)
wmei_wpdo_phase<-round(TabS1_results$MeanPhase[TabS1_results$Predictor=="Winter MEI" & TabS1_results$Response=="Winter PDO"],4)
abunwmeiSyncExp47<-Tab1_results$SynchronyExplained[Tab1_results$Predictor=="Winter MEI" &  Tab1_results$Timescale=="4-7"]

#load Table S2 results
TabS2_results<-readRDS(file = "Results/TableS2.rds")
usda_dvcabunSyncExp<-TabS2_results$SynchronyExplained[TabS2_results$Response=="DVCs" &
                                                        TabS2_results$Predictor=="Abundance"]
usda_adjdvcabunSyncExp<-TabS2_results$SynchronyExplained[TabS2_results$Response=="Traffic-adj. DVCs" &
                                                        TabS2_results$Predictor=="Abundance"]
usda.snowabunSyncExp3_4<-TabS2_results$SynchronyExplained[TabS2_results$Response=="Abundance" &  TabS2_results$Predictor=="Snow Depth" & TabS2_results$Timescale=="3-4"]
avgdeerfluc<-readRDS(file="Results/totaldeerfluctuations.rds")
avgdvcfluc<-readRDS(file="Results/totaldvcfluctuations.rds")
avgdvccostsLow<-readRDS(file="Results/totaldvccostsLow.rds")
avgdvccostsHigh<-readRDS(file="Results/totaldvccostsHigh.rds")
```

#Abstract
Population cycles are fundamentally linked with spatial synchrony, the
prevailing paradigm being that populations with cyclic dynamics are easily synchronized. 
That is, population cycles help give rise to spatial synchrony. Here we demonstrate this process can work in reverse, 
with synchrony causing population cycles. We show that timescale-specific environmental effects, by synchronizing local 
population dynamics on certain timescales only, cause major population cycles over large areas in white-tailed deer. An 
important aspect of the new mechanism is specificity of synchronizing effects to certain timescales, which causes local 
dynamics to sum across space to a substantial cycle on those timescales. We also demonstrate, to our knowledge for the 
first time, that synchrony can be transmitted not only from environmental drivers to populations (deer), but also 
from there to human systems (deer-vehicle collisions). Because synchrony of drivers may be altered by climate change, 
changes to population cycles may arise via our mechanism.

<!--Authors must submit a 120-word-maximum statement about the significance of their research paper written at a level understandable to an undergraduate-educated scientist outside their field of specialty. The primary goal of the Significance Statement is to explain the relevance of the work in broad context to a broad readership. Significance statements are not required for Brief Reports. TOM: My first attempt below is currently 140 words-->
<!-- **Significance Statement:** Understanding the mechanisms behind cycles in population abundances has long been a goal in ecology. Here we demonstrate, in white-tailed deer, a new mechanism by which population cycles can arise, and the consequences of cycles for humans. Fluctuations in deer abundances in different counties of Wisconsin, USA, were highly correlated due to correlated fluctuations in winter climate. Importantly, these synchronized dynamics occurred only on 3-7-year timescales. Deer synchrony in turn produced synchronized oscillations in deer-vehicle collisions across Wisconsin, again with 3-7-year periodicity. When aggregated spatially, fluctuations for individual counties reinforced each other on 3-7-year timescales, producing state-wide cycles. In addition to revealing new mechanisms, our results are important for hunting, insurance industries, and ecosystems influenced by deer.    -->

#Introduction

Population cycles have intrigued ecologists for over 100 years 
[@Myers2018_cycles]. An early hypothesis for the cause of cycles, for the Canadian 
lynx-snowshoe hare system and 
other systems, posited that strongly cyclic environmental phenomena such as sun spots may drive 
population cycles [@Sinclair1992_sunspots;@Selas2004_sunspots]. The sun spot hypothesis was 
refuted, however [@Nilssen2007_nosunspots;@Myers2013_nosunspots], and whereas it is now recognized that environmental drivers can alter cycles through their interactions with intrinsic population
processes [@watson2000;@kausrud2008], for decades 
research has focused instead on explanations of the origin of cycles 
which involve nonlinear or delayed density dependence, 
often via predator-prey interactions and parasitism 
[@kausrud2008;@Krebs2018_harereview;@Krebs2013_book;@Myers2013_nosunspots;@MartinezPadilla2014_grouse].

Population cycles are fundamentally linked with spatial synchrony, the 
tendency for geographically distinct populations to 
exhibit correlated temporal fluctuations. 
Theory predicts that cyclic populations can be synchronized via weak 
coupling from dispersal or via shared exogenous forcing (Moran effects), in a 
process known as nonlinear phase locking [@Bjornstad2000]. Empirical evidence from laboratory 
experiments [@VassuerFox_2009;@HopsonFox_2019] and observational data 
[@Haynes_etal2019] support these predictions. Indeed, the 
lynx-hare system features cycles thought to be synchronized regionally 
by phase locking through Moran effects [@stenseth1999;@stenseth2004]. 
As a consequence, the prevailing paradigm is that cycles and synchrony are linked because 
populations whose dynamics are intrinsically cyclic are easy to synchronize: population 
cycles help cause spatial synchrony.

We here propose a mechanism by which the direction of causality can be 
reversed: instead, spatial synchrony can give rise to population cycles. The
mechanism comprises the following two steps. First, an exogenous driver synchronizes non-cyclic local
population fluctuations *on a particular timescale*. Many large climate phenomena
fluctuate, episodically or consistently, with emphasis on particular timescales. 
For example, from ca.
1965-1995, the North Atlantic Oscillation (NAO) contained a substantial 
10-year oscillatory component 
[@Sheppard2016_aphids]. Thus local weather variables influenced by 
climatic oscillations can contain periodic components which are synchronous across
large areas. These components may be weak and therefore hardly noticable, 
in comparison with asynchronous,
local fluctuations with which they are combined.
Synchronous periodic components in local weather can nevertheless propogate, via Moran
effects, to produce weak, synchronous periodic components 
in local population dynamics. Timescale-specific spatial synchrony of populations is probably common 
[@Defriez2016_plankton;@Sheppard2016_aphids;@Defriez2017_marine;@Defriez2017_terrestrial;@Sheppard2017_surrogates;@Walter2017_geogsync;@Anderson2019_LO;@Sheppard_wavmodel;@Haynes_etal2019]. 
Second, the weak periodic components
of local population dynamics, being synchronous across space, reinforce each other
in the total regional population, whereas stronger, asynchronous
components on other timescales cancel in the spatial total. Thus periodicities
in dynamics which may be unnoticable at the local spatial scale become
dominant at the regional scale. 
The key here is that synchrony occurs *on a particular timescale*.
Dynamics of the total, regional population are then dominated by a cycle on that timescale. See
Fig. \ref{Fig1_example} for a simulated example. A theoretical approach is summarized below and elaborated in
Supporting Information (SI) section \ref{SM-SI:theory}. The bulk of the paper 
presents empirical evidence for the mechanism.

One potential applied concern is that if
fluctuations in local time series do become amplified in a
timescale-specific manner as outlined above, outbreaks of pests or disease
could follow [@Rohani_etal1999]. For example, timescale-specific
synchrony of local populations of pests may produce cycles in the
regional total pest population and therefore in the damage they cause. 
Timescale-specific
synchrony could also be important for industries, such as hunting,
that exploit multiple populations across space if it causes cycling in
the regional total population. 
If the time series of Fig. \ref{Fig1_example}D-F were deer populations 
in counties within a state so that 
Fig. \ref{Fig1_example}J-L are state-total populations, differences between the 
scenarios of Fig. \ref{Fig1_example} could
impact the hunting industry.
Cycles may then also cause cascading effects 
on human economic and social systems. 

The ecological system for which we explore the importance of these ideas 
centers on white-tailed deer (*Odocoileus virginianus*) populations across
the state of Wisconsin (WI), USA, and on rates of deer-vehicle collisions (DVCs).
White-tailed deer is a species of interest due to its
economic and cultural benefits and ecosystem impacts [@Cote2004_deerimpacts;@Conover2011_bookchp]. Across 
the USA in 2011, US\$$34$Bn were spent hunting big game, including deer [@usfws2011_survey]. 
In WI, hunting-related economic activity is ~$1\%$ of state GDP [@usfws2011_survey]. DVCs cause about 
50 human fatalities and US\$$1.5$Bn in vehicular damage annually across the USA [@Conover2011_bookchp]. Thus it 
is an important question what drives the year-to-year variability in statewide deer abundance and DVC 
totals. We consider potential drivers of deer populations including both environmental factors and the direct 
effects of hunting. We use data from 1981-2016 (Fig.  \ref{Fig2_Timeseries}) aggregated at three 
spatial resolutions: counties ($n=71$), United States Department of Agricultural (USDA) districts 
($n=9$), and statewide (*Methods*). 

Here we demonstrate, using theory and deer data, how timescale-specific
spatial synchrony can produce regional population cycles. We also show that
timescale-specific deer population synchrony begets timescale-specific synchrony
in DVCs across the state, producing cycles in 
state-total DVCs as well. Fluctuations in deer population abundances in
the counties of Wisconsin were synchronous due to
fluctuations in winter climate, and synchrony 
occurred mainly on $3-7$-year timescales. Deer synchrony in turn
produced synchrony in DVCs on $3-7$-year timescales. 
Timescale-specific synchrony caused substantial fluctuations on $3-7$-year
timescales in both state-total deer populations and DVCs. 
Our results provide a theoretical description and empirical support for 
a new mechanism of cycles, and for its
potential applied importance. In contrast to the well-understood
mechanism by which cycles can beget synchrony, our mechanism shows 
that synchrony, when specific to certain timescales, can also produce
cycles.

# Summary of theory

The statement of our main theoretical result requires a conceptual understanding 
of the *power spectrum* (or, simply, *spectrum*) of a population or other 
time series, and of the *cospectrum* of two time series. 
The spectrum is useful for studying population cycling because peaks in the 
spectrum indicate cyclic dynamics. Given a time series
or stochastic process, $w_i(t)$, measured in location $i$, the spectrum $S_{w_iw_i}(f)$ 
is a function of frequency, $f$. The frequency of a periodic oscillation is one over
its timescale, i.e., one over its period. The value of the spectrum $S_{w_iw_i}(f)$ is larger for 
frequencies on which $w_i(t)$ oscillates with greater strength. So, 
for instance, a population that exhibits strong (respectively, weak) 10-year-timescale 
oscillations will have a spectrum with a large (respectively, small) peak at 
the frequency $1/10 \text{yr}^{-1}$. The spectrum decomposes the variance, $\var(w_i(t))$, according 
to the timescales on which it occurs, i.e., an appropriate integral of 
$S_{w_iw_i}(f)$ across all frequencies equals $\var(w_i(t))$. The cospectrum
of two time series, $w_i(t)$ and $w_j(t)$, decomposes their covariance, $\cov(w_i(t),w_j(t))$,
in a similar way: the cospectrum, $C_{w_iw_j}(f)$, is a function of frequency
with peaks occurring at frequencies for which oscillations in $w_i(t)$ and
$w_j(t)$ are strong and coincident in their phases, i.e., synchronous; an 
integral of $C_{w_i w_j}(f)$ equals $\cov(w_i(t),w_j(t))$.

Our main theoretical result expresses the spectrum, $S_{\text{tot}}$, of a total
population across a region composed of $i=1,\ldots,N$ local areas in the form
\begin{equation}
S_{\text{tot}}(f) = F_{\text{popdyn}}(f) \left[ \sum_{i=1}^N S_{\epsilon_{i}\epsilon_{i}}(f) +
\sum_{\substack{i,j=1 \\ i \neq j}}^N C_{\epsilon_{i}\epsilon_{j}}(f)\right],\label{mainresmt}
\end{equation}
\noindent where $F_{\text{popdyn}}(f)$ reflects 
population dynamics within the local areas, and the $S_{\epsilon_i\epsilon_i}$
are spectra and the $C_{\epsilon_i\epsilon_j}$ are cospectra of environmental noise time 
series in the local areas. This result is for a model exhibiting 
autoregressive moving average (ARMA) dynamics 
within a habitat patch, and $F_{\text{popdyn}}(f)$
is a function that depends only on the ARMA coefficients (SI section \ref{SM-SI:theory}). 
If dynamics are not intrinsically cycling, 
then $F_{\text{popdyn}}(f)$ has no peaks and does not contribute to whether $S_{\text{tot}}$
is peaked. The term $\sum_{i=1}^N S_{\epsilon_{i}\epsilon_{i}}(f)$ in
(\ref{mainresmt}) encapsulates the effects of 
periodicities in local environmental noise time series on whether the total 
population exhibits cycling. The term 
$\sum_{i \neq j} C_{\epsilon_{i}\epsilon_{j}}(f)$ in (\ref{mainresmt}) encapsulates
the effects of timescale-specific synchrony on whether the total population
exhibits cycling. The key observation here is that, when $N$ is large so that the
region being considered is a combination of a large number of local areas, the first term
is likely to make a smaller contribution to $S_{\text{tot}}$ than the second
term; the first term is a sum of only $N$ quantities and the second is a sum of
$N^2-N$ quantities, and $N^2-N \gg N$ for large $N$. Thus theory predicts
that timescale-specific patterns of synchrony are very important for whether
spatial-total populations exhibit cycling. Of course, which of the two terms
in brackets in (\ref{mainresmt}) is larger depends not only on the number of summands
in each term, but also on their values. We explore 
reasonable scenarios in the SI
and find that the initial intuition is correct, 
that timescale-specific synchrony is often a very important determinant of
regional population cycles. See SI section \ref{SM-SI:theory}
for the theory and for examples, including SI Figs \ref{SM-TheoryExample1}, \ref{SM-TheoryExample2},
\ref{SM-TheoryExample3}.

# Methods

\textbf{Empirical Data.} Using data provided by the Wisconsin Department of Natural Resources 
(WIDNR), we calculated annual deer abundance estimates for the 
period `r minyear`-`r maxyear` for each of `r length(unique(dat$COUNTY_NAM))` WI counties; 
and an index of hunter abundance for each 
county for 1992-`r maxyear`. We obtained 
estimates of DVCs for each county for 1987-`r maxyear` 
from the WI Department of Transportation (WIDOT), as well as traffic 
volume estimates (millions of miles 
driven per county) for 1988-`r maxyear`, to control for traffic in our 
analysis of DVCs. We obtained daily weather station data from the 
National Climatic Data Center ([`https://www.ncdc.noaa.gov/`](https://www.ncdc.noaa.gov/))
and performed averaging to produce annual time series of winter ($\text{Dec}_{t-1}$ 
to $\text{Mar}_t$) weather by county for snow depth, 
precipitation amount, and minimum and maximum air temperature - here called Tmin and Tmax. 
We also calculated a Winter Severity Index (WSI), previously known to relate to winter 
impacts on deer populations [@widnr2001]. 
Winter and growing-season (Apr-Nov) annual time series for the NAO, the multivariate El Ni$\~{n}$o Southern Oscillation Index (MEI), and the Pacific 
Decadal Oscillation (PDO) were obtained by averaging monthly values.
All of these have been shown to 
influence large ungulate population dynamics in the north central 
USA [@Post1998_popgrowth;@Post2004_greenland]. 
High values of these indices generally correspond to mild winter weather over 
WI [@Rodionov2003_elnino].
Thus a total of 11 weather and climate variables were 
used: winter Tmin, Tmax, snow depth, and precipitation; winter NAO, MEI, and PDO; 
growing-season NAO, MEI and PDO; and the WSI. 
See Fig. \ref{Fig2_Timeseries} for plots of some data, and 
SI section \ref{SM-SI:methods} for details of all data and methods.

\textbf{Data pre-processing.} We performed all analyses described below 
at the three spatial scales mentioned in the *Introduction*: counties (N = 44-`r dim(cty.list$Abun)[1]` depending on 
the variables involved, see SI section \ref{SM-SI:methods}; mean county 
area $\pm$ SD = `r round(mean(gps$Area_km2),0)` $\pm$ `r round(sd(gps$Area_km2),0)` $\text{km}^2$), 
USDA districts ($N = 9$; mean 
district area $\pm$ SD = `r round(mean(usda.centroid$Area_km2),0)` $\pm$ `r round(sd(usda.centroid$Area_km2),0)` $\text{km}^2$), 
and statewide (SI Fig. \ref{SM-FigS1_map}). USDA district data were 
the sum or average of county-level variables, as districts were always conglomerates of 
counties (SI Fig. \ref{SM-FigS1_map}); likewise for statewide data. 
Results remained unchanged in their main substance using larger 
spatial scales (see *Results*).

Each time series was Box-Cox transformed to ensure Gaussian marginal 
distributions [@Sheppard2016_aphids], which are necessary for
the Fourier-surrogate-based spatial wavelet coherence 
testing procedures described below. This process also included linear detrending, 
to remove trends that might otherwise
obscure patterns of synchrony [@Buonaccorsi2001], as well as variance standardization 
(SI section \ref{SM-SI:methods}).

\textbf{Analysis.} We use wavelet transforms in all analyses.
Wavelets are ideal for detecting periodic components in complex field data
[@addison2002]. The methods we use are 
well established, standard tools in many fields, or are simple extensions thereof. 
So we provide in the main text an operational description, only,
of each method, i.e., what it takes as input, what it provides as output, and how to
interpret the output, with technical details in SI 
section \ref{SM-SI:methods}. 
<!--***DAN: this may end up getting moved elsewhere:
It is also explained in SI section \ref{SM-SI:methods}
why it is appropriate to use Fourier methods for the theory (above) and 
to additionally use wavelet methods
for the empirical analyses (below). -->
All wavelet analyses were done via the open-source `wsyn`
software package for the R language [@wsyn].

We characterized the spatial synchrony of spatiotemporal variables (e.g., deer
abundance) using wavelet mean field magnitudes and wavelet
phasor mean field magnitudes. Both methods take as input
standardized (see above) time series data from multiple locations and provide as
output plots which display the degree of synchrony among locations as a function
of time and timescale (Figs \ref{Fig1_example}G-I, SI 
Fig. \ref{SM-PedagogFigWMF}). Thus the plots reveal at what times, during the 
duration of data collection, and on what timescales, synchrony was prevalent. 
The wavelet phasor mean field magnitude plot describes the degree 
of phase synchrony among time series, whereas the wavelet mean field magnitude plot describes
the degree of synchrony in both phases and amplitudes of oscillation.
The statistical significance of a wavelet phasor mean field magnitude plot, at any given
time and timescale, can be assessed through 
comparison to a null hypothesis of random phases (SI 
section \ref{SM-SI:methods}; [@Sheppard2013; @Sheppard2016_aphids; @Sheppard_wavmodel]). 
We used a significance threshold  of $P < 0.001$. 

For both county- and district-level data, we assessed pairwise relationships
between spatiotemporal variables using the spatial wavelet coherence technique.
Spatial wavelet coherence tests for timescale-specific relationships between variables
(SI Fig. \ref{SM-PedagogFigSpatCoh}). Specifically, the technique tests whether two
spatiotemporal variables exhibit consistent phase differences and correlated
magnitudes of oscillation through time and across space, as a function of
timescale. Significance is determined by comparing spatial wavelet coherence values for
real data with spatial wavelet coherences made from appropriately randomized,
Fourier surrogate data sets ($N=$ `r nsurrogs`) that retain the same temporal and
spatial autocorrelation properties as the original time series but are otherwise
unrelated to each other (SI section \ref{SM-SI:methods}; [@Theiler1992;@Schreiber2000;@Sheppard2013;@Sheppard2016_aphids;@Sheppard2017_surrogates]). 
We calculated how often observed spatial wavelet coherence between pairs of variables was 
ranked higher than coherences of surrogate data sets, aggregating across the timescale 
band of interest using an
average-rank-based method (SI section \ref{SM-SI:methods}; @Sheppard2016_aphids). A $P$-value 
of less than $0.05$ means that the average empirical spatial wavelet coherence rank was higher than 
the same quantity calculated for at least 95% of the surrogate data sets
which represent the null hypothesis of no relationship. Ranks of
actual spatial wavelet coherences in surrogate spatial wavelet coherences were
also plotted against timescale. 
Finally, when two variables had significant spatial wavelet coherence over a timescale
band, we calculated the mean phase difference between the variables over the band,
$\bar{\theta}$, in units of $\pi$ radians, to determine if oscillations had
approximately in-phase ($|\bar{\theta}| = 0$ to $0.25$), quarter-cycle
($|\bar{\theta}| =  0.25$ to $0.75$), or anti-phase ($|\bar{\theta}| =  0.75$ to $1$)
relationships. We tested spatial wavelet coherence between deer abundance and weather 
and climate variables and hunter abundance, 
over the $3-7$-year timescale band, where the
majority of deer synchrony was present (see *Results*).
We also used spatial wavelet
coherence to test for relationships between deer abundance and DVCs, and between deer
abundance and DVCs adjusted for traffic volume.

We calculated percentages of deer synchrony that could be explained by the synchrony of individual weather 
variables, through Moran effects, as well the percentage of DVC synchrony explained by deer
synchrony, following the wavelet Moran theorem of @Sheppard2016_aphids.
As part of these procedures, we evaluated 
model "cross terms", which are the component of  synchrony that is due to variability 
explained by the drivers at a given location being synchronous with model residuals at other 
locations [@Sheppard2016_aphids]. Cross terms must be small ($<10\%$) 
in order for inferences about synchrony to be valid (SI section \ref{SM-SI:methods}). 
There are reasons to expect large cross terms to be mitigated at larger spatial scales
under some circumstance (SI section \ref{SM-SI:methods}), one reason why we performed analyses
at both county and district scales. 
The approach described here is most straightforwardly applied to predictors which 
are spatially as well as temporally
resolved, e.g., local weather variables as opposed to climate indices such as 
MEI. Attempts to use these techniques to quantify the fraction of deer synchrony explained by
MEI were more complex, and were not needed 
for our research goals, so are in SI sections \ref{SM-SI:methods} (methods) 
and \ref{SM-SI:results} (results).

To quantify the degree to which synchronizing effects are responsible for
cycling in statewide total deer and DVC time series, we adapted the standard
Fourier surrogate procedure of @Theiler1992. We constructed artificial
time series data, for both variables for each location, that had the same spectral
characteristics as the original data, except they were not spatially synchronized over
the $3-7$-year timescale band. We accomplished this by randomizing
phases of the Fourier transforms of the county-level time series 
(SI section \ref{SM-SI:methods}). We summed each set of
surrogate time series spatially to produce surrogate statewide total time series for both
deer and DVCs, representing the hypothetical case of no synchrony over $3-7$-year
timescales. We then compared these surrogate time series to the 
actual state-total deer abundance and DVC time series to determine to what extent 
additional variance and cycling intensity in the
empirical state-total time series were attributable to synchrony. 

Although wavelet tools were the most appropriate 
overall choice for our research goals and are now standard, some readers may be more familiar with Fourier analyses,
which have a longer history. Also, some readers may perceive a mismatch between our theory, which uses
the Fourier language of stochastic processes, and our empirical analyses based on wavelets (see SI section \ref{SM-SI:methods:methodsmismatch}).
A suite of alternative analyses based solely on Fourier approaches is presented in SI section \ref{SM-sect:alt_methods},
and the results of these alternative analyses are in SI section \ref{SM-sect:alt_results}.
<!--***DAN: You may get into trouble with citing an SI section out of order here, in which case you'll have to reorder the methods mismatch section, maybe.
It seems I only cite top-level sections of SI from the MT, so perhaps the best way to do this is to put the mismatch subsection first in the alt-methods
SI section, and then just cite the top-level alt-methods section here. You are also going to have to cite all the SI figures and maybe also tables
from the main text somewhere (just do summary citations of ranges of figures as appropriate).-->
All main conclusions were dually supported by both wavelet and Fourier approaches.

\textbf{Data availability.} All data will be published in the Dryad repository. 
<!--***DAN: Tom to mention in the cover letter that data will be so deposited when the paper
is accepted, and Tom also to replace XYZ with the repo number when the time comes, please.-->
All data and code producing this paper are also archived at `www.github.com/reumandc/TomAnderson_WIdeer`.
Analyses can be re-run and the paper re-compiled via the codes there.

# Results

The wavelet mean field of county deer abundance time series revealed that synchrony occurred 
chiefly at $3-7$-year timescales (Fig. \ref{Fig3_DeerSync}A). The wavelet 
phasor mean field showed the strong statistical significance of this phenomenon (Fig.
\ref{Fig3_DeerSync}B). We therefore focused subsequent analysis on the $3-7$-year timescale band.
Mean field magnitude plots for all
winter weather variables are in SI Fig. \ref{SM-FigS8_WMFsAndWPMFsAncillary}.

Spatial wavelet coherence demonstrated that 
deer abundance synchrony on $3-7$-year timescales was partly caused by climatic Moran effects. 
Deer synchrony was related to winter snow depth 
(spatial wavelet coherence, $P =$ `r round(Tab1_results$Pvalue[Tab1_results$Predictor=="Snow Depth" &Tab1_results$Timescale=="3-7"],2)`, 
Table \ref{Tab1_CountyCoherenceSummary}), in a near anti-phase relationship (mean phase 
difference between abundance and snow depth across $3-7$-year timescales, in units of ${\pi}$ 
radians, was $\bar{\theta} =$ `r round(Tab1_results$MeanPhase[Tab1_results$Predictor=="Snow Depth"&Tab1_results$Timescale=="3-7"],2)`; SI Fig. \ref{SM-FigS4_PhaseDiffs}D). Thus deer 
abundances tended to peak in years of below-average snow. No other local weather variables were 
significantly related with deer abundance (Table \ref{Tab1_CountyCoherenceSummary}). 
Deer abundance was also related, with
approximately quarter-cycle phase shifts, 
to the multivariate El Ni$\~{n}$o Southern Oscillation index
($P =$ `r round(Tab1_results$Pvalue[Tab1_results$Predictor=="Winter MEI"&Tab1_results$Timescale=="3-7"],2)`, $\bar{\theta}  =$ `r round(Tab1_results$MeanPhase[Tab1_results$Predictor=="Winter MEI"&Tab1_results$Timescale=="3-7"],2)`) and to the Pacific Decadal Oscillation 
($P =$ `r round(Tab1_results$Pvalue[Tab1_results$Predictor=="Winter PDO"&Tab1_results$Timescale=="3-7"],2)`, $\bar{\theta}  =$ `r round(Tab1_results$MeanPhase[Tab1_results$Predictor=="Winter PDO"&Tab1_results$Timescale=="3-7"],2)`, 
Table \ref{Tab1_CountyCoherenceSummary}, SI Fig. \ref{SM-FigS4_PhaseDiffs}A,C), both averaged over the winter months
($\text{Dec}_{t-1}$ to $\text{Mar}_t$, abbreviations $\text{MEI}_{\text{w}}$ and
$\text{PDO}_{\text{w}}$). These two indices were strongly coherent and in-phase with
each other on $3-7$-year timescales over 1981-2016 (wavelet coherence, $P =$ `r round(TabS1_results$Pvalue[TabS1_results$Predictor=="Winter MEI" & TabS1_results$Response=="Winter PDO" ],4)`, $\bar{\theta}  =$ `r round(TabS1_results$MeanPhase[TabS1_results$Predictor=="Winter MEI" & TabS1_results$Response=="Winter PDO" ],2)`, SI Table \ref{SM-TabS1_PairwiseClimVarSpatCoh}). Deer abundance was also significantly coherent with growing-season MEI (Table \ref{Tab1_CountyCoherenceSummary}), but likely only because $\text{MEI}_{\text{w}}$ and growing-season MEI were coherent (SI Table \ref{SM-TabS1_PairwiseClimVarSpatCoh}). 
At $5$-yr periodicity, the quarter-cycle phase lags between deer abundance and $\text{MEI}_{\text{w}}$/$\text{PDO}_{\text{w}}$
correspond to deer populations
peaking about `r round(-5*Tab1_results$MeanPhase[Tab1_results$Predictor=="Winter PDO"&Tab1_results$Timescale=="3-7"]/2,1)` or 
`r round(-5*Tab1_results$MeanPhase[Tab1_results$Predictor=="Winter MEI"&Tab1_results$Timescale=="3-7"]/2,1)` years after mild winters. 
Significant spatial wavelet coherence with an environmental variable which is itself spatially synchronous 
(e.g., snow depth or some combination of weather variables related to MEI and PDO) is evidence of transmission 
of synchrony from the environment to deer via Moran effects (SI section \ref{SM-SI:methods}; @Sheppard2016_aphids, @Sheppard_wavmodel). 

Snow depth and climate index results seemed to 
indicate distinct or partly distinct phenomena 
synchronizing deer on different timescales within the $3-7$-year band, and examining these distinct phenomena
provides additional evidence that Moran effects on deer are occurring.
Deer dynamics at $3-4$-year timescales were not significantly spatially coherent with either 
$\text{MEI}_{\text{w}}$ or $\text{PDO}_{\text{w}}$, which instead were strongly coherent with
deer on $4-7$-year timescales (Table \ref{Tab1_CountyCoherenceSummary}, see also 
Fig. \ref{SM-FigS5_RanksInSurrogs}A and Table \ref{SM-TabS2_SummaryDistrict}).
In contrast, spatial wavelet coherence between snow depth and deer was 
not significant on $4-7$-year timescales, but was significant on $3-4$-year timescales
(Table \ref{Tab1_CountyCoherenceSummary}). 
Applying the wavelet Moran theorem, snow depth explained $`r round(snowabunSyncExp3_7,0)`\%$ of 
synchrony in deer abundance at $3-7$-year timescales, but $`r round(snowabunSyncExp3_4,0)`\%$ of synchrony in deer
abundance at $3-4$-year timescales (Table \ref{Tab1_CountyCoherenceSummary}; see also
USDA district level results for the $3-4$-year band, Table \ref{SM-TabS2_SummaryDistrict},
which were similar but had smaller cross terms). The timescale
bands $3-4$ and $4-7$ were chosen by examination of Fig. \ref{SM-FigS5_RanksInSurrogs}.
The wavelet Moran theorem of @Sheppard2016_aphids could not straightforwardly be
applied to estimate percentages of synchrony in deer explained by $\text{MEI}_{\text{w}}$,
because $\text{MEI}_{\text{w}}$ is not a spatially resolved variable (*Methods*).
Such estimates are not necessary, however,
for our argument that deer are environmentally synchronized in the $3-7$-year
band since this is already adequately demonstrated by our wavelet coherence results.
See SI section \ref{SM-SI:results} (also Fig. \ref{SM-InflationFig} and Table \ref{SM-InflationTable}) 
for further discussion.

Deer synchrony on $3-7$-year timescales was not caused by hunting. County-level hunter and deer abundances were not significantly spatially coherent 
($P =$ `r round(Tab1_results$Pvalue[Tab1_results$Predictor=="Hunters" & Tab1_results$Timescale=="3-7"],2)`) over $3-7$-year timescales 
(Table \ref{Tab1_CountyCoherenceSummary}, SI Figs \ref{SM-FigS5_RanksInSurrogs}B, \ref{SM-FigS6_WMFHunter}). Hunter abundance was significantly related to deer on shorter timescales, around 
$2-2.5$-years ($P =$ `r round(Tab1_results$Pvalue[Tab1_results$Predictor=="Hunters" & Tab1_results$Timescale=="2-2.5"],3)`), and was in approximate anti-phase with deer abundance 
on those timescales ($\bar{\theta}  =$ `r round(Tab1_results$MeanPhase[Tab1_results$Predictor=="Hunters" & Tab1_results$Timescale=="2-2.5"],2)`; 
Table \ref{Tab1_CountyCoherenceSummary}, SI Figs \ref{SM-FigS4_PhaseDiffs}G,
\ref{SM-FigS5_RanksInSurrogs}B). 
But these short-timescale results, while interesting (*Discussion*), are not particularly
relevant for synchrony because deer showed limited synchrony on 
those timescales (Fig. \ref{Fig3_DeerSync}).

The wavelet and wavelet phasor mean field magnitude plots for DVCs revealed time-and timescale-specific synchrony, and visual patterns were remarkably similar to those shown by deer (compare Figs \ref{Fig3_DeerSync}A,B and \ref{Fig4_DvcSync}A,B). Spatial wavelet coherence showed that deer abundance and DVCs were significantly related 
($P =$ `r round(Tab1_results$Pvalue[Tab1_results$Response=="DVCs" & Tab1_results$Predictor=="Abundance"],3)`, 
Table \ref{Tab1_CountyCoherenceSummary}) on $3-7$-year timescales, and had a nearly in-phase 
relationship ($\bar{\theta} =$ `r round(Tab1_results$MeanPhase[Tab1_results$Response=="DVCs" & Tab1_results$Predictor=="Abundance"],2)`, SI Fig. \ref{SM-FigS4_PhaseDiffs}E). 
Deer abundance explained $`r round(usda_dvcabunSyncExp,0)`\%$ of the synchrony in 
DVCs when data were aggregated to USDA districts (SI section \ref{SM-SI:results}, 
Fig. \ref{SM-FigS7_WMFAndPredictionsDistricts} and Table \ref{SM-TabS2_SummaryDistrict}). 
As a visual manifestation of this statistical result, a "predicted synchrony" plot, which 
shows patterns of synchrony in DVCs 
which would manifest if deer were the sole cause of DVC synchrony (see SI section \ref{SM-SI:methods} for
how such a plot is produced), closely matched the patterns actually present in DVC
synchrony (Fig. \ref{Fig4_DvcSync}C). 
Statistical significance, phase relationships, and the amount of DVC synchrony explained by deer
abundance remained similar after adjusting DVCs to account for
fluctuations in traffic volume (Table \ref{Tab1_CountyCoherenceSummary}, SI section \ref{SM-SI:results} and Table \ref{SM-TabS2_SummaryDistrict}, Fig. \ref{SM-FigS4_PhaseDiffs}F).

Statewide oscillations due to synchrony correspond to a typical difference of about $`r round(avgdeerfluc,-4)`$ deer and about $`r round(avgdvcfluc,-2)`$ 
DVCs between peak and trough years 
(Fig. \ref{Fig5_StateWideFluctuations}). Importantly, these high amplitude 
statewide oscillations in deer and DVCs disappear (Fig. 
\ref{Fig5_StateWideFluctuations}) when synchrony between counties on $3-7$-year 
timescales is removed via a randomization procedure (*Methods*). Therefore, it is spatial synchrony 
that causes local dynamics to sum to a substantial 
semi-periodic oscillatory phenomenon on statewide spatial scales. Thus, the essence of our theoretical results was borne out in our empirical results (Figure \ref{SM-PedagogEmp_Fig}). Depending on 
assumptions made about DVC costs and reporting rates (SI section \ref{SM-SI:methods};  @Huijser2008_dvc; @Marcoux2010_dvc; @Conover2011_bookchp), 
statewide DVC damages 
vary by US\$$`r floor(round(2*avgdvccostsLow,-6))/1000000`-`r floor(round(2*avgdvccostsHigh,-6))/1000000`$Mn 
between peak and trough years. 

Supplementary results using Fourier approaches (*Methods*, SI section \ref{SM-sect:alt_methods})
agreed with wavelet results, confirming the same main conclusions that: 
1) synchrony in deer and DVCs was timescale specific and had a peak at $3-7$-year timescales;
2) deer synchrony was due to environmental Moran effects and DVC synchrony was due to the effects of deer; and
3) statewide cycles in deer and DVCs occurred, and can be attributed to the timescale-specific synchrony
in these variables (SI section \ref{SM-sect:alt_results}).
<!--***DAN: also cite all the spectral figs and tables here, all in one batch, to maintain the custom of citing all
sup mat figs and tables from the main text and having them appear in the sup mat in the order they are cited in the
main text-->

# Discussion

<!--summary-->
Overall, we demonstrated that timescale-specific Moran effects can cause cycles in a 
regional population. This is a new mechanism by which the environment can 
produce population cycles. 
Previous attemps to link environmental factors to population cycles have either been unsubstantiated [e.g., 
sun spots; @Sinclair1992_sunspots;@Selas2004_sunspots], or provided a useful but partial explanation for cycles, 
as in studies of tetraonids [@watson2000] or small mammals [@kausrud2008] which demonstrated that environmental 
factors tend to modify cycles produced by density dependent factors, rather than producing them outright.
Our study differs from previous work, in part because the new mechanism does not require a regular environmental 
oscillator like sun spots. Rather, it is sufficient for irregular environmental drivers 
such as winter climate to be coherent with populations over a particular timescale band, thereby inducing 
synchrony and leading to cycles in that band.
We also presented what is, to our knowledge, the first demonstration that synchrony 
can be transmitted in a 
Moran-like effect from ecological to human systems (deer to DVCs).
If climate change alters synchrony, as a growing number
of papers indicate is likely occurring 
[@Post2004_greenland;@Allstad2015_gypsysync;@Defriez2016_plankton;
@Sheppard2016_aphids;@Koenig2016_birdsync;@Shestakova2016_forestsync;@Black2018;@hansen2020], 
our results show that cycles
in populations and human systems may change as a consequence.

<!--Variants of the mechanism. -->
There are two alternative elaborations of our mechanism.
First, environmental drivers may have produced
small oscillations on $3-7$-year timescales in county-level deer populations, and 
also brought these oscillations into synchrony, thereby causing them to sum across
counties to produce statewide cycles. Second, the local population oscillations 
may have been due partly to the intrinsic dynamics of deer, resulting 
from density-dependent effects, maternal effects, or 
other mechanisms of nonlinear dynamics. Synchronous environmental drivers 
would then have brought the intrinsically oscillating local dynamics into synchrony. 
The difference between these alternatives is whether local population
oscillations were produced and synchronized, or synchronized but 
not produced, by environmental drivers. The alternatives would be difficult to distinguish, but the former mechanism
may be more likely because winter climate is known to influence ungulate
dynamics [@Mysterud2008_deerclimate; @Hegel2012_caribou; @Post2004_greenland; @Grotan2005_roedeer],
and we are unaware of any evidence suggesting that deer populations are intrinsically
cyclic. In either case our mechanism is the
same whereby Moran effects synchronize local dynamics in a timescale-specific manner,
and then local dynamics sum to produce cycles. Randomizations that removed synchrony 
but preserved local fluctuations eliminated the state-wide 
cycles (Fig. \ref{Fig5_StateWideFluctuations}), demonstrating that synchrony likely produced them.

<!--Reasonableness of the drivers we found in relation to other studies and biological
mechanisms-->
Large-scale climatic patterns and local winter weather are known to influence synchrony 
via Moran effects in other ungulates, including reindeer, 
caribou [@post2002;@aanes2003;@Post2004_greenland;@Hegel2012_caribou], roe deer [@Grotan2005_roedeer], 
and musk oxen [@post2002], and thus the cycling mechanism we observed may also occur 
in other ungulates. Here, winter snow depth and $\text{MEI}_{\text{w}}$/$\text{PDO}_{\text{w}}$ 
were the primary determinants of deer synchrony. Deer populations lagged $\text{MEI}_{\text{w}}$ 
by 1-2 years, a similar delay to other reported lag times of climatic effects on deer 
populations [@Post1998_popgrowth;@forchhammer1998]. Winter weather and climate probably influenced deer
because deeper snow limits access to winter food, which then increases 
susceptibility to starvation or predators and reduces 
parturition rates [@Mysterud2008_deerclimate;@Hegel2012_caribou]. 
Winter weather influences deer beyond WI, so timescale-specific synchrony may extend further than WI. Cycling 
over larger regions may then be even stronger than what we documented because
a larger number of synchronized local populations can accentuate the effect we described.

<!--It aint hunters. -->
Hunters were not the principal driver of deer synchrony. These results are consistent with the
hypothesis that hunters can regulate deer populations over short timescales and small 
areas [@fryxell1991;@denicola2008] but have limited effects over longer timescales and 
larger areas [@brown2000;@simard2013]. The lack of longer-timescale ($3-7$-year) spatial wavelet 
coherence between hunters and deer abundance likely also relates to the facts that hunters 
typically prefer to shoot bucks, the removal of
which has limited population effects; and the state regulates antlerless harvest to
maintain desirable population sizes.

<!--applied reasons to care, link to climate change-->
Our results provide a previously unrecognized applied motivation for studies 
of synchrony in populations.
Fluctuations of a quarter million deer, statewide, caused by synchrony, 
likely will have considerable consequences for agricultural crop damage, forest regeneration 
[@Cote2004_deerimpacts], and hunting success, all of which can have socioeconomic impacts. 
We also showed that Moran effects subsequently pass from deer to DVCs in a cascading manner. 
Deer synchrony almost certainly caused DVC synchrony rather than the opposite causation,
because DVCs were too infrequent to
meaningfully drive deer fluctuations (Fig. \ref{Fig2_Timeseries}), and if they did an anti-phase
relationship would have been found (as opposed to the in-phase relationship we observed).
Thus Moran effects are ultimately responsible for synchrony of DVCs, and synchrony 
causes fluctuations of more than 2500 annual documented DVCs through time.
Such fluctuations may influence insurance and accident recovery industries, though we have not inquired 
whether a fluctuation of this magnitude is big enough to be of substantial concern to these industries.
Our results provide another reason that
additional studies of climate-change influences on synchrony [@hansen2020] should be undertaken: 
such influences may alter the cycling of populations, including populations of socioeconomic importance. 

<!--specificity to timescales, and the strengths of our methods-->
An important aspect in the phenomena we explore is their specificity to 
certain timescales. Deer cycles would not have manifested had deer synchrony 
not been specific to certain timescales.
Timescale specificity in synchrony is now know to be common 
[@Sheppard2016_aphids;@Sheppard2017_surrogates;@Walter2017_geogsync;@Sheppard_wavmodel;@Haynes_etal2019;@Anderson2019_LO].
Many past analyses of synchrony have 
used correlation methods [@Leibhold2004_sync] that conflate timescales, but this can hinder 
detection of synchronous phenomena and inferences of drivers of synchrony 
by obscuring synchrony on some timescales with 
asynchronous dynamics on other timescales [@Sheppard2016_aphids].
The methods used here [@Sheppard2016_aphids;@Sheppard2017_surrogates;@Walter2017_geogsync;@Sheppard_wavmodel]
were able to isolate the timescales on which synchrony occurred,
facilitating precise determination of drivers of synchrony.
Historically, determination of causes of synchrony was called 
difficult [@Leibhold2004_sync;@Kendall2000_moran;@Abbott2007_moran], but in our previous
work [@Sheppard2016_aphids;@Sheppard2017_surrogates;@Walter2017_geogsync;@Sheppard_wavmodel],
and again here, we showed that using timescale-specific methods can help 
ameliorate this difficulty and can reveal new phenomena and mechanisms.

# Acknowledgements
D. Lyden, J. Steinglein and the WIDNR for data; D. Storm, T. Van Deelen, J. Millspaugh, J. Fox, K. Haynes, and A. Liebhold for consultation; and anonymous referees for very useful feedback and some code. Funding was provided by the James S McDonnell Foundation, United States National Science Foundation grants 1442595 and 1714195, United States Department of Agriculture award 2016-67012-24694, and The Nature
Conservancy. 

<!--Load sup mat workspace for the methods, which used to be in the sup mat-->
```{r echo=F, results="hide"}
load(file="WholeWorkspace.RData")
```

# References {-}

<div id="refs"></div>

\newpage


```{r Table1,echo=F,results='markup'}
options(knitr.kable.NA = '') #makes it put blanks instead of "NA" for NAs
rownames(Tab1_MT)<-NULL
knitr::kable(Tab1_MT[,c(1:7)],
             format="latex", align="l",escape=F,
             digits=4,
             col.names=c("Predictor","Response", "Timescale","P-value","Mean phase ($\\bar{\\theta}$)","Synchrony explained","Average cross terms"),
             caption = "Summary of spatial wavelet coherence tests and related results using county-level data. $P$-values and mean phases are from tests of spatial wavelet coherence (\\textit{Methods}) of the variables in columns 1 and 2 over the indicated timescale band. Mean phases over the timescale band (units of $\\pi$ radians) are provided only when meaningful, i.e., when spatial wavelet coherence was significant, $P < 0.05$, or marginally significant. A positive mean phase indicates the variable in the column \\textit{Response} is leading, and a negative mean phase indicates the variable in the column \\textit{Predictor} is leading. The \\textit{Synchrony explained} and \\textit{Average cross terms}  columns, which give percentages, are based on the wavelet Moran theorem (\\textit{Methods}). When cross terms are low, synchrony explained is interpretable as the approximate percentage of synchrony in the response explainable as due to the synchrony in the predictor. Cross terms were low enough except for tests involving DVCs, for which USDA district-level results were more reliable (SI Table \\ref{SM-TabS2_SummaryDistrict}). \\textit{Synchrony explained} and \\textit{Average cross terms} are not reported for climate-index predictors because the climate indices are not spatially resolved (\\textit{Methods}). All values were rounded to display four digits. Abundance = deer abundance. \\label{Tab1_CountyCoherenceSummary}",
             booktabs=T)
```

\newpage

\begin{figure}[!h]  
\includegraphics[width=.5\textwidth]{Results/Fig1.png}
\caption{Simulations demonstrating our mechanism whereby
timescale-specific synchrony can cause population cycling. Three scenarios with different patterns of
environmental synchrony result in different outcomes: scenario 1, an asynchronous-environment scenario, results in no 
population synchrony or cycling; scenario 2 has environmental synchrony on a 3-year timescale, resulting in
3-year population synchrony and 3-year regional-scale cycles; scenario 3 has environmental synchrony on a 10-year timescale, resulting in
10-year population synchrony and 10-year cycles. In scenario 1, 
environmental noise in location $i$ ($i=1,\dots,71$; first $15$ locations
shown in A) was modeled as 
$e^{(1)}_i(t)=\sin(2 \pi t/w^{(1)}+2 \pi u^{(1)}_i)+3\epsilon^{(1)}_i(t)$, where $w^{(1)}=3$, 
the $u^{(1)}_i$ were independent
draws from the uniform distribution on $(0,1)$, and the local noises $\epsilon^{(1)}_i(t)$ were
standard normal and independent across locations and times. In scenario 2, 
noise ($15$ locations shown in B) 
was $e^{(2)}_i(t)=\sin(2 \pi t/w^{(2)}+2 \pi u^{(2)})+3\epsilon^{(2)}_i(t)$
where $w^{(2)}=3$, $u^{(2)}$ was one single standard-normally distributed random number used for all
locations, and the $\epsilon^{(2)}_i(t)$ were again
standard normal and independent. Noise for scenario 3 (C), 
was like scenario 2 but used $w^{(3)}=10$. Thus the noises in 
scenarios 2 and 3 had weak and not visually obvious timescale-specific synchrony and those in 1 had no synchrony.
Populations were modelled as $p^{(s)}_{i}(t+1)=\rho p^{(s)}_i (t)+e^{(s)}_i(t)$ using $\rho=0.4$
for scenarios $s=1,2,3$, and are displayed (D-F) for locations 1-15.
Although population synchrony was weak and 
cannot readily be detected visually, wavelet mean field magnitide plots (\emph{Methods})
revealed that synchrony was present, in scenarios 2 and 3, on 3- and 10-year
timescales (G-I); on such plots, higher values mean greater synchrony at the given 
time on the given timescale. Total populations (J-L) show clear cycling in scenarios 2 and 3,
of periods $3$ and $10$, respectively, consistent with the proposed mechanism 
from the \emph{Introduction}: synchrony on a given timescale 
produces cycling on that timescale in the total population.
We used $71$ locations because the empirical analyses of this study focus on white-tailed
deer populations in Wisconsin, and Wisconsin has 71 counties for which deer 
data were available. Time series on A-F were
shifted vertically for readability.
\label{Fig1_example}}
\end{figure}

\newpage

\begin{figure}[h!]
\includegraphics[width=\textwidth]{Results/Fig2.png}
\caption{Raw time series of (A) deer abundance, (B) deer-vehicle collisions (DVCs), 
(C) average winter ($\text{Dec}_{t-1}$ to $\text{Mar}_t$) snow depth, and 
(D) winter climate indices (Multivariate El Ni$\~{n}$o Southern Oscillation Index, 
MEI; Pacific Decadal Oscillation, PDO), and hunter abundances (E). 
Each line in A-C, E represents a county. Other variables were 
used in analyses (see text) but are not shown here due to lack of significant 
assocations with the deer and DVC variables that were the focus of the 
study.\label{Fig2_Timeseries}}
\end{figure}

\newpage

\begin{figure}[!h]
\includegraphics[height=.7\textheight]{Results/Fig3.png}
\caption{Wavelet mean field (A) and wavelet phasor mean field (B) magnitudes
of deer abundance time series (\textit{Methods}). Higher values in A, B indicate greater synchrony (A) or phase synchrony (B) 
at the indicated times and timescales. Contours on (B) indicate statistical significance of phase synchrony at 
$P =$ 0.001. Horizontal lines in (A) and (B) indicate the timescales that were the focus of subsequent statistical 
analysis, $3-7$-years. 
\label{Fig3_DeerSync}}
\end{figure}

\newpage

\begin{figure}[!h]
\includegraphics[height=.7\textheight]{Results/Fig4.png}
\caption{Wavelet mean field (A) and wavelet phasor mean field (B) magnitudes, and predicted synchrony (C) 
of DVCs. Colors, contours, and horizontal lines in A and B indicate the same as in 
Fig. \ref{Fig3_DeerSync}A,B, but for DVC data. Colors in (C) 
show synchrony in DVCs predicted by a wavelet model with sole predictor deer abundance, while contours in (C) 
show the actual wavelet mean field magnitude of DVCs from (A), for comparison. Agreement is good, giving a 
visual indication of the statistical result that deer synchrony explains DVC synchrony.
\label{Fig4_DvcSync}}
\end{figure}

\newpage

\begin{figure}[!h]
\includegraphics[width=.6\textwidth]{Results/Fig5.png}
\caption{Magnitude of the effects of synchrony on state-total deer abundance
and DVC fluctuations. Small panels show state-total deer abundance (A, in millions) and DVC (C, in thousands) time 
series (black lines) and surrogate totals obtained via a randomization procedure 
(\textit{Methods}) that eliminates synchrony among counties at $3-7$-year timescales 
(grey lines, 1000 surrogates plotted). Increased variation and semi-periodic 
oscillations due to synchrony are particularly notable when trends and long-timescale 
(longer than 7 years) variability are filtered out by subtracting the mean of the 
grey lines (B for deer, in thousands; D for DVCs). $\Delta$ represents the difference in values from 
the surrogate mean. Red arrows highlight examples of the strong cycling that is 
attributable to synchrony.
\label{Fig5_StateWideFluctuations}}
\end{figure}

 <!--Deleted from Fig. 5 caption. For DVCs, the typical fluctuation in numbers of DVCs due 
to synchrony amounts to an estimated difference between peak and trough years of 
US\$$`r floor(round(2*avgdvccostsLow,-6))/1000000`-`r floor(round(2*avgdvccostsHigh,-6))/1000000`$Mn in damages 
statewide (\textit{Methods}).-->
