---
title: "Supplementary Materials for Moran Effects Cause Major Cycles in Deer Populations and Deer-Vehicle Collisions Across Wisconsin"
author: "Thomas L. Anderson, Lawrence W. Sheppard, Jonathan A. Walter, Robert E. Rolley, Daniel C. Reuman"
date: "correspondence to: anderstl@gmail.com, reuman@ku.edu"
geometry: "left=1cm,right=1cm,top=2.5cm,bottom=2.8cm"

output: 
  pdf_document:
    keep_tex: yes
    fig_caption: yes
tables: True
link-citations: True
urlcolor : blue

csl: science.csl
bibliography: refs.bib
---

\textbf{This PDF file includes:}

* Materials and Methods
* Supplementary Text
* Figs. S1 to S8
* Tables S1 to S5

<!--***DAN: Tom, in order for me to put in demos of how some things work here are some 
items I need from you first:
1) the name of an R variable in your code for some in-text number, and an indication of which number, e.g., your response might look like "the p-value in the first line under Table S1 is stored in the R variable  SnowDepthAbundanceCohP35".
2) the name of an R data frame that contains Table S1, e.g., you might just call it TableS1df. 
3) Please open Endnote and try to figure out how to expport your library of ciations that you are using
for this paper as a bibtex file (.bib extension). This reference may help: https://www.reed.edu/cis/help/LaTeX/EndNote.html. Then commit and push the file to the repo in the same folder as this Rmd and let me know.-->

<!--***DAN: Currently we are using science citation style in the supp mat, is this 
required, Tom? If not, wwe should use the callout style we prefer. What style do you
prefer, Tom?-->

<!--checkpoint package for CRAN-available packages-->
```{r checkpoint_chunk, echo=F, results="hide",warning=F, message=F, results="hide"}
if (!dir.exists("./.checkpoint/")){
  dir.create("./.checkpoint/")
}
library(checkpoint,quietly=T)
checkpoint("2018-04-20",checkpointLocation = "./")
```

<!--Install the correct reumannplatz version from github, i.e., the one current on 2018 04 20, tagged
in git with AndersonDeer-->
```{r rp_install, echo=F, message=FALSE}
library(devtools,quietly=T,verbose=F)
library(withr,quietly=T,verbose=F)
withr::with_libpaths(new=.libPaths()[1],
                     devtools::install_github(repo="reumandc/reumannplatz",ref="AndersonDeer",force=T))
```

<!--Libraries-->
```{r libraries, echo=F, message=FALSE}
library(reshape2,quietly=T)
library(tidyr,quietly=T)
library(plyr,quietly=T)
library(tools,quietly=T)
library(fields,quietly=T)
library(shape,quietly=T)
library(Reumannplatz,lib.loc=.libPaths(),quietly=T)
```

<!--Flags controlling the run and other setup-->
```{r setup, echo=F}
#which density to use ("dnr" for original estimates, "tom" for your own calculations)
dens.flag<-"dnr"

#what is the earliest year you want? (hunting only: 1960; abundance: 1981; DVC: 1987; Traffic: 1988, Hunters: 1992)
minyear<-1981

#what is the last year you want? (max is 2016)
maxyear<-2016
  
#normalize the data ("norm" for Box-Cox, "detr" for linearly detrend, or "none")
##Code currently work using the "none" option; the normalization steps happen 
##individually with each analysis step (e.g. see line 15 in "DeerAbundance.R" file), 
##because of the conflicting data lengths and dimensions. 
detr.flag<-"none"

#set spatial scale flag for data generation ("usda", "county" or "both")
scale.flag<-"both"

#***DAN: Not sure what this is, Tom, if this is not necessary should be deleted? 
#calculate how many missing climate values there were (don't need to run anymore)
samplesize.flag<-"n"

#set number of surrogates to use for spatial coherence
nsurrogs<-10000

#a function needed for caching
source("Code/mtime.R") 
```

<!--***DAN: Run time is sufficiently long that we should probably use caching, but
that will require a good understanding of dependencies of everything. I have begun.
Other reasons to use cacheing include it will enforce a clearer specification of
dependencies between chunks.-->

<!--Clean the population data-->
```{r clean_pop_data, echo=F, cache=T, results="hide", fig.show="hide", cache.extra=list(dens.flag,minyear,maxyear,detr.flag,mtime("Code/wi.deer.cleandata.R"),mtime("Data/State_Kill.csv"),mtime("Data/County_KillTotal.csv"),mtime("Data/County_Kill_BuckOnly.csv"),mtime("Data/County_Kill_DoeOnly.csv"),mtime("Data/County_KillGunBuck.csv"),mtime("Data/County_KillGunDoe.csv"),mtime("Data/County_KillBowBuck.csv"),mtime("Data/County_KillBowDoe.csv"),mtime("Data/WI_Cty_CentroidDD.csv"),mtime("Data/DOT_Accidents1987_2016.csv"),mtime("Data/Adj_DOT_Accidents1987_2016.csv"),mtime("Data/OpeningDay.csv"),mtime("Data/CornHarvest.csv"),mtime("Data/WI_USDA_Sections.csv"),mtime("Data/USDA_centroids.csv"),mtime("Data/Cty_Traffic.csv"),mtime("Data/DMU_DeerAbundance1981_2013.csv"),mtime("Data/DMU_DeerDensity1981_2013.csv"),mtime("Data/DMU_HunterAbundance1975_2013.csv"),mtime("Data/DMU_HunterDensity1975_2013.csv"),mtime("Data/hunterdens1416.csv"),mtime("Data/hunterabun1416.csv"),mtime("Data/Cty_Abun_2014_2016.csv"),mtime("Data/permit_issued1970_2013.csv"))}
#Clean population data: end result are two lists of matrices called "cty.list" and "usda.list"
system.time(source("Code/wi.deer.cleandata.R"))
  #list of files generated by wi.deer.cleandata.R:
    #scale.flag == "usda": usda.list, includes data on abundance, kill data, crops, DVCs and traffic
    #scale.flag == "county": cty.list, includes data on abundance, kill data, crops, DVCs and traffic
```

<!--Clean the climate data-->
```{r clean_clim_data, cache=T, echo=F, results="hide", fig.show="hide", cache.extra=list(minyear,maxyear,detr.flag,scale.flag,mtime("Code/wi_climate.R"),mtime("Data/noaa_latlongUpd.csv"),mtime("Data/WI_USDA_Sections.csv"),mtime("Data/Cty_NOAA_Temp1981_2015_set1.csv"),mtime("Data/Cty_NOAA_Temp1981_2015_set2.csv"),mtime("Data/Cty_NOAA_Temp1981_2015_set3.csv"),mtime("Data/Cty_NOAA_Temp1981_2015_set4.csv"),mtime("Data/Cty_NOAA_Temp1981_2015_set5.csv"),mtime("Data/Cty_NOAA_Temp1981_2015_set6.csv"),mtime("Data/Cty_NOAA_Temp2016_2017.csv"),mtime("Data/Cty_NOAA_Precip1981_1985.csv"),mtime("Data/Cty_NOAA_Precip1986_1991.csv"),mtime("Data/Cty_NOAA_Precip1992_1997.csv"),mtime("Data/Cty_NOAA_Precip1998_2003.csv"),mtime("Data/Cty_NOAA_Precip2004_2009.csv"),mtime("Data/Cty_NOAA_Precip2010_2015.csv"),mtime("Data/Cty_NOAA_Precip2016_2017.csv"),mtime("Data/climate_indices.csv"),mtime("Data/mei.csv"),mtime("Data/pdo.csv"),mtime("Data/npi.csv"),mtime("Data/ENSO.csv"))}
system.time(source("Code/wi_climate.R"))
#list of files generated by wi_climate.R:
  #scale.flag == "usda": wsi.usda.mat,hunt.clim.usda, winter.clim.usda,nao.mat,pdo.mat,mei.mat,winter.nao,winter.pdo,winter.mei
  #scale.flag == "county": wsi.mat,wsi.matN,hunt.clim,winter.clim,nao.mat,pdo.mat,mei.mat,winter.nao,winter.pdo,winter.mei
  #large climate indices are the same values (nao,mei,pdo), just replicated into different dimensions to match scale.flag
```

<!--County analysis-->
<!--***DAN: DANGER this list of dependencies not complete, need to fill in!-->
```{r county_analysis, echo=F, results="hide", fig.show="hide", cache=T, cache.extra=list(mtime("Code/DeerAbundance.R"))}
#Run analysis on deer abundance and DVCs using county data
set.seed(101)
system.time(source("Code/DeerAbundance.R"))
```

<!--District analysis-->
<!--***DAN: DANGER this list of dependencies not complete, need to fill in!-->
```{r district_analysis, echo=F, results="hide", fig.show="hide", cache=T, cache.extra=list(mtime("Code/DeerAbundanceUSDA.R"))}
#Run analysis on deer abundance and DVCs using USDA district data
set.seed(201)
system.time(source("Code/DeerAbundanceUSDA.R"))
```

<!--Make plots-->
<!--***DAN: DANGER this list of dependencies not complete, need to fill in!-->
```{r make_plots, echo=F, results="hide", fig.show="hide", cache=T, cache.extra=list(mtime("Code/deer_plots.R"))}
#generate raw data plots
set.seed(301)
system.time(source("Code/deer_plots.R"))

#***DAN: I got the following warnings when I ran deer_plots.R, we have to get 
#to the bottom of these and make them stop by solving the underlying problem:
#Warning messages:
#  1: In plot.window(...) : "smallplot" is not a graphical parameter
#2: In plot.xy(xy, type, ...) : "smallplot" is not a graphical parameter
#3: In title(...) : "smallplot" is not a graphical parameter
#4: In plot.window(...) : "smallplot" is not a graphical parameter
#5: In plot.xy(xy, type, ...) : "smallplot" is not a graphical parameter
#6: In title(...) : "smallplot" is not a graphical parameter
#7: In plot.window(...) : "smallplot" is not a graphical parameter
#8: In plot.xy(xy, type, ...) : "smallplot" is not a graphical parameter
#9: In title(...) : "smallplot" is not a graphical parameter
#10: In plot.window(...) : "smallplot" is not a graphical parameter
#11: In plot.xy(xy, type, ...) : "smallplot" is not a graphical parameter
#12: In title(...) : "smallplot" is not a graphical parameter
#13: In plot.window(...) : "smallplot" is not a graphical parameter
#14: In plot.xy(xy, type, ...) : "smallplot" is not a graphical parameter
#15: In title(...) : "smallplot" is not a graphical parameter
#16: In plot.window(...) : "smallplot" is not a graphical parameter
#17: In plot.xy(xy, type, ...) : "smallplot" is not a graphical parameter
#18: In title(...) : "smallplot" is not a graphical parameter

#***DAN: Fig 1 does not appear, the code for it needs to be put in
```

# Materials and methods

## Data and pre-processing

<!--***DAN: I inserted a comment in each place that needs to be made 
or may need to be made an automatic link to a reference, table, figure, 
or R-produced value. As these are done, the comment can be deleted,
as can the non-automatic text that the automatic link should replace.
But leave the non-automatic text until you do a knit to make sure you
get the same thing.-->

We obtained deer abundance estimates from the Wisconsin Department of 
Natural Resources (WIDNR) spanning a 36<!--***DAN-->-yr period 
(1981–2016<!--***DAN-->; Fig. S1<!--***DAN-->). 
From 1981–2013<!--***DAN-->, deer abundance was estimated for deer management units 
(DMUs, N = 101–139<!--***DAN--> depending on the year). For most years and DMUs, 
abundance was estimated using a sex-age-kill (SAK) model, with aerial surveys or 
accounting models used for a smaller number of years and DMUs (1)<!--***DAN-->. 
In $2014$, DMU boundaries were changed to match the administrative level of 
county (N = 72<!--***DAN-->). Because of this, and because other 
spatial variables (e.g., DVCs, climate) were gathered at the county level, 
we transformed the 1981–2013<!--***DAN--> abundance estimates for DMUs 
into county-level estimates. To do so, we first calculated the areas 
of intersection between DMU and county boundaries for each year (as 
DMU boundaries changed yearly), using the ‘intersect’ function in 
Arc GIS 10.3 (ESRI, Redlands, CA). We then calculated deer abundance 
for each intersection area by multiplying the amount of deer habitat in 
each intersection area by the WIDNR density estimate of deer for the DMU 
that the intersection fell within. Deer habitat was considered to comprise 
areas of permanent cover (e.g., forests) that were $>4$ ha in size. DMU deer 
densities were calculated by dividing the estimated abundance for the DMU 
by the amount of deer habitat within the DMU (1)<!--***DAN-->. This procedure 
assumes approximately spatially uniform densities of deer within deer 
habitat across each DMU. The abundance for a county was the sum of 
the estimated abundances of the county’s intersections with the DMUs. 
We dropped one county (Menominee) due to incomplete data for most 
variables, leaving 71<!--***DAN--> counties for statistical analysis.

We obtained estimated hunter abundance data for each DMU from the WIDNR 
that spanned the years 1992<!--***DAN-->-2016<!--***DAN--> 
(Fig. S1<!--***DAN-->), which constituted an index of the estimated 
number of hunters during the first two days of the 9-day gun 
hunting season (1<!--***DAN-->), assuming this was a good proxy 
for hunter numbers across the season. 
<!--***DAN: I think we should add 1-2 sentences arguing why this is a  reasonable assumption, perhaps by giving data, if available, on how many gun v other types of hunters there are, and how many people go the opening 2 days versus later in the week. Let's discuss, Tom.-->
We estimated hunter abundances 
for counties using the same method as for deer abundance. We 
dropped 18<!--***DAN--> counties due to incomplete survey 
efforts of hunters or missing data, leaving 53<!--***DAN--> 
counties for subsequent analysis. 

We obtained estimates of DVCs for each county, spanning the 
years 1987<!--***DAN-->–2016<!--***DAN-->, from the WI Department 
of Transportation (WIDOT; Fig. S1<!--***DAN-->). These data reflect 
the number of police-investigated DVCs. These data represent an 
index of DVCs, as opposed to an absolute number of incidents, 
as not all DVCs are reported. In Michigan, nearly half of 
DVCs went unreported to the police or insurance companies 
over a five year period (2<!--***DAN-->). Additionally, 
DVC information was collected with reporting effort that 
may have varied for different counties 
(Daniel Storm, WIDNR, Personal Communication), and the 
threshold of damage for the police to respond to and report on 
an incident increased statewide in 1996 from US\$500 to US\$1000. 
However, features of the data such as these could only have obscured 
and could not have artefactually generated the strongly 
significant spatial coherence relationships we report between 
DVCs and deer abundance. Shortcomings of the DVC data added 
noise to the DVC data and likely obscured the relationship 
with deer abundance, so that if better DVC data were available 
this relationship may have been even stronger than that reported. 
We also obtained traffic volume estimates (millions of miles 
driven) for each county spanning the years 1988–2016 from the 
WIDOT, to control for traffic in our analysis of DVCs.

We obtained daily weather data (minimum and maximum air temperature 
[here called Tmin and Tmax], snow depth, and precipitation) from 
all available weather stations in each county from the 
National Climatic Data Center ([`https://www.ncdc.noaa.gov/`](https://www.ncdc.noaa.gov/)). We 
averaged daily values across the winter ($\text{Dec}_{t-1}$ 
to $\text{Mar}_t$) and across stations within a county to produce 
annual time series of winter weather by county which were used in 
all subsequent analyses (Fig. S1<!--***DAN-->). The number of 
weather stations per county and data collected at each station 
varied, resulting in different spatial extents for each variable 
in order to retain the full temporal extent: Tmin ($N=$ 62<!--***DAN--> 
counties), Tmax ($N=$ 62<!--***DAN-->), precipitation 
($N=$ 64<!--***DAN-->), snow depth ($N=$ 60<!--***DAN-->). 
We also calculated a Winter Severity Index (WSI, $N=$ 44<!--***DAN-->), 
a metric known to impact deer populations (1). WSI is the cumulative 
sum of the total number of days for which snow depth was greater 
than 18 inches and Tmin was simultaneously less than $0^\circ$F. 

We obtained monthly estimates of three global climate indices: 
the North Atlantic Oscillation (NAO), the multivariate 
El Ni\~no/Southern Oscillation Index (MEI), and the Pacific 
Decadal Oscillation (PDO), all of which have been shown to 
influence large ungulate population dynamics in the northcentral 
USA (3, 4<!--***DAN-->). We averaged monthly values across the 
approximate growing season (Apr–Nov), and the winter 
($\text{Dec}_{t-1}$ to $\text{Mar}_t$) for each index 
(Fig. S1<!--***DAN-->). We used a total of 11 weather and 
climate variables in subsequent analyses: winter Tmin, Tmax, 
snow depth, and precipitation; winter NAO, MEI, and PDO; 
growing-season NAO, MEI and PDO; and the WSI. All were annual 
time series at the county level.

## Analysis

We performed all analyses described below and in the main text 
at three spatial scales: counties (N = 44–71<!--***DAN--> depending on 
the variables involved; mean county 
area $\pm$ SD = 2017<!--***DAN--> $\pm$ 799<!--***DAN--> $\text{km}^2$), 
United States Department of Agriculture (USDA) districts ($N = 9$; mean 
district area $\pm$ SD = 16138<!--***DAN--> $\pm$ 799<!--***DAN--> $\text{km}^2$), 
and statewide (Fig. S2<!--***DAN-->). Analyses at three spatial scales 
were performed because the SAK model used by the WIDNR to estimate deer 
abundance for each management unit has been criticized as less accurate 
at smaller spatial scales (5)<!--***DAN-->. However, the bias in abundance 
estimates previously reported may have been overestimated, as more recent analyses 
suggest the uncertainty in SAK estimates is substantially less 
($\pm 20\%$ error rates; WIDNR, personal communication). 
<!--***DAN: In was Robert that told is this, and he is a coathor, so perhaps we should take out the personal communication parens? It would be best if we could get a citation for this, even a white paper, if there is one and we can fit it in.-->
In addition, the criticism of the SAK (5)<!--***DAN--> was based 
on a model that did not match perfectly with practices of the WIDNR. 
<!--***DAN: I added the above sentence, Tom, based on a comment of Roberts that had been inserted as a marginal note in the Word doc that this Rmd replaces. Please check. If we keep the "personal communication" paren above then we should also insert one here.-->
Regardless, 
to ensure our results were not biased by spatial extent, we analyzed 
synchrony at multiple spatial extents. USDA district data were 
produced as the sum or 
average of county-level variables, as the two scales had nested boundaries. 
There were roughly $6$ to $8$ counties per district (Fig. S2<!--***DAN-->). 
Statewide analysis used state total or average time series for each 
variable. In general, the patterns of significance and phase relationships 
we observed remained unchanged in their main substance using data 
aggregated to larger spatial scales (Table S4<!--***DAN-->, S5<!--***DAN-->; 
Fig. S6<!--***DAN-->), suggesting the results for counties are not 
meaningfully biased by possible inaccuracies in SAK abundance estimation 
procedures at the county level.



# Supplementary text

## USDA district results

## Statewide results

\newpage

# Supplementary figures
\begin{figure}[h!]
\includegraphics[width=\textwidth]{Results/FigS1.png}
\caption{Raw time series of (A) deer abundance, (B) deer-vehicle collisions, 
(C) average winter ($\text{Dec}_{t-1}$ to $\text{Mar}_t$) snow depth, and 
(D) winter climate indices (Multivariate El Ni\~no Southern Oscillation Index, 
MEI; Pacific Decadal Oscillation, PDO), and hunter abundances (E). 
Each line in A-C, E represents a county. Other variables are not shown 
(e.g., Tmin) due to their lack of significance in this 
study.\label{FigS1_Timeseries}}
\end{figure}

\newpage

<!--***DAN: currently the figure is not include until Tom adds code for S2 or adds the figure itself to the repo if it was not created using R-->
\begin{figure}[!h]
%\includegraphics[width=\textwidth]{}
\caption{Map of Wisconsin, USA. Black lines and yellow dots delineate county boundaries and centroids. Color shading and black dots delineate USDA district boundaries and centroids.\label{FigS2_CountiesDistricts}}
\end{figure}

\newpage

\begin{figure}[!h]
\includegraphics[width=\textwidth]{Results/FigS3.png}
\caption{Phase differences between variables as a function of timescale for all significant relationships of Table S1<!--***DAN: -->, county-level data: A) winter MEI and deer abundance, B) summer MEI and deer abundance, C) winter PDO and deer abundance, D) snow depth and deer abundance, E) deer abundance and deer-vehicle collisions, F) deer abundance and deer-vehicle collisions adjusted for traffic volume, and G) deer abundance and hunter abundance. $\bar{\theta}$ indicates the mean phase (in units of $\pi$ radians) for the shaded timescale bands. Horizontal axes differ in extent among panels due to different time ranges of available data for different variables. Vertical axes differ between left and right panels because time series were shorter for DVCs.\label{FigS3_PhaseDiffs}}
\end{figure}

\newpage

\begin{figure}[!h]
\includegraphics[width=\textwidth]{Results/FigS4.png}
\caption{Rank, by timescale, of spatial coherence values in surrogate coherence distributions (\emph{Methods}) for significant relationships (Table S1<!--***DAN: -->) using county-level data. The red horizontal dashed line corresponds to a nominal confidence level of $95\%$. In A), results are comparisons of the listed variables with deer abundance. In B), `Abun-DVCs' is a comparison of deer abundance and DVCs, `Abun-Adj. DVCs' is a comparison of deer abundance and DVCs adjusted for traffic, and `Hunters-Abun' is a comparison of hunter numbers and deer abundance. The red horizontal line would be analogous to a $p$-value of $0.05$ for an individual timescale (\emph{Methods}).\label{FigS4_RanksInSurrogs}}
\end{figure}

\newpage

\begin{figure}[!h]
\includegraphics[width=.6\textwidth]{Results/FigS5.png}
\caption{Wavelet mean field (A) and wavelet phasor mean field (B) magnitudes of county-level hunter abundance time series, and predicted synchrony of deer abundance based on a wavelet model with sole predictor hunters (C). Higher values in A, B indicate greater synchrony (A) or phase synchrony (B) at the indicated times and timescales. Contours on (B) indicate statistical significance of phase synchrony at $P = 0.001$. Colors in (C) show predicted synchrony in deer abundance, while contours show the actual wavelet mean field magnitude of deer abundance (from Fig. 2A<!--***DAN: -->). Agreement is poor, giving a visual indication of the statistical result that hunter abundance does not explain deer abundance or synchrony well on $3–7$-year timescales.\label{FigS5_WMFHunter}}
\end{figure}

\newpage

\begin{figure}[!h]
\includegraphics[height=.8\textheight]{Results/FigS6.png}
\caption{Wavelet mean field (A, B) and phasor mean field (C, D) magnitudes, and predicted synchrony (E, F) for deer abundance (A, C, E) and DVCs (B, D, F) using USDA-district data. Higher values in A-D indicate greater synchrony (A, B) or phase synchrony (C, D) at the indicated times and timescales. Contours on (C, D) indicate statistical significance of phase synchrony at $P = 0.001$. Colors in (E) show synchrony in deer abundance predicted from a wavelet model using snow depth, MEIw, and PDOw as predictors, while contours show the actual wavelet mean field magnitude of deer abundance from (A), for comparison; agreement is good, giving a visual indication of the statistical result that climatic Moran effects explain deer synchrony. Colors in (F) show synchrony in DVCs predicted by a wavelet model with sole predictor deer abundance, while contours are from (B) and again show good agreement. Vertical axes differ between left and right panels because time series were shorter for DVCs. \label{FigS6_WMFAndPredictionsDistricts}}
\end{figure}

\newpage

\begin{figure}[!h]
\includegraphics[height=.9\textheight]{Results/FigS7.png}
\caption{Wavelet mean fields (left column) and wavelet phasor mean fields (right column) of winter Tmin (A, B), Tmax (C, D), snow depth (E, F) and precipitation (G, H) using county-level data. Colors and contours as in Fig 2A,B<!--***DAN: -->. \label{FigS7_WMFsAndWPMFsAncillary}}
\end{figure}

\newpage

\begin{figure}[!h]
\includegraphics[height=.9\textheight]{Results/FigS8.png}
\caption{Wavelet amplitude of the winter and summer PDO (A,B), NAO (C,D), and MEI (E,F), respectively. Amplitude is the magnitude of the wavelet transform, normalized by the square root of the time-averaged, squared magnitudes of the wavelet transform. Higher values indicate greater variation content at that time/or timescale. \label{FigS8_WaveletAmps}}
\end{figure}

\newpage
\clearpage

# Supplementary tables

<!--***DAN: Table S1, currently commented until Dan gets the name of the R dataframe containing table S1 from Tom-->
```{r TableS1,echo=F,results='markup'}
#knitr::kable(,#Insert name of R dataframe containing table S1 here
#             format='pandoc',
#             caption = "Summary of spatial coherence tests and related results using county-level data #(\emph{Methods}). Each row is based on a test of spatial coherence of the variables in the first two columns. Tests were across the $3–7$-year timescale band, except for ancillary results in the footnotes. $P$-values are provided for all tests, and average phases over the timescale band used, fractions of synchrony explained, and average cross terms (\emph{Methods}) are provided where meaningful (for significant tests; $P < 0.05$). A positive mean phase indicates the response is leading, and a negative mean phase indicates the predictor is leading. Cross-terms are zero for the climate indices because they used the same values for all locations.\\label{TabS1_CountyCoherenceSummary}",
#             booktabs=T)
```