---
title: "Moran effects cause major cycles in deer populations and deer-vehicle collisions across Wisconsin: Supplementary Materials"
author: "Anderson, Sheppard, Walter, Rolley, Reuman"
date: "April 20, 2018"
geometry: "left=1cm,right=1cm,top=2.5cm,bottom=2.8cm"

output: 
  pdf_document:
    keep_tex: yes
    fig_caption: yes
tables: True
link-citations: True
urlcolor : blue

csl: science.csl
bibliography: refs.bib
---

<!--***DAN: Tom, in order for me to put in demos of how some things work here are some 
items I need from you first:
1) the name of an R variable in your code for some in-text number, and an indication of which number, e.g., your response might look like "the p-value in the first line under Table S1 is stored in the R variable  SnowDepthAbundanceCohP35".
2) the name of an R data frame that contains Table S1, e.g., you might just call it TableS1df. 
3) Please open Endnote and try to figure out how to expport your library of ciations that you are using
for this paper as a bibtex file (.bib extension). This reference may help: https://www.reed.edu/cis/help/LaTeX/EndNote.html. Then commit and push the file to the repo in the same folder as this Rmd and let me know.-->

<!--***DAN: Currently we are using science citation style in the supp mat, is this 
required, Tom? If not, wwe should use the callout style we prefer. What style do you
prefer, Tom?-->

<!--checkpoint package for CRAN-available packages-->
```{r checkpoint_chunk, echo=F, warning=F, message=F, results="hide"}
if (!dir.exists("./.checkpoint/")){
  dir.create("./.checkpoint/")
}
library(checkpoint)
checkpoint("2018-04-20",checkpointLocation = "./")
#.libPaths()[1]
```

<!--Install the correct reumannplatz version from github, i.e., the one current on 2018 04 20, tagged
in git with AndersonDeer-->
```{r rp_install}
library(withr)
library(devtools)
#with_libpaths(new = "/usr/lib/R/site-library/", install_github('rCharts', 'ramnathv'))
withr::with_libpaths(new=.libPaths()[1],
                     devtools::install_github(repo="reumandc/reumannplatz",ref="AndersonDeer",force=T))
```

<!--Libraries. Should use checkpoint, but having trouble using reumannplatz from github and checkpoint
at the same time. To be solved later if possible.-->
```{r libraries}
library(reshape2)
library(tidyr)
library(plyr)
library(tools)
library(fields)
library(shape)
#x<-.libPaths()[1]
#y<-library(lib.loc=.libPaths())
#save(x,y,file="LibraryPath.RData")
library(Reumannplatz,lib.loc=.libPaths())
```

<!--Flags controlling the run and other setup-->
```{r setup}
#which density to use ("dnr" for original estimates, "tom" for your own calculations)
dens.flag<-"dnr"

#what is the earliest year you want? (hunting only: 1960; abundance: 1981; DVC: 1987; Traffic: 1988, Hunters: 1992)
minyear<-1981

#what is the last year you want? (max is 2016)
maxyear<-2016
  
#normalize the data ("norm" for Box-Cox, "detr" for linearly detrend, or "none")
##Code currently work using the "none" option; the normalization steps happen 
##individually with each analysis step (e.g. see line 15 in "DeerAbundance.R" file), 
##because of the conflicting data lengths and dimensions. 
detr.flag<-"none"

#set spatial scale flag for data generation ("usda", "county" or "both")
scale.flag<-"both"

#***DAN: Not sure what this is, Tom, if this is not necessary should be deleted? 
#calculate how many missing climate values there were (don't need to run anymore)
samplesize.flag<-"n"

#set number of surrogates to use for spatial coherence
nsurrogs<-10000

#a function needed for caching
source("Code/mtime.R") 
```

<!--***DAN: Rum time is sufficiently long that we should probably use caching, but
that will require a good understanding of dependencies of everything. I have begun.
Other reasons to use cacheing include it will enforce a clearer specification of
dependencies between chunks.-->

<!--Clean the population data-->
```{r clean_pop_data, cache=T,cache.extra=list(dens.flag,minyear,maxyear,detr.flag,mtime("Code/wi.deer.cleandata.R"),mtime("Data/State_Kill.csv"),mtime("Data/County_KillTotal.csv"),mtime("Data/County_Kill_BuckOnly.csv"),mtime("Data/County_Kill_DoeOnly.csv"),mtime("Data/County_KillGunBuck.csv"),mtime("Data/County_KillGunDoe.csv"),mtime("Data/County_KillBowBuck.csv"),mtime("Data/County_KillBowDoe.csv"),mtime("Data/WI_Cty_CentroidDD.csv"),mtime("Data/DOT_Accidents1987_2016.csv"),mtime("Data/Adj_DOT_Accidents1987_2016.csv"),mtime("Data/OpeningDay.csv"),mtime("Data/CornHarvest.csv"),mtime("Data/WI_USDA_Sections.csv"),mtime("Data/USDA_centroids.csv"),mtime("Data/Cty_Traffic.csv"),mtime("Data/DMU_DeerAbundance1981_2013.csv"),mtime("Data/DMU_DeerDensity1981_2013.csv"),mtime("Data/DMU_HunterAbundance1975_2013.csv"),mtime("Data/DMU_HunterDensity1975_2013.csv"),mtime("Data/hunterdens1416.csv"),mtime("Data/hunterabun1416.csv"),mtime("Data/Cty_Abun_2014_2016.csv"),mtime("Data/permit_issued1970_2013.csv"))}
#Clean population data: end result are two lists of matrices called "cty.list" and "usda.list"
system.time(source("Code/wi.deer.cleandata.R"))
  #list of files generated by wi.deer.cleandata.R:
    #scale.flag == "usda": usda.list, includes data on abundance, kill data, crops, DVCs and traffic
    #scale.flag == "county": cty.list, includes data on abundance, kill data, crops, DVCs and traffic
```

<!--Clean the climate data-->
```{r clean_clim_data, cache=T, cache.extra=list(minyear,maxyear,detr.flag,scale.flag,mtime("Code/wi_climate.R"),mtime("Data/noaa_latlongUpd.csv"),mtime("Data/WI_USDA_Sections.csv"),mtime("Data/Cty_NOAA_Temp1981_2015_set1.csv"),mtime("Data/Cty_NOAA_Temp1981_2015_set2.csv"),mtime("Data/Cty_NOAA_Temp1981_2015_set3.csv"),mtime("Data/Cty_NOAA_Temp1981_2015_set4.csv"),mtime("Data/Cty_NOAA_Temp1981_2015_set5.csv"),mtime("Data/Cty_NOAA_Temp1981_2015_set6.csv"),mtime("Data/Cty_NOAA_Temp2016_2017.csv"),mtime("Data/Cty_NOAA_Precip1981_1985.csv"),mtime("Data/Cty_NOAA_Precip1986_1991.csv"),mtime("Data/Cty_NOAA_Precip1992_1997.csv"),mtime("Data/Cty_NOAA_Precip1998_2003.csv"),mtime("Data/Cty_NOAA_Precip2004_2009.csv"),mtime("Data/Cty_NOAA_Precip2010_2015.csv"),mtime("Data/Cty_NOAA_Precip2016_2017.csv"),mtime("Data/climate_indices.csv"),mtime("Data/mei.csv"),mtime("Data/pdo.csv"),mtime("Data/npi.csv"),mtime("Data/ENSO.csv"))}
system.time(source("Code/wi_climate.R"))
#list of files generated by wi_climate.R:
  #scale.flag == "usda": wsi.usda.mat,hunt.clim.usda, winter.clim.usda,nao.mat,pdo.mat,mei.mat,winter.nao,winter.pdo,winter.mei
  #scale.flag == "county": wsi.mat,wsi.matN,hunt.clim,winter.clim,nao.mat,pdo.mat,mei.mat,winter.nao,winter.pdo,winter.mei
  #large climate indices are the same values (nao,mei,pdo), just replicated into different dimensions to match scale.flag
```

<!--County analysis-->
<!--***DAN: DANGER this list of dependencies not complete, need to fill in!-->
```{r county_analysis, cache=T, cache.extra=list(mtime("Code/DeerAbundance.R"))}
#Run analysis on deer abundance and DVCs using county data
set.seed(101)
#Rin<-require(Reumannplatz)
#save(Rin,file="Rin.RData")
system.time(source("Code/DeerAbundance.R"))
```

<!--***DAN: You want this whole code suite to be *exactly* reproducible, so any random bits
need to have the seed set first. Check for that.-->

<!--District analysis-->
<!--***DAN: DANGER this list of dependencies not complete, need to fill in!-->
```{r district_analysis, cache=T, cache.extra=list(mtime("Code/DeerAbundanceUSDA.R"))}
#Run analysis on deer abundance and DVCs using USDA district data
set.seed(201)
system.time(source("Code/DeerAbundanceUSDA.R"))
```

<!--Make plots-->
<!--***DAN: DANGER this list of dependencies not complete, need to fill in!-->
```{r make_plots, cache=T, cache.extra=list(mtime("Code/deer_plots.R"))}
#generate raw data plots
set.seed(301)
system.time(source("Code/deer_plots.R"))

#***DAN: I got the following warnings when I ran deer_plots.R, we have to get 
#to the bottom of these and make them stop by solving the underlying problem:
#Warning messages:
#  1: In plot.window(...) : "smallplot" is not a graphical parameter
#2: In plot.xy(xy, type, ...) : "smallplot" is not a graphical parameter
#3: In title(...) : "smallplot" is not a graphical parameter
#4: In plot.window(...) : "smallplot" is not a graphical parameter
#5: In plot.xy(xy, type, ...) : "smallplot" is not a graphical parameter
#6: In title(...) : "smallplot" is not a graphical parameter
#7: In plot.window(...) : "smallplot" is not a graphical parameter
#8: In plot.xy(xy, type, ...) : "smallplot" is not a graphical parameter
#9: In title(...) : "smallplot" is not a graphical parameter
#10: In plot.window(...) : "smallplot" is not a graphical parameter
#11: In plot.xy(xy, type, ...) : "smallplot" is not a graphical parameter
#12: In title(...) : "smallplot" is not a graphical parameter
#13: In plot.window(...) : "smallplot" is not a graphical parameter
#14: In plot.xy(xy, type, ...) : "smallplot" is not a graphical parameter
#15: In title(...) : "smallplot" is not a graphical parameter
#16: In plot.window(...) : "smallplot" is not a graphical parameter
#17: In plot.xy(xy, type, ...) : "smallplot" is not a graphical parameter
#18: In title(...) : "smallplot" is not a graphical parameter

#***DAN: Fig 1 does not appear, the code for it needs to be put in
```

# Materials and methods

## Data and pre-processing

<!--***DAN: I inserted a comment in each place that needs to be made 
or may need to be made an automatic link to a reference, table, figure, 
or R-produced value. As these are done, the comment can be deleted,
as can the non-automatic text that the automatic link should replace.
But leave the non-automatic text until you do a knit to make sure you
get the same thing.-->

<!--***DAN: demo of ordinary text-->
We obtained deer abundance estimates from the Wisconsin Department of 
Natural Resources (WIDNR) spanning a 36<!--***DAN-->-yr period 
(1981–2016<!--***DAN-->; Fig. S1<!--***DAN-->). 
From 1981–2013<!--***DAN-->, deer abundance was estimated for deer management units 
(DMUs, N = 101–139<!--***DAN--> depending on the year). For most years and DMUs, 
abundance was estimated using a sex-age-kill (SAK) model, with aerial surveys or 
accounting models used for a smaller number of years and DMUs (1)<!--***DAN-->. 
In $2014$, DMU boundaries were changed to match the administrative level of 
county (N = 72<!--***DAN-->). Because of this, and because other 
spatial variables (e.g., DVCs, climate) were gathered at the county level, 
we transformed the 1981–2013<!--***DAN--> abundance estimates for DMUs 
into county-level estimates. To do so, we first calculated the areas 
of intersection between DMU and county boundaries for each year (as 
DMU boundaries changed yearly), using the ‘intersect’ function in 
Arc GIS 10.3 (ESRI, Redlands, CA). We then calculated deer abundance 
for each intersection area by multiplying the amount of deer habitat in 
each intersection area by the WIDNR density estimate of deer for the DMU 
that the intersection fell within. Deer habitat was considered to comprise 
areas of permanent cover (e.g., forests) that were $>4$ ha in size. DMU deer 
densities were calculated by dividing the estimated abundance for the DMU 
by the amount of deer habitat within the DMU (1)<!--***DAN-->. This procedure 
assumes approximately spatially uniform densities of deer within deer 
habitat across each DMU. The abundance for a county was the sum of 
the estimated abundances of the county’s intersections with the DMUs. 
We dropped one county (Menominee) due to incomplete data for most 
variables, leaving 71<!--***DAN--> counties for statistical analysis.

<!--***DAN: Much text skipped for now-->

## Analysis

We performed all analyses described below and in the main text 
at three spatial scales: counties (N = 44–71<!--***DAN--> depending on 
the variables involved; mean county 
area $\pm$ SD = 2017<!--***DAN--> $\pm$ 799<!--***DAN--> $\text{km}^2$), 
United States Department of Agriculture (USDA) districts ($N = 9$; mean 
district area $\pm$ SD = 16138<!--***DAN--> $\pm$ 799<!--***DAN--> $\text{km}^2$), 
and statewide (Fig. S2<!--***DAN-->). Analyses at three spatial scales 
were performed because the SAK model used by the WIDNR to estimate deer 
abundance for each management unit has been criticized as less accurate 
at smaller spatial scales (5)<!--***DAN-->. However, the bias in abundance 
estimates previously reported may overestimated, as more recent analyses 
suggest the uncertainty in SAK estimates is substantially less 
($\pm 20\%$ error rates; WIDNR, personal communication). Regardless, 
to ensure our results were not biased by spatial extent, we analyzed 
synchrony at multiple spatial extents. USDA districts were the sum or 
average of county-level variables, as the two scales had nested boundaries, 
roughly equating to $6$ to $8$ counties per district (Fig. S2<!--***DAN-->). 
Statewide analysis used state total or average time series for each 
variable. In general, the patterns of significance and phase relationships 
we observed remained unchanged in their main substance using data 
aggregated to larger spatial scales (Table S4<!--***DAN-->, S5<!--***DAN-->; 
Fig. S6<!--***DAN-->), suggesting the results for counties are not 
meaningfully biased by any possible inaccuracies in SAK abundance estimation 
procedures at the county level.

<!--***DAN: Much text skipped for now-->

# Supplementary text

## USDA district results

## Statewide results

\newpage

# Supplementary figures
\begin{figure}[h!]
\includegraphics[width=\textwidth]{Results/FigS1.png}
\caption{Raw time series of (A) deer abundance, (B) deer-vehicle collisions, 
(C) average winter ($\text{Dec}_{t-1}$ to $\text{Mar}_t$) snow depth, and 
(D) winter climate indices (Multivariate El Ni\~no Southern Oscillation Index, 
MEI; Pacific Decadal Oscillation, PDO), and hunter abundances (E). 
Each line in A-C, E represents a county. Other variables are not shown 
(e.g., Tmin) due to their lack of significance in this 
study.\label{FigS1_Timeseries}}
\end{figure}

\newpage

<!--***DAN: currently the figure is not include until Tom adds code for S2 or adds the figure itself to the repo if it was not created using R-->
\begin{figure}[!h]
%\includegraphics[width=\textwidth]{}
\caption{Map of Wisconsin, USA. Black lines and yellow dots delineate county boundaries and centroids. Color shading and black dots delineate USDA district boundaries and centroids.\label{FigS2_CountiesDistricts}}
\end{figure}

\newpage

\begin{figure}[!h]
\includegraphics[width=\textwidth]{Results/FigS3.png}
\caption{Phase differences between variables as a function of timescale for all significant relationships of Table S1<!--***DAN: -->, county-level data: A) winter MEI and deer abundance, B) summer MEI and deer abundance, C) winter PDO and deer abundance, D) snow depth and deer abundance, E) deer abundance and deer-vehicle collisions, F) deer abundance and deer-vehicle collisions adjusted for traffic volume, and G) deer abundance and hunter abundance. $\bar{\theta}$ indicates the mean phase (in units of $\pi$ radians) for the shaded timescale bands. Horizontal axes differ in extent among panels due to different time ranges of available data for different variables. Vertical axes differ between left and right panels because time series were shorter for DVCs.\label{FigS3_PhaseDiffs}}
\end{figure}

\newpage

\begin{figure}[!h]
\includegraphics[width=\textwidth]{Results/FigS4.png}
\caption{Rank, by timescale, of spatial coherence values in surrogate coherence distributions (\emph{Methods}) for significant relationships (Table S1<!--***DAN: -->) using county-level data. The red horizontal dashed line corresponds to a nominal confidence level of $95\%$. In A), results are comparisons of the listed variables with deer abundance. In B), `Abun-DVCs' is a comparison of deer abundance and DVCs, `Abun-Adj. DVCs' is a comparison of deer abundance and DVCs adjusted for traffic, and `Hunters-Abun' is a comparison of hunter numbers and deer abundance. The red horizontal line would be analogous to a $p$-value of $0.05$ for an individual timescale (\emph{Methods}).\label{FigS4_RanksInSurrogs}}
\end{figure}

\newpage

\begin{figure}[!h]
\includegraphics[width=.6\textwidth]{Results/FigS5.png}
\caption{Wavelet mean field (A) and wavelet phasor mean field (B) magnitudes of county-level hunter abundance time series, and predicted synchrony of deer abundance based on a wavelet model with sole predictor hunters (C). Higher values in A, B indicate greater synchrony (A) or phase synchrony (B) at the indicated times and timescales. Contours on (B) indicate statistical significance of phase synchrony at $P = 0.001$. Colors in (C) show predicted synchrony in deer abundance, while contours show the actual wavelet mean field magnitude of deer abundance (from Fig. 2A<!--***DAN: -->). Agreement is poor, giving a visual indication of the statistical result that hunter abundance does not explain deer abundance or synchrony well on $3–7$-year timescales.\label{FigS5_WMFHunter}}
\end{figure}

\newpage

\begin{figure}[!h]
\includegraphics[height=.8\textheight]{Results/FigS6.png}
\caption{Wavelet mean field (A, B) and phasor mean field (C, D) magnitudes, and predicted synchrony (E, F) for deer abundance (A, C, E) and DVCs (B, D, F) using USDA-district data. Higher values in A-D indicate greater synchrony (A, B) or phase synchrony (C, D) at the indicated times and timescales. Contours on (C, D) indicate statistical significance of phase synchrony at $P = 0.001$. Colors in (E) show synchrony in deer abundance predicted from a wavelet model using snow depth, MEIw, and PDOw as predictors, while contours show the actual wavelet mean field magnitude of deer abundance from (A), for comparison; agreement is good, giving a visual indication of the statistical result that climatic Moran effects explain deer synchrony. Colors in (F) show synchrony in DVCs predicted by a wavelet model with sole predictor deer abundance, while contours are from (B) and again show good agreement. Vertical axes differ between left and right panels because time series were shorter for DVCs. \label{FigS6_WMFAndPredictionsDistricts}}
\end{figure}

\newpage

\begin{figure}[!h]
\includegraphics[height=.9\textheight]{Results/FigS7.png}
\caption{Wavelet mean fields (left column) and wavelet phasor mean fields (right column) of winter Tmin (A, B), Tmax (C, D), snow depth (E, F) and precipitation (G, H) using county-level data. Colors and contours as in Fig 2A,B<!--***DAN: -->. \label{FigS7_WMFsAndWPMFsAncillary}}
\end{figure}

\newpage

\begin{figure}[!h]
\includegraphics[height=.9\textheight]{Results/FigS8.png}
\caption{Wavelet amplitude of the winter and summer PDO (A,B), NAO (C,D), and MEI (E,F), respectively. Amplitude is the magnitude of the wavelet transform, normalized by the square root of the time-averaged, squared magnitudes of the wavelet transform. Higher values indicate greater variation content at that time/or timescale. \label{FigS8_WaveletAmps}}
\end{figure}

\newpage
\clearpage

# Supplementary tables

<!--***DAN: Table S1, currently commented until Dan gets the name of the R dataframe containing table S1 from Tom-->
```{r TableS1,echo=F,results='markup'}
#knitr::kable(,#Insert name of R dataframe containing table S1 here
#             format='pandoc',
#             caption = "Summary of spatial coherence tests and related results using county-level data #(\emph{Methods}). Each row is based on a test of spatial coherence of the variables in the first two columns. Tests were across the $3–7$-year timescale band, except for ancillary results in the footnotes. $P$-values are provided for all tests, and average phases over the timescale band used, fractions of synchrony explained, and average cross terms (\emph{Methods}) are provided where meaningful (for significant tests; $P < 0.05$). A positive mean phase indicates the response is leading, and a negative mean phase indicates the predictor is leading. Cross-terms are zero for the climate indices because they used the same values for all locations.\\label{TabS1_CountyCoherenceSummary}",
#             booktabs=T)
```